---
title: "Build Model Data"
params:
  bootstrap_rep: 3
format: html
---

```{r}
#| label: timestamp-start
#| echo: false

timestamp()
```

```{r}
#| label: setup
#| include: false
library(here)
source(here("utils","check_packages.R"))
source(here("utils","functions.R"))
```

```{r}
#| label: prepare-data
#| include: false

load(here("data", "data_constructed", "data_couples.RData"))

census <- census |>
  # relevel year
  mutate(year = relevel(year, "2000"))
```

This script reads in the full couple-level level dataset across all census/ACS years and creates a list of actual and bootstrap sampled model level data. This model level data is a contingency table and is produced by the function `create_model-data` although here that is wrapped in a `generate_bootstrap_data` function that repeats this procedure multiple times. The first dataset in the produced list is for the actual data and the remaining datasets in the list are for the bootstrapped samples. This list structure can then be read into `bootstrap_model` to estimate actual models.

```{r}
#| label: tbl-show-boostrap-values
#| tbl-cap: Table summarizing important values for the bootstrap procedure
#| echo: false

tribble( 
  ~name, ~value,
  "Number of bootstrap samples", params$bootstrap_rep,
  "Number of couple observations", nrow(census)
) |>
  gt() |>
  fmt_number(value, decimals = 0)
```

# State marriage data

The workhorse model data structure will group observations by the race of both spouses, the state, and year. For models which estimate different rates by region, I also need a region variable. However, since each state can only be in one region, adding region to the grouping would lead to a bloated dataset with a bunch of zero values. So, instead, I just attach region afterwards with a join. 

Note that I can still collapse racial groups on the model data structure later by mutating `race_husband` and `race_wife` and then re-running a `group_by` and `summarize` on the frequencies.

I also use `control_var = "gen"` here to separate results by generational status so that I can build those models for comparison.

```{r}
#| label: generate-state-region-model-data

tic()
data_model_state <- census |>
  # only native born
  filter(gen_husband == "2nd+" & gen_wife == "2nd+") |>
  # only marriages
  filter(union_type == "marriage") |>
  generate_bootstrap_data(params$bootstrap_rep, 
                          composition_var = "state")
toc()

# add in the region for each state
region_crosswalk <- census |>
  select(state, region) |>
  distinct()

data_model_state <- data_model_state |>
  map(~ {
    .x |>
      left_join(region_crosswalk, by = "state") |>
      relocate(region, .before = freq)
  })
  
save(data_model_state, 
     file = here("data", "data_constructed", "data_model_state.RData"))
rm(data_model_state)
gc()
```

# Generational status

```{r}
#| label: generate-state-gen-data

tic()
data_model_state_no_first_gen <- census |>
  # only native born
  filter(gen_husband != "1st" & gen_wife != "1st") |>
  # drop 1960
  filter(year != "1960") |>
  # only marriages
  filter(union_type == "marriage") |>
  generate_bootstrap_data(params$bootstrap_rep, 
                          composition_var = "state")
toc()

tic()
data_model_state_all_gen <- census |>
  # drop 1960
  filter(year != "1960") |>
  # only marriages
  filter(union_type == "marriage") |>
  generate_bootstrap_data(params$bootstrap_rep, 
                          composition_var = "state")
toc()

save(data_model_state_no_first_gen, data_model_state_all_gen,
     file = here("data", "data_constructed", "data_model_state_gen.RData"))
rm(data_model_state_all_gen, data_model_state_no_first_gen)
gc()
```

# Cohabitation

```{r}
#| label: generate-cohab-data

tic()
data_model_state_cohab <- census |>
  # only native born
  filter(gen_husband == "2nd+" & gen_wife == "2nd+") |>
  # only marriages
  filter(union_type == "cohabitation") |>
  generate_bootstrap_data(params$bootstrap_rep, composition_var = "state")
toc()

data_model_state_cohab <- data_model_state_cohab |>
  map(~ {
    .x |>
      left_join(region_crosswalk, by = "state") |>
      relocate(region, .before = freq)
  })
  
save(data_model_state_cohab, 
     file = here("data", "data_constructed", "data_model_state_cohab.RData"))
rm(data_model_state_cohab)
gc()
```


# Best race data

For the models using a person's best race, I need to generate a separate model data structure as this variable changes at the individual level.

```{r}
#| label: generate-best-race-model-data

transform_race <- function(race, best_race) {
  
  race <- as.character(race)
  race <- case_when(
    race %in% c("EastAsian", "SEAsian", "SouthAsian", "OtherAsian", 
                "PI") ~ "API",
    TRUE ~ race
  )
  
  race <- if_else(!is.na(best_race), best_race, race)
  race <- factor(race, levels = c("White", "Black", "AIAN", "API"))
  return(race)
}

tic()
data_model_best_race_state <- census |>
  # only native born
  filter(gen_husband == "2nd+" & gen_wife == "2nd+") |>
  # only marriages
  filter(union_type == "marriage") |>
  mutate(race_husband = transform_race(race_husband, race_best_husband),
         race_wife = transform_race(race_wife, race_best_wife)) |>
  filter(!is.na(race_husband) & !is.na(race_wife)) |>
  generate_bootstrap_data(params$bootstrap_rep, composition_var = "state")
toc()

save(data_model_best_race_state, 
     file = here("data", "data_constructed", "data_model_best_race_state.RData"))
```

```{r}
#| label: timestamp-end
#| echo: false

timestamp()
```




