---
title: "Check model approach"
format: html
---

```{r}
#| label: setup
#| include: false
library(here)
source(here("utils","check_packages.R"))
source(here("utils","functions.R"))
load(here("data", "data_constructed", "mar_table.RData"))
```


Lets create a function that will calculate the log-odds ratio and its standard error from `tab-mar` for two groups.

```{r}
calculate_or <- function(group1, group2) {
  lor <- log(tab_mar[group1, group2,])+log(tab_mar[group2, group1,])-
    log(tab_mar[group1, group1,])-log(tab_mar[group2, group2,])
  se <- sqrt(1/tab_mar[group1, group2,]+1/tab_mar[group2, group1,]+
               1/tab_mar[group1, group1,]+1/tab_mar[group2, group2,])
  tibble(year = as.numeric(names(lor)), lor, se, 
         type = paste(group1, group2, sep="-"))
}
```

We can now use `pmap` to calculate log odds ratios across all groups of interest.

```{r}
#| label: get-lor-df

all_combos <- tribble(
  ~race1, ~race2,
  "White", "Black",
  "White", "AIAN",
  "White", "API",
  "White", "Hispanic",
  "Black", "AIAN",
  "Black", "API",
  "Black", "Hispanic",
  "AIAN", "API",
  "AIAN", "Hispanic",
  "API", "Hispanic",
  "White/Black", "White",
  "White/Black", "Black",
  "White/API", "White",
  "White/API", "API",
  "White/AIAN", "White",
  "White/AIAN", "AIAN"
)

lor_data <- pmap(all_combos, function(race1, race2) {
  calculate_or(race1, race2)
  }) |>
  bind_rows() |>
  filter(lor > -Inf)
```

Now lets try a model approach.

First convert to a frequency based dataset

```{r}
mar_data_full <- as.data.frame.table(tab_mar) |>
  tibble() |>
  rename(race_husband = Var1, race_wife = Var2, year = Var3, freq = Freq) |>
  filter(year != 1960)
```


## Two-group case

Start with two groups. By default you will get an endogamy parameter so switch this to an intermarriage parameter.

```{r}
mar_data_white_black <- mar_data_full |>
  filter(race_husband %in% c("White", "Black") & 
           race_wife %in% c("White", "Black")) |>
  mutate(inter_w_b = (race_husband == "White" & race_wife == "Black"))

model_wb <- glm(freq~(race_husband + race_wife + inter_w_b) * year, 
                data = mar_data_white_black, family = poisson)

wb_lor <- avg_slopes(model_wb, variables = "inter_w_b", by = "year", 
                     type = "link")
```

How does this compare to cross-product?

```{r}

temp <- wb_lor |>
  as_tibble() |>
  mutate(year = as.numeric(paste(year))) |>
  rename(lor_model = estimate, se_model = std.error) |>
  select(lor_model, se_model, year)

temp <- lor_data |> 
  filter(type == "White-Black") |>
  rename(lor_cross = lor, se_cross = se) |>
  select(-type) |>
  inner_join(temp)
  
ggplot(temp, aes(x = lor_cross, y = lor_model))+
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "red")+
  geom_point()+
  labs(x = "White/Black log odds ratio by cross-product",
       y = "White/Black lod odds ratio by model")+
  #geom_text_repel(aes(label = year))+
  theme_bw()

ggplot(temp, aes(x = se_cross, y = se_model))+
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "red")+
  geom_point()+
  labs(x = "White/Black log odds ratio standard error by cross-product",
       y = "White/Black lod odds ratio standard error by model")+
  theme_bw()
```

Perfect fit.

## Three-group case

Now we have a spare parameter to fit.

```{r}
mar_data_three_group <- mar_data_full |>
  filter(race_husband %in% c("White", "Black", "API") & 
           race_wife %in% c("White", "Black", "API")) |>
  mutate(inter_w_b = (race_husband == "White" & race_wife == "Black"),
         inter_w_a = (race_husband == "White" & race_wife == "API"),
         inter_b_a = (race_husband == "Black" & race_wife == "API"),
         gender_b_a = case_when(
           race_husband == "API" & race_wife == "Black" ~ 1,
           race_husband == "Black" & race_wife == "API" ~ -1,
           TRUE ~ 0
         ))

model_three_group <- glm(freq~(race_husband + race_wife + 
                                 inter_w_b +
                                 inter_w_a + 
                                 inter_b_a + 
                                 gender_b_a) * year, 
                data = mar_data_three_group, family = poisson)

inter_lor <- avg_slopes(model_three_group, 
                        variables = c("inter_w_b", "inter_w_a", "inter_b_a"), 
                        by = "year", 
                        type = "link")
```



```{r}

temp <- inter_lor |>
  as_tibble() |>
  mutate(
    year = as.numeric(paste(year)),
    type = case_when(
      term == "inter_w_b" ~ "White-Black",
      term == "inter_w_a" ~ "White-API",
      term == "inter_b_a" ~ "Black-API",
    )
  ) |>
  rename(lor_model = estimate, se_model = std.error) |>
  select(lor_model, se_model, year, type)

temp <- lor_data |> 
  filter(type %in% c("White-Black", "White-API", "Black-API")) |>
  rename(lor_cross = lor, se_cross = se) |>
  inner_join(temp)
  
ggplot(temp, aes(x = lor_cross, y = lor_model))+
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "red")+
  geom_point()+
  labs(x = "White/Black log odds ratio by cross-product",
       y = "White/Black lod odds ratio by model")+
  facet_wrap(~type)+
  theme_bw()

ggplot(temp, aes(x = se_cross, y = se_model))+
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "red")+
  geom_point()+
  labs(x = "White/Black log odds ratio standard error by cross-product",
       y = "White/Black lod odds ratio standard error by model")+
  facet_wrap(~type)+
  theme_bw()
```


# All single race groups

Sparseness problems are created for years where we get zero cells. This causes all of the SEs in the AMEs to be wrong. Filtering out frequencies of zero seems to fix the problem. The `avg_slopes` gives a warning about rank deficient matrices but it fits everything fine.

```{r}
mar_data_single <- mar_data_full |>
  #filter(year != "1970") |>
  filter(!str_detect(race_husband, "/|Other"), 
         !str_detect(race_wife, "/|Other")) |>
  filter(freq > 0) |>
  mutate(
    inter_w_b = (race_husband == "White" & race_wife == "Black"),
    inter_w_a = (race_husband == "White" & race_wife == "API"),
    inter_w_i = (race_husband == "White" & race_wife == "AIAN"),
    inter_w_h = (race_husband == "White" & race_wife == "Hispanic"),
    inter_b_a = (race_husband == "Black" & race_wife == "API"),
    inter_b_i = (race_husband == "Black" & race_wife == "AIAN"),
    inter_b_h = (race_husband == "Black" & race_wife == "Hispanic"),
    inter_a_i = (race_husband == "API" & race_wife == "AIAN"),
    inter_a_h = (race_husband == "API" & race_wife == "Hispanic"),
    inter_i_h = (race_husband == "AIAN" & race_wife == "Hispanic"),
    gender_b_a = case_when(
      race_husband == "API" & race_wife == "Black" ~ 1,
      race_husband == "Black" & race_wife == "API" ~ -1,
      TRUE ~ 0
    ),
    gender_b_i = case_when(
      race_husband == "AIAN" & race_wife == "Black" ~ 1,
      race_husband == "Black" & race_wife == "AIAN" ~ -1,
      TRUE ~ 0
    ),
    gender_b_h = case_when(
      race_husband == "Hispanic" & race_wife == "Black" ~ 1,
      race_husband == "Black" & race_wife == "Hispanic" ~ -1,
      TRUE ~ 0
    ),
    gender_a_i = case_when(
      race_husband == "AIAN" & race_wife == "API" ~ 1,
      race_husband == "API" & race_wife == "AIAN" ~ -1,
      TRUE ~ 0
    ),
    gender_a_h = case_when(
      race_husband == "Hispanic" & race_wife == "API" ~ 1,
      race_husband == "API" & race_wife == "Hispanic" ~ -1,
      TRUE ~ 0
    ),
    gender_i_h = case_when(
      race_husband == "Hispanic" & race_wife == "AIAN" ~ 1,
      race_husband == "AIAN" & race_wife == "Hispanic" ~ -1,
      TRUE ~ 0
    )
  )

model_single <- glm(freq~(race_husband + race_wife + 
                            inter_w_b +
                            inter_w_a + 
                            inter_w_i +
                            inter_w_h + 
                            inter_b_a + 
                            inter_b_i +
                            inter_b_h +
                            inter_a_i + 
                            inter_a_h + 
                            inter_i_h + 
                            gender_b_a +
                            gender_b_i + 
                            gender_b_h + 
                            gender_a_i +
                            gender_a_h + 
                            gender_i_h) * year, 
                data = mar_data_single, family = poisson)

inter_lor <- avg_slopes(model_single, 
                        variables = c("inter_w_b", 
                                      "inter_w_a",
                                      "inter_w_i",
                                      "inter_w_h",
                                      "inter_b_a",
                                      "inter_b_i",
                                      "inter_b_h",
                                      "inter_a_i",
                                      "inter_a_h",
                                      "inter_i_h"), 
                        by = "year", 
                        type = "link")
```


```{r}

temp <- inter_lor |>
  as_tibble() |>
  mutate(
    year = as.numeric(paste(year)),
    type = case_when(
      term == "inter_w_b" ~ "White-Black",
      term == "inter_w_a" ~ "White-API",
      term == "inter_w_i" ~ "White-AIAN",
      term == "inter_w_h" ~ "White-Hispanic",
      term == "inter_b_a" ~ "Black-API",
      term == "inter_b_i" ~ "Black-AIAN",
      term == "inter_b_h" ~ "Black-Hispanic",
      term == "inter_a_i" ~ "AIAN-API",
      term == "inter_a_h" ~ "API-Hispanic",
      term == "inter_i_h" ~ "AIAN-Hispanic",
    )
  ) |>
  rename(lor_model = estimate, se_model = std.error) |>
  select(lor_model, se_model, year, type)

temp <- lor_data |> 
  filter(type %in% c("White-Black", "White-API", "White-AIAN", "White-Hispanic",
                     "Black-API", "Black-AIAN", "Black-Hispanic",
                     "API-Hispanic", "AIAN-Hispanic", "AIAN-API")) |>
  rename(lor_cross = lor, se_cross = se) |>
  inner_join(temp)
  
ggplot(temp, aes(x = lor_cross, y = lor_model))+
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "red")+
  geom_point()+
  labs(x = "White/Black log odds ratio by cross-product",
       y = "White/Black lod odds ratio by model")+
  facet_wrap(~type)+
  theme_bw()

ggplot(temp, aes(x = se_cross, y = se_model))+
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "red")+
  geom_point()+
  labs(x = "White/Black log odds ratio standard error by cross-product",
       y = "White/Black lod odds ratio standard error by model")+
  facet_wrap(~type)+
  theme_bw()
```

# try plotting some output

```{r}
inter_lor |>
  as_tibble() |>
  filter(str_detect(term, "_w_")) |>
  mutate(
    year = as.numeric(paste(year)),
    type = case_when(
      term == "inter_w_b" ~ "White-Black",
      term == "inter_w_a" ~ "White-API",
      term == "inter_w_i" ~ "White-AIAN",
      term == "inter_w_h" ~ "White-Hispanic"
    )
  ) |>
  ggplot(aes(x = year, y = exp(estimate), color = type))+
  geom_vline(xintercept = 2000, linetype = 2)+
  geom_ribbon(aes(ymin = exp(conf.low), ymax = exp(conf.high), fill = type),
              alpha = 0.3, color = NA)+
  geom_line()+
  geom_point(size = 0.5)+
  scale_y_log10()+
  theme_bw()

```

# Try automating


```{r}
selected_groups <- c("White", "Black", "AIAN", "API", "Hispanic")

mar_data <- mar_data_full |>
  filter(race_husband %in% selected_groups & race_wife %in% selected_groups) |>
  filter(freq > 0)

for(i in 1:(length(selected_groups)-1)) {
  for(j in (i+1):length(selected_groups)) {
    
    race1 <- selected_groups[i]
    race2 <- selected_groups[j]
    
    var_suffix <- paste(str_to_lower(race1), str_to_lower(race2), sep = "_")
    var_suffix <- str_replace(var_suffix, "/", ".")
    
    var_name <- paste0("inter_", var_suffix)                 
    mar_data <- mar_data |>
      mutate({{ var_name }} := (race_husband == race1 & race_wife == race2))
    
    if(i > 1) {
      # we need a gender adjuster
      var_name <- paste0("gender_", var_suffix)
      mar_data <- mar_data |>
        mutate(
          {{ var_name }} := case_when(
            race_husband == race1 & race_wife == race2 ~ 1,
            race_husband ==  race2 & race_wife ==  race1 ~ -1,
            TRUE ~ 0
          )
        )
    }
  }
}

create_formula <- function(mar_data) {
  terms <- colnames(mar_data)
  terms <- terms[-which(terms == "year")]
  terms <- terms[-which(terms == "freq")]
  terms <- paste0("(", paste(terms, collapse="+"), ")*year")
  reformulate(terms, "freq")
}

model <- glm(create_formula(mar_data), data = mar_data, family = poisson)

terms <- colnames(mar_data)[str_detect(colnames(mar_data), "^inter_")]

inter_lor <- avg_slopes(model, variables = terms, by = "year", type = "link")

inter_lor |>
  as_tibble() |>
  mutate(year = as.numeric(paste(year))) |>
  ggplot(aes(x = year, y = exp(estimate), color = term))+
  geom_vline(xintercept = 2000, linetype = 2)+
  geom_ribbon(aes(ymin = exp(conf.low), ymax = exp(conf.high), fill = term),
              alpha = 0.3, color = NA)+
  geom_line()+
  geom_point(size = 0.5)+
  scale_y_log10()+
  theme_bw()
```

```{r}
calculate_lor_model <- function(selected_groups) {
  
  # trim down the data to selected groups and frequency counts greater than zero
  mar_data <- mar_data_full |>
    filter(race_husband %in% selected_groups & race_wife %in% selected_groups) |>
    filter(freq > 0)
  
  # create required variables
  for(i in 1:(length(selected_groups)-1)) {
    for(j in (i+1):length(selected_groups)) {
      
      race1 <- selected_groups[i]
      race2 <- selected_groups[j]
      
      var_suffix <- paste(str_to_lower(race1), str_to_lower(race2), sep = "_")
      var_suffix <- str_replace_all(var_suffix, "/", ".")
      
      var_name <- paste0("inter_", var_suffix)                 
      mar_data <- mar_data |>
        mutate({{ var_name }} := (race_husband == race1 & race_wife == race2))
      
      if(i > 1) {
        # we need a gender adjuster
        var_name <- paste0("gender_", var_suffix)
        mar_data <- mar_data |>
          mutate(
            {{ var_name }} := case_when(
              race_husband == race1 & race_wife == race2 ~ 1,
              race_husband ==  race2 & race_wife ==  race1 ~ -1,
              TRUE ~ 0
            )
          )
      }
    }
  }

  # create formula
  terms <- colnames(mar_data) |> str_subset("freq|year", negate = TRUE)
  formula_model <- reformulate(paste0("(", paste(terms, collapse="+"), ")*year"),
                               "freq")
  # run the model
  glm(formula_model, data = mar_data, family = poisson)
  
}

get_marg_effects <- function(model, groups) {
  vars <- map_chr(groups, function(x) {
    x <- str_to_lower(x)
    x <- str_replace(x, "/", ".")
    paste0("inter_", paste(x, collapse = "_"))
  })
  
  avg_slopes(model, variables = vars, by = "year", type = "link") |>
    as_tibble() |>
    mutate(year = as.numeric(paste(year)))
}

```

```{r}
selected_groups <- c("White", "Black", "AIAN", "API", "Hispanic")
model <- calculate_lor_model(selected_groups)

get_marg_effects(model, 
                 list(c("White", "Black"),
                      c("White", "API"),
                      c("White", "AIAN"),
                      c("White", "Hispanic"))) |>
  ggplot(aes(x = as.numeric(paste(year)), y = exp(estimate), color = term))+
  geom_vline(xintercept = 2000, linetype = 2)+
  geom_ribbon(aes(ymin = exp(conf.low), ymax = exp(conf.high), fill = term),
              alpha = 0.3, color = NA)+
  geom_line()+
  geom_point(size = 0.5)+
  scale_y_log10()+
  theme_bw()

get_marg_effects(model, 
                 list(c("AIAN", "API"),
                      c("White", "API"))) |>
  ggplot(aes(x = as.numeric(paste(year)), y = exp(estimate), color = term))+
  geom_vline(xintercept = 2000, linetype = 2)+
  geom_hline(yintercept = exp(-9.394), linetype = 3)+
  geom_ribbon(aes(ymin = exp(conf.low), ymax = exp(conf.high), fill = term),
              alpha = 0.3, color = NA)+
  geom_line()+
  geom_point(size = 0.5)+
  scale_y_log10()+
  theme_bw()
```

# All groups

The model can run for the full model, but extracting average slopes from this model takes forever. Its more efficient to simply run the model on the subset that you want.

```{r}
selected_groups <- c("White", "Black", "White/Black")

model <- calculate_lor_model(selected_groups)

get_marg_effects(model, 
                 list(c("White", "Black"),
                      c("White", "White/Black"),
                      c("Black", "White/Black"))) |>
  ggplot(aes(x = as.numeric(paste(year)), y = exp(estimate), color = term))+
  geom_vline(xintercept = 2000, linetype = 2)+
  geom_ribbon(aes(ymin = exp(conf.low), ymax = exp(conf.high), fill = term),
              alpha = 0.3, color = NA)+
  geom_line()+
  geom_point(size = 0.5)+
  scale_y_log10()+
  theme_bw()
```


