---
title: "Check model approach"
format: html
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(here)
source(here("utils","check_packages.R"))
source(here("utils","functions.R"))
load(here("data", "data_constructed", "individ_data.RData"))
```

# Set up the data

Lets start by creating a summary dataset by year and race of spouses

```{r}
#| label: create-agg-mar-data

mar_data <- census |>
  # restrict age of husbands to get recent marriages
  filter(age_husband >= 25 & age_husband <= 34) |>
  mutate(year = factor(year),
         year = relevel(year, "2000")) |>
  group_by(race_husband, race_wife, year, .drop = FALSE) |>
  summarize(freq = n()) |>
  ungroup()
```


Lets also calculate the raw log odds via a cross product

```{r}
#| label: calculate-lor-crossproduct

temp <- census |>
  filter(age_husband >= 25 & age_husband <= 34)

tab_mar <- table(temp$race_husband, temp$race_wife, temp$year)

calculate_or <- function(group1, group2) {
  lor <- log(tab_mar[group1, group2,])+log(tab_mar[group2, group1,])-
    log(tab_mar[group1, group1,])-log(tab_mar[group2, group2,])
  se <- sqrt(1/tab_mar[group1, group2,]+1/tab_mar[group2, group1,]+
               1/tab_mar[group1, group1,]+1/tab_mar[group2, group2,])
  tibble(year = as.numeric(names(lor)), lor, se, 
         type = paste(group1, group2, sep="-"))
}

all_combos <- tribble(
  ~race1, ~race2,
  "White", "Black",
  "White", "AIAN",
  "White", "API",
  "White", "Hispanic",
  "Black", "AIAN",
  "Black", "API",
  "Black", "Hispanic",
  "AIAN", "API",
  "AIAN", "Hispanic",
  "API", "Hispanic",
  "White/Black", "White",
  "White/Black", "Black",
  "White/API", "White",
  "White/API", "Black",
  "White/API", "API",
  "White/AIAN", "White",
  "White/AIAN", "AIAN"
)

lor_data <- pmap(all_combos, function(race1, race2) {
  calculate_or(race1, race2)
  }) |>
  bind_rows() |>
  filter(lor > -Inf)
```

# Model it within a function

Now lets do this within a model.

```{r}
#| label: modeling-function

calculate_lor_model <- function(selected_groups, terms_add = NULL) {
  
  # trim down the data to selected groups and frequency counts greater than zero
  model_data <- mar_data |>
    filter(race_husband %in% selected_groups & race_wife %in% selected_groups) |>
    filter(freq > 0)
  
  # create required variables
  for(i in 1:(length(selected_groups)-1)) {
    for(j in (i+1):length(selected_groups)) {
      
      race1 <- selected_groups[i]
      race2 <- selected_groups[j]
      
      var_suffix <- paste(str_to_lower(race1), str_to_lower(race2), sep = "_")
      var_suffix <- str_replace_all(var_suffix, "/", ".")
      
      var_name <- paste0("inter_", var_suffix)                 
      model_data <- model_data |>
        mutate({{ var_name }} := (race_husband == race1 & race_wife == race2))
      
      # for cases not involving the reference category, we need to add a 
      # variable for the two gender combinations with 1 and -1 coding so that 
      # the general effect is the average of the two
      if(i > 1) {
        # we need a gender adjuster
        var_name <- paste0("gender_", var_suffix)
        model_data <- model_data |>
          mutate(
            {{ var_name }} := case_when(
              race_husband == race1 & race_wife == race2 ~ 1,
              race_husband ==  race2 & race_wife ==  race1 ~ -1,
              TRUE ~ 0
            )
          )
      }
    }
  }

  # create formula
  terms <- colnames(model_data) |> str_subset("freq|year", negate = TRUE)
  terms_str <- paste(terms, collapse="+")
  if(!is.null(terms_add)) {
    terms_str <- paste(terms_str, terms_add, sep = "+")
  }
  formula_model <- reformulate(paste0("(", terms_str, ")*year"),
                               "freq")
  # run the model
  model <- glm(formula_model, data = model_data, family = poisson)
  
  # identify results that are missing due to zero values
  missing <- coef(model) |> 
    enframe(name = "variable", value = "coef") |>
    filter(is.na(coef), 
           str_detect(variable, "^inter_")) |>
    mutate(term = str_split_i(variable, "TRUE", 1),
           year = as.numeric(str_sub(variable, start = -4)),
           year = if_else(is.na(year), 2000, year),
           missing = TRUE) |>
    select(term, year, missing)
    
  marg <- avg_slopes(model, 
                     variables = str_subset(terms, "^inter_"),
                     by = "year",
                     type = "link") |>
    as_tibble() |>
    mutate(year = as.numeric(paste(year))) |> 
    left_join(missing) |>
    filter(is.na(missing)) |>
    select(-missing)
  
  return(marg)
}
```

Try it out for a few cases:

```{r}
#| label: fig-five-single
#| fig-cap: Log odds of interracial marriage with white partners

selected_groups <- c("White", "Black", "AIAN", "API", "Hispanic")
calculate_lor_model(selected_groups) |>
  filter(str_detect(term, "_white_")) |>
  ggplot(aes(x = as.numeric(paste(year)), y = exp(estimate), color = term))+
  geom_vline(xintercept = 2000, linetype = 2)+
  geom_ribbon(aes(ymin = exp(conf.low), ymax = exp(conf.high), fill = term),
              alpha = 0.3, color = NA)+
  geom_line()+
  geom_point(size = 0.5)+
  scale_y_log10()+
  theme_bw()
```

```{r}
#| label: fig-aian-intermar
#| fig-cap: Log odds of interracial marriage with AIAN partners

calculate_lor_model(selected_groups) |>
  filter(str_detect(term, "_aian")) |>
  ggplot(aes(x = as.numeric(paste(year)), y = exp(estimate), color = term))+
  geom_vline(xintercept = 2000, linetype = 2)+
  geom_ribbon(aes(ymin = exp(conf.low), ymax = exp(conf.high), fill = term),
              alpha = 0.3, color = NA)+
  geom_line()+
  geom_point(size = 0.5)+
  scale_y_log10()+
  theme_bw()
```

```{r}
#| label: fig-wb-intermar
#| fig-cap: Log odds of interracial marriage between White, Black, and White/Black partners

calculate_lor_model(c("White", "Black", "White/Black")) |>
  ggplot(aes(x = as.numeric(paste(year)), y = exp(estimate), color = term))+
  geom_vline(xintercept = 2000, linetype = 2)+
  geom_ribbon(aes(ymin = exp(conf.low), ymax = exp(conf.high), fill = term),
              alpha = 0.3, color = NA)+
  geom_line()+
  geom_point(size = 0.5)+
  scale_y_log10()+
  theme_bw()
```


# Test it out

## Two-group case

Start with two groups. Does it match results from lor by cross-product?

```{r}
#| label: fig-two-group
#| fig-cap: Comparison of log odds ratio and standard errors for White/Black intermarriage by model and cross-product
#| layout-ncol: 2
#| fig-subcap:
#|    - Estimate
#|    - Standard Error

wb_lor <- calculate_lor_model(c("White", "Black")) |>
  rename(lor_model = estimate, se_model = std.error) |>
  select(lor_model, se_model, year)

temp <- lor_data |> 
  filter(type == "White-Black") |>
  rename(lor_cross = lor, se_cross = se) |>
  select(-type) |>
  inner_join(wb_lor)

ggplot(temp, aes(x = lor_cross, y = lor_model))+
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "red")+
  geom_point()+
  labs(x = "White/Black log odds ratio by cross-product",
       y = "White/Black lod odds ratio by model")+
  theme_bw()

ggplot(temp, aes(x = se_cross, y = se_model))+
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "red")+
  geom_point()+
  labs(x = "White/Black log odds ratio standard error by cross-product",
       y = "White/Black lod odds ratio standard error by model")+
  theme_bw()
```

Perfect fit.

## Three-group case

We now have a spare parameter to fit.

```{r}
#| label: fig-three-group
#| fig-cap: Comparison of log odds ratio and standard errors for White/Black, Black/API, and White/API intermarriage by model and cross-product
#| layout-ncol: 2
#| fig-subcap:
#|    - Estimate
#|    - Standard Error

inter_lor <- calculate_lor_model(c("White", "Black", "API")) |>
  mutate(type = case_when(
    term == "inter_black_api" ~ "Black-API",
    term == "inter_white_black" ~ "White-Black",
    term == "inter_white_api" ~ "White-API"
  )) |>
  rename(lor_model = estimate, se_model = std.error) |>
  select(type, lor_model, se_model, year)

temp <- lor_data |> 
  filter(type %in% c("White-Black", "White-API", "Black-API")) |>
  rename(lor_cross = lor, se_cross = se) |>
  inner_join(inter_lor)

ggplot(temp, aes(x = lor_cross, y = lor_model))+
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "red")+
  geom_point()+
  labs(x = "log odds ratio by cross-product",
       y = "log odds ratio by model")+
  facet_wrap(~type)+
  theme_bw()

ggplot(temp, aes(x = se_cross, y = se_model))+
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "red")+
  geom_point()+
  labs(x = "log odds ratio standard error by cross-product",
       y = "log odds ratio standard error by model")+
  facet_wrap(~type)+
  theme_bw()
```

perfect fit again! Its working.

## Four-group case

Lets put in a multiracial group this time.

```{r}
#| label: fig-four-group
#| fig-cap: Comparison of log odds ratio and standard errors for several intermarriages by model and cross-product
#| layout-ncol: 2
#| fig-height: 7
#| fig-subcap:
#|    - Estimate
#|    - Standard Error

inter_lor <- calculate_lor_model(c("White", "Black", "API", "White/API")) |>
  mutate(type = case_when(
    term == "inter_black_api" ~ "Black-API",
    term == "inter_white_black" ~ "White-Black",
    term == "inter_white_api" ~ "White-API",
    term == "inter_white_white.api" ~ "White/API-White",
    term == "inter_api_white.api" ~ "White/API-API",
    term == "inter_black_white.api" ~ "White/API-Black"
  )) |>
  rename(lor_model = estimate, se_model = std.error) |>
  select(type, lor_model, se_model, year)

temp <- lor_data |> 
  filter(type %in% c("White-Black", "White-API", "Black-API", 
                     "White/API-White", "White/API-API", "White/API-Black")) |>
  rename(lor_cross = lor, se_cross = se) |>
  inner_join(inter_lor)

ggplot(temp, aes(x = lor_cross, y = lor_model))+
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "red")+
  geom_point()+
  labs(x = "log odds ratio by cross-product",
       y = "log odds ratio by model")+
  facet_wrap(~type)+
  theme_bw()

ggplot(temp, aes(x = se_cross, y = se_model))+
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "red")+
  geom_point()+
  labs(x = "log odds ratio standard error by cross-product",
       y = "log odds ratio standard error by model")+
  facet_wrap(~type)+
  theme_bw()
```

# Add covariates

```{r}

selected_groups <- c("White", "Black", "AIAN", "API", "Hispanic")
results_single_no_controls <- calculate_lor_model(selected_groups) |>
  mutate(type = "no controls")

```


```{r}
#| label: create-agg-mar-data-covar

mar_data <- census |>
  # restrict age of husbands to get recent marriages
  filter(age_husband >= 25 & age_husband <= 34) |>
  mutate(year = factor(year),
         year = relevel(year, "2000")) |>
  group_by(race_husband, race_wife, educ_husband, educ_wife, year, .drop = FALSE) |>
  summarize(freq = n()) |>
  ungroup()
```

```{r}
results_single_ed_controls <- calculate_lor_model( 
  selected_groups,
  terms_add = "educ_husband*race_husband+educ_wife*race_wife+educ_husband*educ_wife"
)

results_single <- results_single_ed_controls |>
  mutate(type = "ed controls") |> 
  bind_rows(results_single_no_controls)

results_single |>
  filter(str_detect(term, "_white_")) |>
  mutate(type = factor(type, levels = c("no controls", "ed controls"))) |>
  ggplot(aes(x = as.numeric(paste(year)), y = exp(estimate), color = term, linetype = type))+
  geom_vline(xintercept = 2000, linetype = 2)+
  geom_ribbon(aes(ymin = exp(conf.low), ymax = exp(conf.high), fill = term),
              alpha = 0.3, color = NA)+
  geom_line()+
  geom_point(size = 0.5)+
  scale_y_log10()+
  theme_bw()

```
