---
title: "Check model approach"
format: html
---

```{r}
#| label: setup
#| include: false
library(here)
source(here("utils","check_packages.R"))
source(here("utils","functions.R"))
load(here("data", "data_constructed", "individ_data.RData"))
```


Lets start by creating a summary dataset by year and race of spouses

```{r}
mar_data <- census |>
  # restrict age of husbands to get recent marriages
  filter(age_husband >= 25 & age_husband <= 34) |>
  mutate(year = factor(year),
         year = relevel(year, "2000")) |>
  group_by(race_husband, race_wife, year, .drop = FALSE) |>
  summarize(freq = n()) |>
  ungroup()
```


Lets create a function that will calculate the log-odds ratio and its standard error from `tab-mar` for two groups.

```{r}
temp <- census |>
  filter(age_husband >= 25 & age_husband <= 34)

tab_mar <- table(census$race_husband, census$race_wife, census$year)

calculate_or <- function(group1, group2) {
  lor <- log(tab_mar[group1, group2,])+log(tab_mar[group2, group1,])-
    log(tab_mar[group1, group1,])-log(tab_mar[group2, group2,])
  se <- sqrt(1/tab_mar[group1, group2,]+1/tab_mar[group2, group1,]+
               1/tab_mar[group1, group1,]+1/tab_mar[group2, group2,])
  tibble(year = as.numeric(names(lor)), lor, se, 
         type = paste(group1, group2, sep="-"))
}
```

We can now use `pmap` to calculate log odds ratios across all groups of interest.

```{r}
#| label: get-lor-df

all_combos <- tribble(
  ~race1, ~race2,
  "White", "Black",
  "White", "AIAN",
  "White", "API",
  "White", "Hispanic",
  "Black", "AIAN",
  "Black", "API",
  "Black", "Hispanic",
  "AIAN", "API",
  "AIAN", "Hispanic",
  "API", "Hispanic",
  "White/Black", "White",
  "White/Black", "Black",
  "White/API", "White",
  "White/API", "API",
  "White/AIAN", "White",
  "White/AIAN", "AIAN"
)

lor_data <- pmap(all_combos, function(race1, race2) {
  calculate_or(race1, race2)
  }) |>
  bind_rows() |>
  filter(lor > -Inf)
```

Now lets try a model approach.

First convert to a frequency based dataset

```{r}
mar_data_full <- as.data.frame.table(tab_mar) |>
  tibble() |>
  rename(race_husband = Var1, race_wife = Var2, year = Var3, freq = Freq) |>
  filter(year != 1960)
```





# All single race groups

Sparseness problems are created for years where we get zero cells. This causes all of the SEs in the AMEs to be wrong. Filtering out frequencies of zero seems to fix the problem. The `avg_slopes` gives a warning about rank deficient matrices but it fits everything fine.

```{r}
mar_data_single <- mar_data_full |>
  #filter(year != "1970") |>
  filter(!str_detect(race_husband, "/|Other"), 
         !str_detect(race_wife, "/|Other")) |>
  filter(freq > 0) |>
  mutate(
    inter_w_b = (race_husband == "White" & race_wife == "Black"),
    inter_w_a = (race_husband == "White" & race_wife == "API"),
    inter_w_i = (race_husband == "White" & race_wife == "AIAN"),
    inter_w_h = (race_husband == "White" & race_wife == "Hispanic"),
    inter_b_a = (race_husband == "Black" & race_wife == "API"),
    inter_b_i = (race_husband == "Black" & race_wife == "AIAN"),
    inter_b_h = (race_husband == "Black" & race_wife == "Hispanic"),
    inter_a_i = (race_husband == "API" & race_wife == "AIAN"),
    inter_a_h = (race_husband == "API" & race_wife == "Hispanic"),
    inter_i_h = (race_husband == "AIAN" & race_wife == "Hispanic"),
    gender_b_a = case_when(
      race_husband == "API" & race_wife == "Black" ~ 1,
      race_husband == "Black" & race_wife == "API" ~ -1,
      TRUE ~ 0
    ),
    gender_b_i = case_when(
      race_husband == "AIAN" & race_wife == "Black" ~ 1,
      race_husband == "Black" & race_wife == "AIAN" ~ -1,
      TRUE ~ 0
    ),
    gender_b_h = case_when(
      race_husband == "Hispanic" & race_wife == "Black" ~ 1,
      race_husband == "Black" & race_wife == "Hispanic" ~ -1,
      TRUE ~ 0
    ),
    gender_a_i = case_when(
      race_husband == "AIAN" & race_wife == "API" ~ 1,
      race_husband == "API" & race_wife == "AIAN" ~ -1,
      TRUE ~ 0
    ),
    gender_a_h = case_when(
      race_husband == "Hispanic" & race_wife == "API" ~ 1,
      race_husband == "API" & race_wife == "Hispanic" ~ -1,
      TRUE ~ 0
    ),
    gender_i_h = case_when(
      race_husband == "Hispanic" & race_wife == "AIAN" ~ 1,
      race_husband == "AIAN" & race_wife == "Hispanic" ~ -1,
      TRUE ~ 0
    )
  )

model_single <- glm(freq~(race_husband + race_wife + 
                            inter_w_b +
                            inter_w_a + 
                            inter_w_i +
                            inter_w_h + 
                            inter_b_a + 
                            inter_b_i +
                            inter_b_h +
                            inter_a_i + 
                            inter_a_h + 
                            inter_i_h + 
                            gender_b_a +
                            gender_b_i + 
                            gender_b_h + 
                            gender_a_i +
                            gender_a_h + 
                            gender_i_h) * year, 
                data = mar_data_single, family = poisson)

inter_lor <- avg_slopes(model_single, 
                        variables = c("inter_w_b", 
                                      "inter_w_a",
                                      "inter_w_i",
                                      "inter_w_h",
                                      "inter_b_a",
                                      "inter_b_i",
                                      "inter_b_h",
                                      "inter_a_i",
                                      "inter_a_h",
                                      "inter_i_h"), 
                        by = "year", 
                        type = "link")
```


```{r}

temp <- inter_lor |>
  as_tibble() |>
  mutate(
    year = as.numeric(paste(year)),
    type = case_when(
      term == "inter_w_b" ~ "White-Black",
      term == "inter_w_a" ~ "White-API",
      term == "inter_w_i" ~ "White-AIAN",
      term == "inter_w_h" ~ "White-Hispanic",
      term == "inter_b_a" ~ "Black-API",
      term == "inter_b_i" ~ "Black-AIAN",
      term == "inter_b_h" ~ "Black-Hispanic",
      term == "inter_a_i" ~ "AIAN-API",
      term == "inter_a_h" ~ "API-Hispanic",
      term == "inter_i_h" ~ "AIAN-Hispanic",
    )
  ) |>
  rename(lor_model = estimate, se_model = std.error) |>
  select(lor_model, se_model, year, type)

temp <- lor_data |> 
  filter(type %in% c("White-Black", "White-API", "White-AIAN", "White-Hispanic",
                     "Black-API", "Black-AIAN", "Black-Hispanic",
                     "API-Hispanic", "AIAN-Hispanic", "AIAN-API")) |>
  rename(lor_cross = lor, se_cross = se) |>
  inner_join(temp)
  
ggplot(temp, aes(x = lor_cross, y = lor_model))+
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "red")+
  geom_point()+
  labs(x = "White/Black log odds ratio by cross-product",
       y = "White/Black lod odds ratio by model")+
  facet_wrap(~type)+
  theme_bw()

ggplot(temp, aes(x = se_cross, y = se_model))+
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "red")+
  geom_point()+
  labs(x = "White/Black log odds ratio standard error by cross-product",
       y = "White/Black lod odds ratio standard error by model")+
  facet_wrap(~type)+
  theme_bw()
```

# try plotting some output

```{r}
inter_lor |>
  as_tibble() |>
  filter(str_detect(term, "_w_")) |>
  mutate(
    year = as.numeric(paste(year)),
    type = case_when(
      term == "inter_w_b" ~ "White-Black",
      term == "inter_w_a" ~ "White-API",
      term == "inter_w_i" ~ "White-AIAN",
      term == "inter_w_h" ~ "White-Hispanic"
    )
  ) |>
  ggplot(aes(x = year, y = exp(estimate), color = type))+
  geom_vline(xintercept = 2000, linetype = 2)+
  geom_ribbon(aes(ymin = exp(conf.low), ymax = exp(conf.high), fill = type),
              alpha = 0.3, color = NA)+
  geom_line()+
  geom_point(size = 0.5)+
  scale_y_log10()+
  theme_bw()

```

# Model within a function

```{r}
calculate_lor_model <- function(selected_groups) {
  
  # trim down the data to selected groups and frequency counts greater than zero
  model_data <- mar_data |>
    filter(race_husband %in% selected_groups & race_wife %in% selected_groups) |>
    filter(freq > 0)
  
  # create required variables
  for(i in 1:(length(selected_groups)-1)) {
    for(j in (i+1):length(selected_groups)) {
      
      race1 <- selected_groups[i]
      race2 <- selected_groups[j]
      
      var_suffix <- paste(str_to_lower(race1), str_to_lower(race2), sep = "_")
      var_suffix <- str_replace_all(var_suffix, "/", ".")
      
      var_name <- paste0("inter_", var_suffix)                 
      model_data <- model_data |>
        mutate({{ var_name }} := (race_husband == race1 & race_wife == race2))
      
      # for cases not involving the reference category, we need to add a 
      # variable for the two gender combinations with 1 and -1 coding so that 
      # the general effect is the average of the two
      if(i > 1) {
        # we need a gender adjuster
        var_name <- paste0("gender_", var_suffix)
        model_data <- model_data |>
          mutate(
            {{ var_name }} := case_when(
              race_husband == race1 & race_wife == race2 ~ 1,
              race_husband ==  race2 & race_wife ==  race1 ~ -1,
              TRUE ~ 0
            )
          )
      }
    }
  }

  # create formula
  terms <- colnames(model_data) |> str_subset("freq|year", negate = TRUE)
  formula_model <- reformulate(paste0("(", paste(terms, collapse="+"), ")*year"),
                               "freq")
  # run the model
  model <- glm(formula_model, data = model_data, family = poisson)
  
  # identify results that are missing due to zero values
  missing <- coef(model) |> 
    enframe(name = "variable", value = "coef") |>
    filter(is.na(coef), 
           str_detect(variable, "^inter_")) |>
    mutate(term = str_split_i(variable, "TRUE", 1),
           year = as.numeric(str_sub(variable, start = -4)),
           year = if_else(is.na(year), 2000, year),
           missing = TRUE) |>
    select(term, year, missing)
    
  marg <- avg_slopes(model, 
                     variables = str_subset(terms, "^inter_"),
                     by = "year",
                     type = "link") |>
    as_tibble() |>
    mutate(year = as.numeric(paste(year))) |> 
    left_join(missing) |>
    filter(is.na(missing)) |>
    select(-missing)
  
  return(marg)
}
```

Try it out for a few cases:

```{r}
selected_groups <- c("White", "Black", "AIAN", "API", "Hispanic")
calculate_lor_model(selected_groups) |>
  filter(str_detect(term, "_white_")) |>
  ggplot(aes(x = as.numeric(paste(year)), y = exp(estimate), color = term))+
  geom_vline(xintercept = 2000, linetype = 2)+
  geom_ribbon(aes(ymin = exp(conf.low), ymax = exp(conf.high), fill = term),
              alpha = 0.3, color = NA)+
  geom_line()+
  geom_point(size = 0.5)+
  scale_y_log10()+
  theme_bw()

calculate_lor_model(selected_groups) |>
  filter(str_detect(term, "_aian_")) |>
  ggplot(aes(x = as.numeric(paste(year)), y = exp(estimate), color = term))+
  geom_vline(xintercept = 2000, linetype = 2)+
  geom_ribbon(aes(ymin = exp(conf.low), ymax = exp(conf.high), fill = term),
              alpha = 0.3, color = NA)+
  geom_line()+
  geom_point(size = 0.5)+
  scale_y_log10()+
  theme_bw()

calculate_lor_model(c("White", "Black", "White/Black")) |>
  ggplot(aes(x = as.numeric(paste(year)), y = exp(estimate), color = term))+
  geom_vline(xintercept = 2000, linetype = 2)+
  geom_ribbon(aes(ymin = exp(conf.low), ymax = exp(conf.high), fill = term),
              alpha = 0.3, color = NA)+
  geom_line()+
  geom_point(size = 0.5)+
  scale_y_log10()+
  theme_bw()
```


# Test it out

## Two-group case

Start with two groups. Does it match results from lor by cross-product?

```{r}
wb_lor <- calculate_lor_model(c("White", "Black")) |>
  rename(lor_model = estimate, se_model = std.error) |>
  select(lor_model, se_model, year)

temp <- lor_data |> 
  filter(type == "White-Black") |>
  rename(lor_cross = lor, se_cross = se) |>
  select(-type) |>
  inner_join(wb_lor)

ggplot(temp, aes(x = lor_cross, y = lor_model))+
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "red")+
  geom_point()+
  labs(x = "White/Black log odds ratio by cross-product",
       y = "White/Black lod odds ratio by model")+
  theme_bw()

ggplot(temp, aes(x = se_cross, y = se_model))+
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "red")+
  geom_point()+
  labs(x = "White/Black log odds ratio standard error by cross-product",
       y = "White/Black lod odds ratio standard error by model")+
  theme_bw()
```

Perfect fit.

## Three-group case

We now have a spare parameter to fit.

```{r}
inter_lor <- calculate_lor_model(c("White", "Black", "API")) |>
  mutate(type = case_when(
    term == "inter_black_api" ~ "Black-API",
    term == "inter_white_black" ~ "White-Black",
    term == "inter_white_api" ~ "White-API"
  )) |>
  rename(lor_model = estimate, se_model = std.error) |>
  select(type, lor_model, se_model, year)

temp <- lor_data |> 
  filter(type %in% c("White-Black", "White-API", "Black-API")) |>
  rename(lor_cross = lor, se_cross = se) |>
  inner_join(inter_lor)

ggplot(temp, aes(x = lor_cross, y = lor_model))+
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "red")+
  geom_point()+
  labs(x = "log odds ratio by cross-product",
       y = "log odds ratio by model")+
  facet_wrap(~type)+
  theme_bw()

ggplot(temp, aes(x = se_cross, y = se_model))+
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "red")+
  geom_point()+
  labs(x = "log odds ratio standard error by cross-product",
       y = "log odds ratio standard error by model")+
  facet_wrap(~type)+
  theme_bw()
```

perfect fit again! Its working.

## Four-group case

Lets put in a multiracial group this time.

```{r}
inter_lor <- calculate_lor_model(c("White", "Black", "API", "White/API")) |>
  mutate(type = case_when(
    term == "inter_black_api" ~ "Black-API",
    term == "inter_white_black" ~ "White-Black",
    term == "inter_white_api" ~ "White-API",
    term == "inter_white_white.api" ~ "White/API-White",
    term == "inter_api_white.api" ~ "White/API-API",
    term == "inter_black_white.api" ~ "White/API-Black"
  )) |>
  rename(lor_model = estimate, se_model = std.error) |>
  select(type, lor_model, se_model, year)

temp <- lor_data |> 
  filter(type %in% c("White-Black", "White-API", "Black-API", 
                     "White/API-White", "White/API-API", "White/API-Black")) |>
  rename(lor_cross = lor, se_cross = se) |>
  inner_join(inter_lor)

ggplot(temp, aes(x = lor_cross, y = lor_model))+
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "red")+
  geom_point()+
  labs(x = "log odds ratio by cross-product",
       y = "log odds ratio by model")+
  facet_wrap(~type)+
  theme_bw()

ggplot(temp, aes(x = se_cross, y = se_model))+
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "red")+
  geom_point()+
  labs(x = "log odds ratio standard error by cross-product",
       y = "log odds ratio standard error by model")+
  facet_wrap(~type)+
  theme_bw()
```