---
title: "Estimate Models"
params:
  n_workers: 5
format: html
---

```{r}
#| label: timestamp-start
#| echo: false

timestamp()
```

```{r}
#| label: setup
#| include: false
library(here)
source(here("utils","check_packages.R"))
source(here("utils","functions.R"))
```

```{r}
#| label: load-state-model-data
#| include: false

load(here("data", "data_constructed", "data_model_state.RData"))
```

This script will estimate all of the important models via bootstrap. It depends on having already run `buld_model_data.qmd` to create lists of the model data required to estimate the models. This speeds up the estimation procedure enormously here. In addition, I use `future` and `furrr` with multiple workers to speed up the process further. 

```{r}
#| label: set-future-plans

if(interactive()) {
  plan(multisession, workers = min(detectCores() - 1, params$n_workers))
} else {
  plan(multicore, workers = min(detectCores() - 1, params$n_workers))
}
```

# Overall models

These models will be estimated for the collapsed five major race groups and five multiracial groups. I estimate all of the single race groups together to get all 10 possible combinations. I estimate each constituent pairing of the single-to-multi interracial marriages separately to avoid taking time to estimate a bunch of pairs that I won't analyze (due too much sampling error largely).

All models adjust for state-level racial composition.

```{r}
#| label: estimate-state-model-overall

# collapse race
tic()
data_model_state_simple <- data_model_state |> 
  future_map(function(model_data) {
    model_data |>
      mutate(race_husband = collapse_race(race_husband),
             race_wife = collapse_race(race_wife)) |>
      group_by(race_husband, race_wife, year, state, region) |>
      summarize(freq = sum(freq), .groups = "drop")
  })
toc()


# first do single race groups
tic()
model_state_bs <- data_model_state_simple |>
  bootstrap_model(c("White", "Black", "AIAN", "API", "Hispanic"), 
                  composition_var = "state", year_separate = TRUE)
toc()

# now add in each multiracial pairwise comparison
tic()
model_state_bs <- data_model_state_simple |>
  bootstrap_model(c("White", "White-Black"),
                  composition_var = "state", year_separate = TRUE,
                  chunk_size = 250) |>
  bind_rows(model_state_bs)
toc()

tic()
model_state_bs <- data_model_state_simple |>
  bootstrap_model(c("Black", "White-Black"),
                  composition_var = "state", year_separate = TRUE,
                  chunk_size = 250) |>
  bind_rows(model_state_bs)
toc()

tic()
model_state_bs <- data_model_state_simple |>
  bootstrap_model(c("White", "White-AIAN"),
                  composition_var = "state", year_separate = TRUE,
                  chunk_size = 250) |>
  bind_rows(model_state_bs)
toc()

tic()
model_state_bs <- data_model_state_simple |>
  bootstrap_model(c("AIAN", "White-AIAN"),
                  composition_var = "state", year_separate = TRUE,
                  chunk_size = 250) |>
  bind_rows(model_state_bs)
toc()

tic()
model_state_bs <- data_model_state_simple |>
  bootstrap_model(c("White", "White-API"),
                  composition_var = "state", year_separate = TRUE,
                  chunk_size = 250) |>
  bind_rows(model_state_bs)
toc()

tic()
model_state_bs <- data_model_state_simple |>
  bootstrap_model(c("API", "White-API"),
                  composition_var = "state", year_separate = TRUE,
                  chunk_size = 250) |>
  bind_rows(model_state_bs)
toc()

tic()
model_state_bs <- data_model_state_simple |>
  bootstrap_model(c("Black", "Black-AIAN"),
                  composition_var = "state", year_separate = TRUE,
                  chunk_size = 250) |>
  bind_rows(model_state_bs)
toc()

tic()
model_state_bs <- data_model_state_simple |>
  bootstrap_model(c("AIAN", "Black-AIAN"),
                  composition_var = "state", year_separate = TRUE,
                  chunk_size = 250) |>
  bind_rows(model_state_bs)
toc()

tic()
model_state_bs <- data_model_state_simple |>
  bootstrap_model(c("Black", "Black-API"),
                  composition_var = "state", year_separate = TRUE,
                  chunk_size = 250) |>
  bind_rows(model_state_bs)
toc()

tic()
model_state_bs <- data_model_state_simple |>
  bootstrap_model(c("API", "Black-API"),
                  composition_var = "state", year_separate = TRUE,
                  chunk_size = 250) |>
  bind_rows(model_state_bs)
toc()

model_state_bs <- model_state_bs |>
  arrange(term, year)

save(model_state_bs, 
     file = here("data", "data_constructed", "model_state_bs.RData"))
```

I will also do a model that does not control for state.

```{r}
#| label: estimate-basic-model

tic()
model_basic_bs <- data_model_state_simple |>
  future_map(function(model_data) {
    model_data |>
      group_by(race_husband, race_wife, year) |>
      summarize(freq = sum(freq), .groups = "drop")
  }) |>
  bootstrap_model(c("White", "Black", "AIAN", "API", "Hispanic"))
toc()

save(model_basic_bs, 
     file = here("data", "data_constructed", "model_basic_bs.RData"))
```

# Generational models

Now lets get models for the single race groups that:

* use all generations including 1st
* use 1.5 and 2nd+ generations

```{r}
#| label: load-gen-model-data
#| include: false

load(here("data", "data_constructed", "data_model_state_gen.RData"))
```


```{r}
#| label: estimate-generational-models

tic()
model_state_single_all_gen_bs <- data_model_state_all_gen |> 
  future_map(function(model_data) {
    model_data |>
      mutate(race_husband = collapse_race(race_husband),
             race_wife = collapse_race(race_wife)) |>
      group_by(race_husband, race_wife, year, state) |>
      summarize(freq = sum(freq), .groups = "drop")
  }) |>
  bootstrap_model(c("White", "Black", "AIAN", "API", "Hispanic"), 
                  composition_var = "state", year_separate = TRUE)
toc()

tic()
model_state_single_no_first_gen_bs <- data_model_state_no_first_gen |> 
  future_map(function(model_data) {
    model_data |>
      mutate(race_husband = collapse_race(race_husband),
             race_wife = collapse_race(race_wife)) |>
      group_by(race_husband, race_wife, year, state) |>
      summarize(freq = sum(freq), .groups = "drop")
  }) |>
  bootstrap_model(c("White", "Black", "AIAN", "API", "Hispanic"), 
                  composition_var = "state", year_separate = TRUE)
toc()

save(model_state_single_all_gen_bs, model_state_single_no_first_gen_bs,
     file = here("data", "data_constructed", "models_state_gen_bs.RData"))
rm(data_model_state_no_first_gen, data_model_state_all_gen)
gc()
```

# Cohabitation models

```{r}
#| label: load-cohab-model-data
#| include: false

load(here("data", "data_constructed", "data_model_state_cohab.RData"))
```

```{r}
#| label: estimate-cohab-model

# collapse race
tic()
data_model_state_cohab_simple <- data_model_state_cohab |> 
  future_map(function(model_data) {
    model_data |>
      mutate(race_husband = collapse_race(race_husband),
             race_wife = collapse_race(race_wife)) |>
      group_by(race_husband, race_wife, year, state) |>
      summarize(freq = sum(freq), .groups = "drop")
  })
toc()

# we don't need the original cohab data any more
rm(data_model_state_cohab)
gc()

tic()
model_state_cohab_bs <- data_model_state_cohab_simple |>
  bootstrap_model(c("White", "Black", "AIAN", "API", "Hispanic"), 
                  composition_var = "state", year_separate = TRUE)
toc()
```

```{r}
#| label: estimate-combine-model

# collapse race
tic()
data_model_state_combine_simple <- map2(
  data_model_state_simple, 
  data_model_state_cohab_simple, 
  bind_rows
) |>
  future_map(function(model_data) {
    model_data |>
      group_by(race_husband, race_wife, year, state) |>
      summarize(freq = sum(freq), .groups = "drop")
  })
toc()

# we don't need the separate cohab data anymore
rm(data_model_state_cohab_simple)
gc()

tic()
model_state_combine_bs <- data_model_state_combine_simple |>
  bootstrap_model(c("White", "Black", "AIAN", "API", "Hispanic"), 
                  composition_var = "state", year_separate = TRUE)
toc()
```


```{r}
#| label: save-cohab-models

save(model_state_cohab_bs, model_state_combine_bs,
     file = here("data", "data_constructed", "models_cohab.RData"))

# memory cleanup
rm(data_model_state_combine_simple)
gc()
```

# Regional models

These models are like the overall models for single race groups, but I allow each log odds ratio to vary by region as well.

```{r}
#| label: estimate-model-regional

# we can do white, black, hispanic together and the other white/* as pairs
tic()
model_state_byregion_bs <- data_model_state_simple |>
  bootstrap_model(c("White", "Black", "Hispanic"),
                  composition_var = "state", conditional_var = "region",
                  year_separate = TRUE, chunk_size = 10)
toc()

tic()
model_state_byregion_bs <- data_model_state_simple |>
  bootstrap_model(c("White", "AIAN"),
                  composition_var = "state", conditional_var = "region",
                  year_separate = TRUE, chunk_size = 25) |>
  bind_rows(model_state_byregion_bs)
toc()

tic()
model_state_byregion_bs <- data_model_state_simple |>
  bootstrap_model(c("White", "API"),
                  composition_var = "state", conditional_var = "region",
                  year_separate = TRUE, chunk_size = 25) |>
  bind_rows(model_state_byregion_bs)
toc()

save(model_state_byregion_bs, 
     file = here("data", "data_constructed", "model_region.RData"))

# we should be done with data_model_state_simple now
rm(data_model_state_simple)
gc()
```

# Subgroup models

For these models I use the full breakout for Hispanic and API respondents to analyze intermarriage with other single race groups and with each other. 

```{r}
#| label: estimate-state-model-subgroups

## Hispanic subgroups ##
tic()
model_hispanic_sub_state_bs <- data_model_state |>
  bootstrap_model(c("White", "Black", "WhiteHispanic", "BlackHispanic", 
                    "OtherHispanic"),
                  composition_var = "state", year_separate = TRUE)
toc()

## Asian subgroups ##
tic()
model_asian_sub_state_bs <- data_model_state |>
  bootstrap_model(c("White", "EastAsian", "SEAsian", "SouthAsian", "PI"),
                  composition_var = "state", year_separate = TRUE)
toc()

# collapse east and SE asian
collapse_asian <- function(race) {
  new_race <- case_when(
    race == "White" ~ "White",
    race == "SouthAsian" ~ "SouthAsian",
    race %in% c("EastAsian", "SEAsian") ~ "EastAsian") 
  factor(new_race, levels = c("White", "EastAsian", "SouthAsian"))
}

tic()
model_asian_sub_simple_state_bs <- data_model_state |> 
  future_map(function(model_data) {
    model_data |>
      mutate(race_husband = collapse_asian(race_husband),
             race_wife = collapse_asian(race_wife)) |>
      group_by(race_husband, race_wife, year, state) |>
      summarize(freq = sum(freq), .groups = "drop")
  }) |>
  bootstrap_model(c("White", "EastAsian", "SouthAsian", "PI"),
                  composition_var = "state", year_separate = TRUE)
toc()

## Indigenous subgroups ##
tic()
model_indig_sub_state_bs <- data_model_state |>
  bootstrap_model(c("White", "AIAN", "PI"),
                  composition_var = "state", year_separate = TRUE)
toc()

save(model_hispanic_sub_state_bs,
     model_asian_sub_state_bs,
     model_asian_sub_simple_state_bs,
     model_indig_sub_state_bs,
     file = here("data", "data_constructed", "models_subgroups.RData"))
```

# Counterfactual models

In these models, I test counterfactual cases for part-White multiple race respondents. I consider:

1. What if they had all responded as White?
2. What if they had all responded as the non-White race?
3. What if they had selected the best race used by IPUMS?

The first two can be done by changing values on spouses's race. The best race model requires the use of a separate model data list.

```{r}
#| label: load-best-model-data
#| include: false

load(here("data", "data_constructed", "data_model_best_race_state.RData"))
```

```{r}
#| label: estimate-state-model-counterfactual

## best race model ##
tic()
model_cf_best_state_bs <- data_model_best_race_state |>
  bootstrap_model(c("White", "Black", "AIAN", "API"),
                  composition_var = "state", year_separate = TRUE) |>
  mutate(counterfactual = "Best")
toc()

counterfactual_race <- function(race, cf) {
  
  race <- collapse_race(race)
  
  if(cf == "white") {
    race <- if_else(race == "White-Black", "White", race)
    race <- if_else(race == "White-API", "White", race)
    race <- if_else(race == "White-AIAN", "White", race)
  }
  if(cf == "minority") {
    race <- if_else(race == "White-Black", "Black", race)
    race <- if_else(race == "White-API", "API", race)
    race <- if_else(race == "White-AIAN", "AIAN", race)
  }

  return(race)

}

## white model ##
tic()
model_cf_white_state_bs <- data_model_state |>
  future_map(function(model_data) {
    model_data |>
      mutate(race_husband = counterfactual_race(race_husband, "white"),
             race_wife = counterfactual_race(race_wife, "white")) |>
      group_by(race_husband, race_wife, year, state) |>
      summarize(freq = sum(freq), .groups = "drop")
  }) |>
  bootstrap_model(c("White", "Black", "AIAN", "API"),
                  composition_var = "state", year_separate = TRUE) |>
  mutate(counterfactual = "White")
toc()

## minority model ##
tic()
model_cf_minority_state_bs <- data_model_state |>
  future_map(function(model_data) {
    model_data |>
      mutate(race_husband = counterfactual_race(race_husband, "minority"),
             race_wife = counterfactual_race(race_wife, "minority")) |>
      group_by(race_husband, race_wife, year, state) |>
      summarize(freq = sum(freq), .groups = "drop")
  }) |>
  bootstrap_model(c("White", "Black", "AIAN", "API"),
                  composition_var = "state", year_separate = TRUE) |>
  mutate(counterfactual = "Minority")
toc()

save(model_cf_best_state_bs,
     model_cf_white_state_bs,
     model_cf_minority_state_bs,
     file = here("data", "data_constructed", "models_counterfactual.RData"))

rm(data_model_best_race_state)
gc()
```

```{r}
#| label: shut-down-workers

plan(sequential)
```

```{r}
#| label: timestamp-end
#| echo: false

timestamp()
```


