---
title: "Organize Data Notebook"
---

```{r}
#| label: setup
#| include: false
library(here)
source(here("utils","check_packages.R"))
source(here("utils","functions.R"))
```


This script will read in Census and ACS data from 1960 to 2019 to construct broad interracial marriage patterns over that entire time period. Reliable data on Hispanic identification is not available until 1980. IPUMS does provide a Hispanic identifier for earlier years that is primarily based on surnames, but this cannot be used by for interracial marriage because the identifier is the same for spouses. The 1970 data do contain responses to the first Hispanicity question on Form 1, but these responses have been shown to be inconsistent with later responses from 1980 on. We did test this response, and found a substantial decrease in Hispanic intermarriage from 1970 to 1980 that is inconsisten with general trends.

## Read in ACS Data

```{r}
#| label: read-data

census <- read_fwf(here("data", "data_raw", "acs_census", "usa_00145.dat.gz"),
                   col_positions = fwf_cols(YEAR = c(1,4),
                                            SAMPLE = c(5, 10),
                                            HHWT = c(32, 41),
                                            REGION = c(55, 56),
                                            STATEFIP = c(57, 58),
                                            SEX = 86,
                                            AGE = c(87, 89),
                                            MARRNO = 91,
                                            YRMARR = c(92, 95),
                                            AGEMARR = c(96, 97),
                                            RACED = c(99, 101),
                                            HISPAN = 102,
                                            HISP1970 = 114,
                                            BPL = c(106, 108),
                                            EDUCD = c(117, 119),
                                            SEX_SP = 123,
                                            AGE_SP = c(124, 126),
                                            MARRNO_SP = 127,
                                            AGEMARR_SP = c(128, 129),
                                            RACED_SP = c(130, 132),
                                            HISPAN_SP = 133,
                                            HISP1970_SP = 137,
                                            BPL_SP = c(134, 136),
                                            EDUCD_SP = c(140, 142),
                                            QRACE = 121,
                                            QHISPAN = 120,
                                            QRACE_SP = 144,
                                            QHISPAN_SP = 143),
                   col_types = cols(.default = "i", HHWT = "d"))
```

In more recent data we will have some same sex couples. To keep consistency across the time we will focus on heterosexual couples It also makes our data extract simpler because we can grab only men and attach spousal characteristics. However, we will still get male gay couples in the sample, so these need to be removed.

```{r}
#| label: check-spouse-sex

table(census$SEX, census$SEX_SP)

census <- census |>
  filter(SEX_SP == 2)

table(census$SEX, census$SEX_SP)
```

Restrict both spouses to native born status:

```{r}
#| label: restrict-nativity

census |> 
  select(BPL, BPL_SP) |>
  summary()

census <- census |>
  filter(BPL <= 99, BPL_SP <= 99)

census |> 
  select(BPL, BPL_SP) |>
  summary()
```

Lets check the ages of men and women. We won't change age here however.

```{r}
#| label: check-age

summary(census$AGE)
summary(census$AGE_SP)

summary(census$AGE)
summary(census$AGE_SP)
```

Lets merge in data from `tidycensus` to get state names.

```{r}
census <- fips_codes |> 
  as_tibble() |>
  mutate(STATEFIP = as.numeric(state_code)) |>
  select(state, STATEFIP) |>
  distinct() |>
  right_join(census)
```

Ok, now we are ready to code race.

```{r}
#| label: create-vars

# TODO: deal with quality flags on race

# lets create a function for assigning race, since we will do it for both spouses
code_race <- function(raced, hispan, year) {
  
  race <- case_when(
    (hispan != 0 | raced == 110) & year > 1970 ~ "Hispanic",
    raced == 100 ~ "White",
    raced == 200 ~ "Black",
    (raced >= 300 & raced <= 399) |
      raced == 630 | (raced >= 680 & raced <= 699) ~ "Indigenous",
    # TODO: leaving out 650 Other Asian or API in 1980 only
    (raced >= 400 & raced <= 620) | 
      (raced >= 640 & raced <= 643) | 
      (raced >= 660 & raced <= 679) ~ "Asian",
    #raced == 700 ~ "Other", 
    # lets do major biracial groups for fun
    raced == 801 ~ "White-Black",
    raced == 802 | (raced >= 820 & raced <= 825) ~ "White-Indigenous",
    raced >= 810 & raced <= 819 ~ "White-Asian",
    raced == 830 | (raced >= 840 & raced <= 842) ~ "Black-Indigenous",
    raced >= 831 & raced <= 838 ~ "Black-Asian")
  
  factor(race, 
         levels = c("White", "Black", "Indigenous", "Asian", "Hispanic",
                    "White-Black", "White-Indigenous", "White-Asian", 
                    "Black-Indigenous", "Black-Asian"))
}

# also create a function for education
code_educ <- function(educd) {
  education <- case_when(
    educd < 62 ~ "LHS",
    educd < 70 ~ "HS",
    educd < 101 ~ "SC",
    TRUE ~ "C"
  )
  
  education <- factor(education,
                      levels = c("LHS", "HS", "SC", "C"))
  
  return(education)
}

# for the age weighting, we will need five year age groups. Just code
# by the midpoint so we can calculate means
create_age_group <- function(age) {
  
  # rather than categorical, we will use midpoints. For the open ended 
  # categories we will just go 5 above and 2 below.
  age_group <- case_when(
    age < 20 ~ 18,    # less than 20
    age < 25 ~ 22.5,  # 20-24
    age < 30 ~ 27.5,  # 25-29
    age < 35 ~ 32.5,  # 30-34
    age < 45 ~ 40,    # 35-44
    age < 55 ~ 50,    # 45-54
    age >= 55 ~ 57.5, # 55+
    TRUE ~ NA
  )
  
  return(age_group)
  
}

census <- census |>
  mutate(race_husband = code_race(RACED, HISPAN, YEAR),
         race_wife = code_race(RACED_SP, HISPAN_SP, YEAR), 
         educ_husband = code_educ(EDUCD),
         educ_wife = code_educ(EDUCD_SP),
         age_husband = AGE,
         age_wife = AGE_SP,
         age_group_husband = create_age_group(age_husband),
         age_group_wife = create_age_group(age_wife),
         dur_mar = case_when(
           YEAR >= 2008 ~ YEAR-YRMARR,
           YEAR <= 1980 & MARRNO == 1 ~ AGE-AGEMARR,
           YEAR <= 1980 & MARRNO_SP == 1 ~ AGE_SP-AGEMARR_SP,
           TRUE ~ NA
         ),
         region = case_when(
           REGION %in% c(11, 12) ~ "NE",
           REGION %in% c(21, 22) ~ "MW",
           REGION %in% c(31, 32, 33) ~ "S",
           REGION %in% c(41, 42) ~ "W"
         ),
         division = case_when(
           REGION == 11 ~ "New England",
           REGION == 12 ~ "Middle Atlantic",
           REGION == 21 ~ "East North Central",
           REGION == 22 ~ "West North Central",
           REGION == 31 ~ "South Atlantic",
           REGION == 32 ~ "East South Central",
           REGION == 33 ~ "West South Central",
           REGION == 41 ~ "Mountain",
           REGION == 42 ~ "Pacific"
         ),
         year = case_when(
           YEAR <= 2000 ~ YEAR,
           YEAR <= 2003 ~ 2002,
           YEAR <= 2006 ~ 2005,
           YEAR <= 2009 ~ 2008,
           YEAR <= 2012 ~ 2011,
           YEAR <= 2015 ~ 2014,
           YEAR <= 2018 ~ 2017
         )) |>
  # remove 2019
  filter(YEAR != 2019) |>
  # remove other and other multiracial
  filter(!is.na(race_husband) & !is.na(race_wife))
```

Lets check ourselves.

```{r}
#| label: data-coding-checks

table(census$YEAR, census$year)
table(census$EDUCD, census$educ_husband)
table(census$EDUCD_SP, census$educ_wife)
table(census$educ_husband, census$educ_wife)


table(census$race_husband)
table(census$race_wife)

table(census$RACED, census$race_husband)

census |>
  group_by(YEAR) |>
  summarize(mean_dur_mar = mean(dur_mar, na.rm = TRUE),
            min_dur_mar = min(dur_mar, na.rm = TRUE)) |>
  gt()

```

## Calculate age weights

We want to calculate age weights that weight each couple by the joint age distribution to mimic an age distribution similar to recently married couples. To do this we need to first calculate the joint distribution of recent marriages. For now, we will do this based on data for years where we have information on marriage timing. <!-- TODO: Replace 1960-1980 with vital stats tables from 1960-1988 because these only consider first marriages -->

```{r}
#| label: calculate-joint-age-distribution

# long format table of joint distribution of spouse's age by year
prop_age_year_recent <- census |>
  filter(dur_mar <= 1) |>
  group_by(age_group_husband, age_group_wife, YEAR) |>
  summarize(n_dur = n()) |>
  ungroup()

prop_age_year_recent <- census |>
  filter(dur_mar <= 1) |>
  group_by(YEAR) |>
  summarize(n_total = n()) |>
  ungroup() |>
  right_join(prop_age_year_recent) |>
  mutate(prop = n_dur / n_total) |>
  select(age_group_husband, age_group_wife, YEAR, prop)
  
# check sums to 1 in each year
prop_age_year_recent |>
  group_by(YEAR) |>
  summarize(sum(prop)) |>
  gt()
```

```{r}
gs4_deauth()

convert_age_string <- function(x) {
  
 x <- x |> str_remove("^wife_|^husband_")
 x <- ifelse(str_detect(x, "under_18"), "17", x)
 x <- ifelse(str_detect(x, "under_20"), "18", x)
 x <- ifelse(str_detect(x, "65_over"), "65", x)
 x <- ifelse(str_detect(x, "55_over"), "55", x)
 x <- x |>str_split_i("_", 1)
 
 x <- as.numeric(x)
 create_age_group(x)
    
}
```

```{r}
#| label: read-1960-vsus

# this ones a mess - its split up by single, widowed, and divorced wives, but
# the age groups for widowed and divorced start with under 35 and under 25, 
# respectively. I think the way to handle this will be to distribut the values
# in this cell in the same proportions as for the single cells

vsus1960_single <- read_sheet("1YVMMhwrIsyux6FQCPqyCqN45BLOruxLfsk60XBra1z4",
                              sheet = "Table 2-18 Single Bride 1960 VSUS") |>
  filter(wife_age != "wife_total") |>
  select(!husband_total) |>
  pivot_longer(cols = starts_with("husband_"),
               names_to = "husband_age", values_to = "freq")

vsus1960_widowed <- read_sheet("1YVMMhwrIsyux6FQCPqyCqN45BLOruxLfsk60XBra1z4",
                              sheet = "Table 2-18 Widowed Bride 1960 VSUS") |>
  filter(wife_age != "wife_total") |>
  select(!husband_total) |>
  pivot_longer(cols = starts_with("husband_"),
               names_to = "husband_age", values_to = "freq")

vsus1960_divorced <- read_sheet("1YVMMhwrIsyux6FQCPqyCqN45BLOruxLfsk60XBra1z4",
                              sheet = "Table 2-18 Divorced Bride 1960 VSUS") |>
  filter(wife_age != "wife_total") |>
  select(!husband_total) |>
  pivot_longer(cols = starts_with("husband_"),
               names_to = "husband_age", values_to = "freq")

# ok first lets distribute the widowed cases
wife_ages <- c("wife_under_20", "wife_20_24", "wife_25_29", "wife_30_34")
for(husb_age in unique(vsus1960_single$husband_age)) {
  x <- vsus1960_single |> 
    filter(husband_age == husb_age & wife_age %in% wife_ages) |>
    pull(freq)
  p <- x/sum(x)
  freq <- vsus1960_widowed |>
    filter(husband_age == husb_age & wife_age == "wife_under_35") |>
    pull(freq)
  freq <- freq*p
  vsus1960_widowed <- vsus1960_widowed |>
    bind_rows(tibble(wife_age = wife_ages, husband_age = husb_age, freq = freq))
}
# now remove the under_35 cases
vsus1960_widowed <- vsus1960_widowed |>
  filter(wife_age != "wife_under_35")

# now divorced
wife_ages <- c("wife_under_20", "wife_20_24")
for(husb_age in unique(vsus1960_single$husband_age)) {
  x <- vsus1960_single |> 
    filter(husband_age == husb_age & wife_age %in% wife_ages) |>
    pull(freq)
  p <- x/sum(x)
  freq <- vsus1960_divorced |>
    filter(husband_age == husb_age & wife_age == "wife_under_25") |>
    pull(freq)
  freq <- freq*p
  vsus1960_divorced <- vsus1960_divorced |>
    bind_rows(tibble(wife_age = wife_ages, husband_age = husb_age, freq = freq))
}
# now remove the under_35 cases
vsus1960_divorced <- vsus1960_divorced |>
  filter(wife_age != "wife_under_25")

# now combine them together and continue processing
vsus1960 <- vsus1960_single |>
  bind_rows(vsus1960_widowed) |>
  bind_rows(vsus1960_divorced) |>
  mutate(age_group_husband = convert_age_string(husband_age),
         age_group_wife = convert_age_string(wife_age)) |>
  group_by(age_group_husband, age_group_wife) |>
  summarize(freq = sum(freq)) |>
  ungroup() |>
  mutate(prop = freq / sum(freq), YEAR = 1960) |>
  select(age_group_husband, age_group_wife, YEAR, prop)

# remove existing 1960 data and replace with this one
prop_age_year_recent <- prop_age_year_recent |>
  filter(YEAR != 1960) |>
  bind_rows(vsus1960)
```


```{r}
#| label: read-1970-vsus

vsus1970 <- read_sheet("1YVMMhwrIsyux6FQCPqyCqN45BLOruxLfsk60XBra1z4",
                       sheet = "Table 1-28 1970 VSUS") |>
  filter(wife_age != "wife_total") |>
  select(!husband_total) |>
  pivot_longer(cols = starts_with("husband_"),
               names_to = "husband_age", values_to = "freq") |>
  mutate(age_group_husband = convert_age_string(husband_age),
         age_group_wife = convert_age_string(wife_age),
         prop = freq / sum(freq)) |>
  group_by(age_group_husband, age_group_wife) |>
  summarize(prop = sum(prop)) |>
  ungroup() |>
  mutate(YEAR = 1970) |>
  select(age_group_husband, age_group_wife, YEAR, prop)

# remove existing 1970 data and replace with this one
prop_age_year_recent <- prop_age_year_recent |>
  filter(YEAR != 1970) |>
  bind_rows(vsus1970)
```

```{r}
#| label: read-1980-vsus

vsus1980 <- read_sheet("1YVMMhwrIsyux6FQCPqyCqN45BLOruxLfsk60XBra1z4",
                       sheet = "Table 1-28 1980 VSUS") |>
  filter(wife_age != "wife_total") |>
  select(!husband_total) |>
  pivot_longer(cols = starts_with("husband_"),
               names_to = "husband_age", values_to = "freq") |>
  mutate(age_group_husband = convert_age_string(husband_age),
         age_group_wife = convert_age_string(wife_age),
         prop = freq / sum(freq)) |>
  group_by(age_group_husband, age_group_wife) |>
  summarize(prop = sum(prop)) |>
  ungroup() |>
  mutate(YEAR = 1980) |>
  select(age_group_husband, age_group_wife, YEAR, prop)

# remove existing 1980 data and replace with this one
prop_age_year_recent <- prop_age_year_recent |>
  filter(YEAR != 1980) |>
  bind_rows(vsus1980)
```

```{r}
#| label: read-1988-vsus

vsus1988 <- read_sheet("1YVMMhwrIsyux6FQCPqyCqN45BLOruxLfsk60XBra1z4",
                             sheet = "Table 1-20, White Spouses, 1988 VSUS") |>
  filter(wife_age != "wife_total") |>
  select(!husband_total) |>
  pivot_longer(cols = starts_with("husband_"),
               names_to = "husband_age", values_to = "freq")

vsus1988 <- read_sheet("1YVMMhwrIsyux6FQCPqyCqN45BLOruxLfsk60XBra1z4",
                             sheet = "Table 1-20, Black Spouses, 1988 VSUS") |>
  filter(wife_age != "wife_total") |>
  select(!husband_total) |>
  pivot_longer(cols = starts_with("husband_"),
               names_to = "husband_age", values_to = "freq") |>
  bind_rows(vsus1988)

vsus1988 <- read_sheet("1YVMMhwrIsyux6FQCPqyCqN45BLOruxLfsk60XBra1z4",
                             sheet = "Table 1-20, Other Spouses, 1988 VSUS") |>
  filter(wife_age != "wife_total") |>
  select(!husband_total) |>
  pivot_longer(cols = starts_with("husband_"),
               names_to = "husband_age", values_to = "freq") |>
  bind_rows(vsus1988)

vsus1988 <- read_sheet("1YVMMhwrIsyux6FQCPqyCqN45BLOruxLfsk60XBra1z4",
                             sheet = "Table 1-20, Spouses not stated, 1988 VSUS") |>
  filter(wife_age != "wife_total") |>
  select(!husband_total) |>
  pivot_longer(cols = starts_with("husband_"),
               names_to = "husband_age", values_to = "freq") |>
  bind_rows(vsus1988)

vsus1988 <- vsus1988 |>
  mutate(age_group_husband = convert_age_string(husband_age),
         age_group_wife = convert_age_string(wife_age)) |>
  group_by(age_group_husband, age_group_wife) |>
  summarize(freq = sum(freq)) |>
  ungroup() |>
  mutate(prop = freq / sum(freq), 
         YEAR = 1988) |>
  select(age_group_husband, age_group_wife, YEAR, prop)

# add 1988 data to full age prop data
prop_age_year_recent <- prop_age_year_recent |>
  bind_rows(vsus1988)
```


We have no information on duration between 1980 and 2008, so we will need to interpolate the joint distribution. After experimenting with more advanced techniques, we found that a simpler linear interpolation actually provided the best fit. Lets first create a function to interpolate.

```{r}
#| label: function-interpolation

# create function to interpolate
interpolate_age_dist <- function(year_interp, year_start, year_end) {

  # starting and end distributions
  dist_start <- prop_age_year_recent |> 
    filter(YEAR == year_start) |>
    pull(prop)
  
  dist_end <- prop_age_year_recent |> 
    filter(YEAR == year_end) |>
    pull(prop)
  
  # simple linear interpolation outperforms more complex methods
  alpha <- (year_interp - year_start) / (year_end - year_start)  

  # Interpolate linearly between the distributions
  dist_interp <- (1 - alpha) * dist_start + alpha * dist_end
  
  # put into same format as prop_age_year_recent
  prop_age_year_recent |>
    filter(YEAR == year_end) |>
    select(age_group_husband, age_group_wife) |>
    mutate(YEAR = year_interp,
           prop = dist_interp)
}
```

Now we can do the interpolation for the missing years and add them to our full joint distribution data.

```{r}
#| label: age-interpolation

# ok lets now interpolate all needed years by what we have
years_interp <- c(1990, 2000:2007)
prop_age_year_recent <- map(years_interp, interpolate_age_dist, 1988, 2008) |>
  bind_rows() |>
  bind_rows(prop_age_year_recent)
```

And lets check these to make sure they look reasonable with some figures.

```{r}
#| label: fig-summary-age-dist
#| fig-cap: Summary measures of age distribution of spouses, interpolated and non-interpolated data
#| layout-ncol: 1
#| fig-subcap:
#|   - Mean age of husbands
#|   - Mean age of wives
#|   - Spousal Age gap

temp <- prop_age_year_recent |>
  group_by(YEAR) |>
  summarize(mean_age_husband = sum(age_group_husband * prop),
            mean_age_wife = sum(age_group_wife * prop),
            age_diff = mean_age_husband - mean_age_wife) |>
  ungroup() |>
  mutate(interpolated = YEAR %in% years_interp)

ggplot(temp, aes(x = YEAR, y = mean_age_husband))+
  geom_line()+
  geom_point(aes(color = interpolated))+
  labs(x = "year", y = "mean husband's age")+
  theme_bw()

ggplot(temp, aes(x = YEAR, y = mean_age_wife))+
  geom_line()+
  geom_point(aes(color = interpolated))+
  labs(x = "year", y = "mean wife's age")+
  theme_bw()

ggplot(temp, aes(x = YEAR, y = age_diff))+
  geom_line()+
  geom_point(aes(color = interpolated))+
  labs(x = "year", y = "mean husband's age - mean wife's age")+
  theme_bw()
```


```{r}
#| label: fig-age-dist
#| fig-cap: Distribution of spouse's age over time, including real and interpolated cases
#| layout-ncol: 2
#| fig-subcap:
#|   - Husbands
#|   - Wives

prop_age_year_recent |>
  filter(YEAR %in% c(1988, 1990, 2000, 2004, 2008)) |>
  group_by(age_group_husband, YEAR) |>
  summarise(prop = sum(prop)) |>
  mutate(YEAR = factor(YEAR),
         age_group_husband = case_when(
           age_group_husband < 20 ~ "Under 20",
           age_group_husband < 25 ~ "20-24",
           age_group_husband < 30 ~ "25-29",
           age_group_husband < 35 ~ "30-34",
           age_group_husband < 45 ~ "35-44",
           age_group_husband < 55 ~ "45-54",
           age_group_husband >= 55 ~ "55+",
         ),
         age_group_husband = factor(
           age_group_husband,
           levels = c("Under 20", "20-24", "25-29", "30-34", "35-44",
                      "45-54", "55+")
         )) |>
  ggplot(aes(x = age_group_husband, fill = YEAR, group = YEAR, y = prop))+
  geom_col(alpha = 0.5, position = "identity", width = 1)+
  labs(x = "age of husband")+
  theme_bw()

prop_age_year_recent |>
  filter(YEAR %in% c(1988, 1990, 2000, 2004, 2008)) |>
  group_by(age_group_wife, YEAR) |>
  summarise(prop = sum(prop)) |>
  mutate(YEAR = factor(YEAR),
         age_group_wife = case_when(
           age_group_wife < 20 ~ "Under 20",
           age_group_wife < 25 ~ "20-24",
           age_group_wife < 30 ~ "25-29",
           age_group_wife < 35 ~ "30-34",
           age_group_wife < 45 ~ "35-44",
           age_group_wife < 55 ~ "45-54",
           age_group_wife >= 55 ~ "55+",
         ),
         age_group_wife = factor(
           age_group_wife,
           levels = c("Under 20", "20-24", "25-29", "30-34", "35-44",
                      "45-54", "55+")
         )) |>
  ggplot(aes(x = age_group_wife, fill = YEAR, group = YEAR, y = prop))+
  geom_col(alpha = 0.5, position = "identity", width = 1)+
  labs(x = "age of wife")+
  theme_bw()
```

```{r}
#| label: fig-age-dist-tile
#| fig-cap: Heat map of joint distribution of spouse's age, real and interpolated years
prop_age_year_recent |>
  filter(YEAR %in% c(1988, 1990, 2000, 2004, 2008)) |>
  mutate(age_group_husband = case_when(
           age_group_husband < 20 ~ "Under 20",
           age_group_husband < 25 ~ "20-24",
           age_group_husband < 30 ~ "25-29",
           age_group_husband < 35 ~ "30-34",
           age_group_husband < 45 ~ "35-44",
           age_group_husband < 55 ~ "45-54",
           age_group_husband >= 55 ~ "55+",
         ),
         age_group_husband = factor(
           age_group_husband,
           levels = c("Under 20", "20-24", "25-29", "30-34", "35-44",
                      "45-54", "55+")
         ),
         age_group_wife = case_when(
           age_group_wife < 20 ~ "Under 20",
           age_group_wife < 25 ~ "20-24",
           age_group_wife < 30 ~ "25-29",
           age_group_wife < 35 ~ "30-34",
           age_group_wife < 45 ~ "35-44",
           age_group_wife < 55 ~ "45-54",
           age_group_wife >= 55 ~ "55+",
         ),
         age_group_wife = factor(
           age_group_wife,
           levels = c("Under 20", "20-24", "25-29", "30-34", "35-44",
                      "45-54", "55+")
         )) |>
  ggplot(aes(x = age_group_husband, y = age_group_wife, fill = prop)) +
  geom_tile() +
  scale_fill_viridis_c(direction = -1) +
  facet_wrap(~YEAR)+
  labs(x = "husband's age", y = "wife's age", fill = "joint p")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

Now that we have the joint distribution of age for recent marriages, we can compute the same proportions across all extant marriages and then use the ratio of the proportions to get the weights, the same as for post-stratification weights.


```{r}
#| label: calculate-weights

prop_age_year <- census |>
  group_by(age_group_husband, age_group_wife, YEAR) |>
  summarize(n_dur = n()) |>
  ungroup()

prop_age_year <- census |>
  group_by(YEAR) |>
  summarize(n_total = n()) |>
  ungroup() |>
  right_join(prop_age_year) |>
  mutate(prop_total = n_dur / n_total) |>
  select(age_group_husband, age_group_wife, YEAR, prop_total)

prop_age_year |>
  group_by(YEAR) |>
  summarize(sum = sum(prop_total)) |>
  gt()

weighting <- full_join(prop_age_year, prop_age_year_recent) |>
  mutate(weight_age = prop / prop_total,
         weight_age = ifelse(is.na(weight_age), 0, weight_age))

census <- census |>
  left_join(weighting)

# confirm that the mean weight in each year is 1
census |>
  group_by(YEAR) |>
  summarize(mean = mean(weight_age)) |>
  gt()
```

As a test case, lets see how the weights adjust the distribution in 2008.

```{r}
#| label: fig-age-weights-hist
#| fig-cap: The distribution of husband's age in 2008. The dotted line gives the distribution of husband's age for recent marriages. The red histogram gives the distribution of husband's age across all observations, while the green histogram applies age weights to the full distribution to bring them in line with recent marriages.

census |>
  filter(year == "2008") |>
  ggplot(aes(x = age_husband, y = after_stat(density)))+
  geom_histogram(alpha = 0.7, fill = "red", binwidth = 1)+
  geom_histogram(alpha = 0.7, fill = "navy", binwidth = 1, 
                 aes(weight = weight_age))+
  geom_histogram(data = filter(census, year == "2008" & dur_mar <= 1),
               fill = NA, color = "black", linetype = 2, 
               binwidth = 1)+
  theme_bw()
```

## Save the data

Now lets save the data.

```{r}
census <- census |> 
  select(race_husband, race_wife, 
         educ_husband, educ_wife, 
         age_husband, age_wife, 
         dur_mar, weight_age, 
         state, division, region, year)
         

save(census, file = here("data", "data_constructed", "individ_data.RData"))
```




