---
title: "Organize Data Notebook"
---

```{r}
#| label: setup
#| include: false
library(here)
source(here("utils","check_packages.R"))
source(here("utils","functions.R"))
```


This script will read in Census and ACS data from 1960 to 2019 to construct broad interracial marriage patterns over that entire time period. Reliable data on Hispanic identification is not available until 1980. IPUMS does provide a Hispanic identifier for earlier years that is primarily based on surnames, but this cannot be used by for interracial marriage because the identifier is the same for spouses. The 1970 data do contain responses to the first Hispanicity question on Form 1, but these responses have been shown to be inconsistent with later responses from 1980 on. We did test this response, and found a substantial decrease in Hispanic intermarriage from 1970 to 1980 that is inconsisten with general trends.

```{r}
#| label: read-data

census <- read_fwf(here("data", "data_raw", "acs_census", "usa_00145.dat.gz"),
                   col_positions = fwf_cols(YEAR = c(1,4),
                                            SAMPLE = c(5, 10),
                                            HHWT = c(32, 41),
                                            REGION = c(55, 56),
                                            STATEFIP = c(57, 58),
                                            SEX = 86,
                                            AGE = c(87, 89),
                                            MARRNO = 91,
                                            YRMARR = c(92, 95),
                                            AGEMARR = c(96, 97),
                                            RACED = c(99, 101),
                                            HISPAN = 102,
                                            HISP1970 = 114,
                                            BPL = c(106, 108),
                                            EDUCD = c(117, 119),
                                            SEX_SP = 123,
                                            AGE_SP = c(124, 126),
                                            MARRNO_SP = 127,
                                            AGEMARR_SP = c(128, 129),
                                            RACED_SP = c(130, 132),
                                            HISPAN_SP = 133,
                                            HISP1970_SP = 137,
                                            BPL_SP = c(134, 136),
                                            EDUCD_SP = c(140, 142),
                                            QRACE = 121,
                                            QHISPAN = 120,
                                            QRACE_SP = 144,
                                            QHISPAN_SP = 143),
                   col_types = cols(.default = "i", HHWT = "d"))
```

In more recent data we will have some same sex couples. To keep consistency across the time we will focus on heterosexual couples It also makes our data extract simpler because we can grab only men and attach spousal characteristics. However, we will still get male gay couples in the sample, so these need to be removed.

```{r}
#| label: check-spouse-sex

table(census$SEX, census$SEX_SP)

census <- census |>
  filter(SEX_SP == 2)

table(census$SEX, census$SEX_SP)
```

Replace HISPAN with HISP1970 for the 1970 data (actuall don't do this).

```{r}
#| label: restrict-1970
#| eval: false

census <- census |>
  mutate(HISPAN = ifelse(YEAR == 1970, HISP1970, HISPAN),
         HISPAN_SP = ifelse(YEAR == 1970, HISP1970_SP, HISPAN_SP))
```

Restrict both spouses to native born status

```{r}
#| label: restrict-nativity

census |> 
  select(BPL, BPL_SP) |>
  summary()

census <- census |>
  filter(BPL <= 99, BPL_SP <= 99)

census |> 
  select(BPL, BPL_SP) |>
  summary()
```

Lets check the ages of men and women. We won't chane age here however.

```{r}
#| label: check-age

summary(census$AGE)
summary(census$AGE_SP)

summary(census$AGE)
summary(census$AGE_SP)

# lets restrict this to min 15 just like men
census <- census |>
  filter(AGE_SP >= 15)
```

Lets merge in data from `tidycensus` to get state names.

```{r}
census <- fips_codes |> 
  as_tibble() |>
  mutate(STATEFIP = as.numeric(state_code)) |>
  select(state, STATEFIP) |>
  distinct() |>
  right_join(census)
```
Hmmm, a bit troubling on the minimum values, but we will leave them be for this analysis.

Ok, now we are ready to code race.

```{r}
#| label: create-vars

# TODO: deal with quality flags on race

# lets create a function for assigning race, since we will do it for both spouses
code_race <- function(raced, hispan, year) {
  
  race <- case_when(
    (hispan != 0 | raced == 110) & year > 1970 ~ "Hispanic",
    raced == 100 ~ "White",
    raced == 200 ~ "Black",
    raced >= 300 & raced <= 399 ~ "AIAN",
    # we have to do API because 1980 and 1990
    (raced >= 400 & raced <= 699) | (raced >= 860 & raced <= 869) ~ "API",
    #raced == 700 ~ "Other", 
    # lets do major biracial groups for fun
    raced == 801 ~ "White-Black",
    raced == 802 ~ "White-AIAN",
    raced >= 803 & raced <= 825 ~ "White-API",
    raced == 830 ~ "Black-AIAN",
    raced >= 831 & raced <= 842 ~ "Black-API")
  
  factor(race, 
         levels = c("White", "Black", "AIAN", "Hispanic", "API",
                    "White-Black", "White-AIAN", "White-API", 
                    "Black-AIAN", "Black-API"))
}

code_educ <- function(educd) {
  education <- case_when(
    educd < 62 ~ "LHS",
    educd < 70 ~ "HS",
    educd < 101 ~ "SC",
    TRUE ~ "C"
  )
  
  education <- factor(education,
                      levels = c("LHS", "HS", "SC", "C"))
  
  return(education)
}

census <- census |>
  mutate(race_husband = code_race(RACED, HISPAN, YEAR),
         race_wife = code_race(RACED_SP, HISPAN_SP, YEAR), 
         educ_husband = code_educ(EDUCD),
         educ_wife = code_educ(EDUCD_SP),
         age_husband = AGE,
         age_wife = AGE_SP,
         dur_mar = case_when(
           YEAR >= 2008 ~ YEAR-YRMARR,
           YEAR <= 1980 & MARRNO == 1 ~ AGE-AGEMARR,
           YEAR <= 1980 & MARRNO_SP == 1 ~ AGE_SP-AGEMARR_SP,
           TRUE ~ NA
         ),
         region = case_when(
           REGION %in% c(11, 12) ~ "NE",
           REGION %in% c(21, 22) ~ "MW",
           REGION %in% c(31, 32, 33) ~ "S",
           REGION %in% c(41, 42) ~ "W"
         ),
         division = case_when(
           REGION == 11 ~ "New England",
           REGION == 12 ~ "Middle Atlantic",
           REGION == 21 ~ "East North Central",
           REGION == 22 ~ "West North Central",
           REGION == 31 ~ "South Atlantic",
           REGION == 32 ~ "East South Central",
           REGION == 33 ~ "West South Central",
           REGION == 41 ~ "Mountain",
           REGION == 42 ~ "Pacific"
         ),
         year = case_when(
           YEAR <= 2000 ~ YEAR,
           YEAR <= 2003 ~ 2002,
           YEAR <= 2006 ~ 2005,
           YEAR <= 2009 ~ 2008,
           YEAR <= 2012 ~ 2011,
           YEAR <= 2015 ~ 2014,
           YEAR <= 2018 ~ 2017
         )) |>
  # remove 2019
  filter(YEAR != 2019) |>
  # remove other and other multiracial
  filter(!is.na(race_husband) & !is.na(race_wife))

table(census$YEAR, census$year)
table(census$EDUCD, census$educ_husband)
table(census$EDUCD_SP, census$educ_wife)
table(census$educ_husband, census$educ_wife)


table(census$race_husband)
table(census$race_wife)

census |>
  group_by(YEAR) |>
  summarize(mean_dur_mar = mean(dur_mar, na.rm = TRUE),
            min_dur_mar = min(dur_mar, na.rm = TRUE))

```

Ok, now lets create age weights based on recent marriages.

```{r}
counts_age_year <- census |>
  group_by(age_husband, age_wife, YEAR) |>
  summarize(n = n())

counts_year <- census |>
  group_by(YEAR) |>
  summarize(total = n())

prop_age_year <- counts_age_year |>
  left_join(counts_year) |>
  mutate(prop_total = n / total) |>
  select(-n, -total)

counts_age_year_dur <- census |>
  filter(dur_mar <= 1 & (YEAR > 1980 | (MARRNO == 1 & MARRNO_SP == 1))) |>
  group_by(age_husband, age_wife, YEAR) |>
  summarize(n = n())

counts_year_dur <- census |>
  filter(dur_mar <= 1 & (YEAR > 1980 | (MARRNO == 1 & MARRNO_SP == 1))) |>
  group_by(YEAR) |>
  summarize(total = n())

prop_age_year_dur <- counts_age_year_dur |>
  left_join(counts_year_dur) |>
  mutate(prop_dur = n / total) |>
  select(-n, -total)

# some checks
prop_age_year |>
  group_by(YEAR) |>
  summarize(sum = sum(prop_total))

prop_age_year_dur |>
  group_by(YEAR) |>
  summarize(sum = sum(prop_dur))

weighting <- left_join(prop_age_year, prop_age_year_dur) |>
  mutate(weight_age = prop_dur / prop_total,
         weight_age = ifelse(is.na(weight_age), 0, weight_age))

# need to center weight_age mean on 1 in each year
weighting <- weighting |>
  group_by(YEAR) |>
  summarize(mean = mean(weight_age)) |>
  right_join(weighting) |>
  mutate(weight_age = weight_age / mean) |>
  select(age_husband, age_wife, YEAR, weight_age)

weighting |>
  group_by(YEAR) |>
  summarize(mean = mean(weight_age))

census <- census |>
  left_join(weighting)


census |>
  filter(year == "1960") |>
  ggplot(aes(x = age_husband, y = after_stat(density)))+
  geom_histogram(alpha = 0.7, fill = "red", binwidth = 1)+
  geom_histogram(data = filter(census, year == "1960" & 
                                 dur_mar <= 1 & MARRNO == 1 & MARRNO_SP == 1),
                 alpha = 0.7, fill = "navy", binwidth = 1)+
  theme_bw()

census |>
  filter(year == "2008") |>
  ggplot(aes(x = age_husband, y = after_stat(density)))+
  geom_histogram(alpha = 0.7, fill = "red", binwidth = 1)+
  geom_histogram(alpha = 0.7, fill = "darkgreen", binwidth = 1, 
                 aes(weight = weight_age))+
  geom_histogram(data = filter(census, year == "2008" & dur_mar <= 1),
               fill = NA, color = "navy", linetype = 2, binwidth = 1)+
  theme_bw()
```

Now lets save the data.

```{r}
census <- census |> 
  select(race_husband, race_wife, 
         educ_husband, educ_wife, 
         age_husband, age_wife, 
         dur_mar, weight_age, 
         state, division, region, year)
         

save(census, file = here("data", "data_constructed", "individ_data.RData"))
```




