---
title: "Run Models"
execute:
  warning: false
  message: false
---

```{r}
#| label: timestamp-start

timestamp()
```

```{r}
#| label: setup
#| include: false
library(here)
source(here("utils","check_packages.R"))
source(here("utils","functions.R"))
```


```{r}
#| label: prepare-data
#| include: false

load(here("data", "data_constructed", "individ_data.RData"))

# relevel year and factor up anything that might go into models
census <- census |>
  mutate(year = factor(year),
         year = relevel(year, "2000"),
         region = factor(region),
         division = factor(division),
         state = factor(state))
```

```{r}
#| label: set-globals

# number of bootstrap samples
B <- 50
```


Lets compare processing time of bootstrapped vs non-bootstrapped model for single race groups

```{r}

tic()
marg_single <- census |>
  group_by(race_husband, race_wife, year, .drop = FALSE) |>
  summarize(freq = sum(weight_age), .groups = "drop") |>
  estimate_lor(c("White", "Black", "AIAN", "API", "Hispanic"))
toc()

tic()
marg_single_bs <- census |>
  bootstrap_model(c("White", "Black", "AIAN", "API", "Hispanic")) |>
  mutate(type = "bootstrap")
toc()

marg_single |>
  mutate(type = "glm") |>
  bind_rows(marg_single_bs) |>
  ggplot(aes(x = year, y = estimate, color = type))+
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill = type),
              alpha = 0.3, color = NA)+
  geom_point(alpha = 0.5)+
  geom_line(alpha = 0.5, linewidth = 1.25)+
  facet_wrap(~term, scales = "free_y")+
  scale_color_manual(values = c("navy", "red"))+
  scale_fill_manual(values = c("navy","red"))+
  theme_bw()
```

How long does it take to run state based model for one pairing?

```{r}
tic()
marg_wb_state_bs <- census |>
  bootstrap_model(c("White", "Black"), composition_var = "state",
                  show_progress = TRUE) |>
  mutate(type = "bootstrap")
toc()

```

About 6.5 minutes with $B = 10$, so for all ten single race pairings, this should take about 65 minutes. Each additional 10 B is going to add an hour basically. 

How long does a division model take?

```{r}
tic()
marg_wb_division_bs <- census |>
  bootstrap_model(c("White", "Black"), composition_var = "division",
                  show_progress = TRUE) |>
  mutate(type = "bootstrap")
toc()

```

So much faster. 

```{r}
marg_wb_state_bs <- marg_wb_state_bs |>
  mutate(composition = "state")

marg_wb_bs <- marg_single_bs |>
  filter(term == "White/Black") |>
  mutate(composition = "none")

marg_wb_division_bs |>
  mutate(composition = "division") |>
  bind_rows(marg_wb_state_bs) |>
  bind_rows(marg_wb_bs) |>
  mutate(factor(composition, levels = c("none", "division", "state"))) |>
  ggplot(aes(x = year, y = estimate, color = composition))+
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill = composition),
              alpha = 0.3, color = NA)+
  geom_point(alpha = 0.5)+
  geom_line(alpha = 0.5, linewidth = 1.25)+
  scale_color_manual(values = c("gold", "navy", "red"))+
  scale_fill_manual(values = c("gold", "navy","red"))+
  theme_bw()
```

```{r}
tic()
marg_single_by_region_bs <- census |>
  bootstrap_model(c("White", "Black", "AIAN", "API", "Hispanic"), 
                  conditional_var = "region",
                  show_progress = TRUE) |>
  mutate(type = "bootstrap")
toc()

tic()
marg_single_by_region <- census |>
  group_by(race_husband, race_wife, region, year, .drop = FALSE) |>
  summarize(freq = sum(weight_age), .groups = "drop") |>
  estimate_lor(c("White", "Black", "AIAN", "API", "Hispanic"),
               conditional_var = "region")
toc()

```

```{r}
marg_single_by_region_bs |>
  ggplot(aes(x = year, y = estimate, color = region))+
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill = region),
              alpha = 0.3, color = NA)+
  geom_point(alpha = 0.5)+
  geom_line(alpha = 0.5, linewidth = 1.25)+
  facet_wrap(~term)+
  scale_color_manual(values = c("darkgreen", "gold", "navy", "red"))+
  scale_fill_manual(values = c("darkgreen", "gold", "navy","red"))+
  theme_bw()

```

```

```{r}
#| label: timestamp-end

timestamp()
```