---
title: "Heterogeneity in the Trends in Interracial Marriage in the US, 1960-2023"
shorttitle: "Trends in Interracial Marriage, 1960-2023"
abstract: An Abstract
keywords: [keyword1, keyword2]
thanks: Thanks to everyone for checking this out.
reference-section-title: References
bibliography: ../bibliography/project.bib
csl: ../bibliography/chicago-parenthetical.csl
format:
  pdf:
    keep-tex: false
    fig-pos: "h"
    include-in-header: 
      text: |
        \usepackage{dcolumn}
#  submittable-pdf:
#     keep-tex: false
#     fig-pos: "!t"
#     include-in-header: 
#       text: |
#         \usepackage{dcolumn}
#  submittable-docx: 
#    fig-format: pdf
---

```{r}
#| label: setup
#| include: false
library(here)
source(here("utils","check_packages.R"))
source(here("utils","functions.R"))
```

```{r}
#| label: load-data

load(here("data", "data_constructed", "exogamy_trends.RData"))
load(here("data", "data_constructed", "model_state_bs.RData"))
load(here("data", "data_constructed", "model_basic_bs.RData"))
load(here("data", "data_constructed", "models_subgroups.RData"))
load(here("data", "data_constructed", "models_counterfactual.RData"))
load(here("data", "data_constructed", "model_region.RData"))
load(here("data", "data_constructed", "models_state_gen_bs.RData"))
load(here("data", "data_constructed", "model_recent_mar.RData"))
load(here("data", "data_constructed", "models_cohab.RData"))
```

```{r}
#| label: set-color-palette

palette <- c(
  # Single race combinations
  "White/Black"                   = "#E69F00",  # rich orange
  "White/AIAN"                    = "#009E73",  # teal green
  "White/API"                     = "#CC79A7",  # muted magenta
  "White/Hispanic"                = "#661100",  # deep rust
  "Black/AIAN"                    = "#A65628",  # warm brown
  "Black/API"                     = "#74AA99",  # muted cyan
  "Black/Hispanic"                = "#882255",  # deep burgundy
  "AIAN/API"                      = "#56B4E9",  # desaturated turquoise
  "AIAN/Hispanic"                 = "#AA4499",  # dusty rose
  "API/Hispanic"                  = "#117733",  # forest green
  
  # Single race to multiracial combinations
  "White/White-Black"             = "#47AC11",  # vivid spring green
  "Black/White-Black"             = "#471624",  # deep maroon-brown
  "White/White-AIAN"              = "#361892",  # royal indigo
  "AIAN/White-AIAN"               = "#23A4BB",  # bright teal-blue
  "White/White-API"               = "#667AA3",  # slate blue
  "API/White-API"                 = "#6D7003",  # olive drab
  "Black/Black-API"               = "#1117AA",  # bold cobalt blue
  "API/Black-API"                 = "#6D4113",  # dark umber
  "Black/Black-AIAN"              = "#7171BB",  # muted periwinkle
  "AIAN/Black-AIAN"               = "#422D13",  # earthy espresso

  # Hispanic subgroups with White and Black
  "White/WhiteHispanic"           = "#9E9E00",  # olive yellow
  "White/BlackHispanic"           = "#AA0055",  # rose plum
  "White/OtherHispanic"           = "#4FAA37",  # leafy green
  "Black/WhiteHispanic"           = "#552288",  # soft purple
  "Black/BlackHispanic"           = "#AA3344",  # reddish brick
  "Black/OtherHispanic"           = "#2C37DA",  # vivid sapphire

  # Asian subgroups with White
  "White/EastAsian"               = "#3366AA",  # steely blue
  "White/E&SE Asian"              = "#3366AA",  # steely blue
  "White/SEAsian"                 = "#118877",  # sea green
  "White/South Asian"              = "#9933AA",  # muted violet-magenta

  # Indigenous subgroups with White
  "White/AIAN"                    = "#CC4477",  # muted raspberry rose
  "White/Pac. Islander"           = "#3311EE",  # steel sky blue
  
  # Intra-Hispanic
  "WhiteHispanic/BlackHispanic"   = "#BB8866",  # muted tan
  "WhiteHispanic/OtherHispanic"   = "#885577",  # dusty mauve
  "BlackHispanic/OtherHispanic"   = "#AA8844",  # muted amber
  
  # Intra-Asian
  "EastAsian/SEAsian"             = "#336633",  # forest moss
  "EastAsian/SouthAsian"          = "#993399",  # grape violet
  "SEAsian/SouthAsian"            = "#CC6666",  # muted rose

  # Intra-Indigenous
  "AIAN/PI"                       = "#447777"   # muted teal gray
)

```

# Introduction

# Background

# Data and Methods

Data for this analysis are derived from the microdata Census samples for Census 1960, 1970, 1980, 1990, and 2000, and the American Community Survey (ACS) data from 2001 to 2023, with the exception of 2020. All data were extracted via the Integrated Public Use Microdata Sample system <!-- TODO: cite source -->. ACS data from 2020 is removed due to issues with data collection during the COVID pandemic.

The data include all extant heterosexual married couples in each sample. I also include a secondary sample of heterosexual cohabiting couples, based on data from 1990 at which point "unmarried partner" is provided as a choice for relationship to head of household. <!-- TODO: justify heterosexual -->

For most of the analysis, I also restrict the sample to couples in which both partners are native born. Foreign-born individuals may have been married prior to arrival in the US. Even in cases, where foreign-born individuals arrived in the US unmarried, they will have substantial preferences for ethnonational endogamy that could easily be conflated with ethnoracial preferences. This may be less true of individuals who arrived as children (i.e. the 1.5 generation). In supplementary analysis discussed later, I consider estimates of interracial marriage that are less restrictive on national origin.

## Measuring Race

In order to measure the frequency of marriage across groups, I first have to establish what those groups will be. Ideally, group distinctions would capture meaningful and salient ideas within a given society of where group boundaries lie. However, over a period of this amount of time, this desire immediately faces both conceptual and methodological issues. Conceptually, the meaning, placement, and salience of boundaries may change over time. Methodologically, the Census's approach to measuring race and ethnicity over this time period has changed. For these reasons, I will confine most of the analysis to the large single race groups of White, Black, American Indian/Alaska Native (AIAN), Asian/Pacific Islander (API), and Hispanic, where conceptual and methodological changes are less likely to cause discrepancy over time.

Nonetheless, several changes in Census coding of race are noteworthy. First, Hispanic identification is not determined by a category on the race question but by a separate question on Hipanicity that precedes the race question. This question was not asked on all Census forms until 1980.[^1] Thus, I cannot measure Hispanic intermarriage patterns prior to 1980 and Hispanics in the census in 1960 and 1970 will instead be distributed into other racial categories. From 1980 onward, individuals who identified as Hispanic on the Hispanicity question are identified as Hispanic in my analytical data regardless of their race question.

[^1]: Form 1 of the 1970 Census also included a Hispanicity question, but the results from this Hispanicity question have been found to be not comparable to the Hispanicity question asked in 1980 onwards. <!-- TODO: find a cite here -->

Second, the collapsing of Asian and Pacific Islander respondents is less than ideal because the experiences of most Pacific Islander populations are distinct from those of Asian populations, and come closer to the experiences of other indigenous populations<!-- TODO: cite? -->. However, the 1980 Census includes a category of "Other Asian and Pacific Islander" that makes it impossible to fully distinguish these two groups in the data. Furthermore, the "best race" data that I discuss more fully below groups Asians and Pacific Islanders together. Thus, for the majority of the analysis I use the combined Asian/Pacific Islander category. However, for some supplementary analysis, I split these two groups up for a more fine grained analysis. I handle the "Other Asian and Pacific Islander" category in 1980 by treating it as a missing data problem. Individuals in these categories are assigned either an Asian or Pacific Islander identity via predictive mean matching.

Third, starting in 2000, individuals are allowed to check all categories that apply in response to the race question. This creates a major discrepancy before and after 2000, as individuals who were previously limited to one race response can now check multiple options. This change may create changes in patterns of interracial marriage that are not easy to predict apriori, because we don't know who is mostly likely to change their race response. I handle this shift in a couple of ways. F Second, I consider different counterfactual scenarios of race reporting if multiracial individuals were instead forced back into a single race option.

Finally, from 2020 onward, based on a format change to the race question on the Census and ACS, the Census Bureau changed its method of allocation of race responses in such a way that the number of multiracial respondents increased explosively <!-- TODO: citation -->. This increase was most notable for individuals who identified as Hispanic and were placed in a multiracial category that included the "Other" category <!-- TODO: citation -->. Because, I am identifying Hispanics by the Hispanicity question, this measurement issue should not lead to large discrepancies in the results here.

Even apart from these specific changes to measurement, it is possible that some patterns of change in interracial marriage may be driven by changes in patterns of racial identification over time. Race is not fixed and longitudinal data has shown that people do change their racial identification over time <!-- TODO: cite -->. Some of this change may be a kind of "churning" with no particular direction or trend over time, in which case, we can think of the racial categories here as measuring a probabilistic assessment of one's race. Other changes may reflect long term secular change which could affect patterns of intermarriage. The clearest case of this kind of change over the time period is the renewal and growth of American Indian identification from the mid-Twentieth century to at least the end of that century<!-- TODO: cite -->. Such growth may have contributed to increases in interracial marriage when formerly non-American Indian spouses switched their identification to American Indian. However, it is also possible that non-American Indians with American Indian spouses were more likely to switch which would then reduce the measure of interracial marriage.

For most of the analysis, I focus on interracial marriage between the five major racial groups of White, Black, American Indian/Alaska Native, Asian/Pacific Islander, and Hispanic, giving me a total of 10 pairings to consider. However, in some parts of the analysis, I disaggregate or consider other groups to explore trends in more depths. First, from 2000 on, I look at the tendency of the largest multiracial groups (White-Black, White-API, and White-AIAN) to intermarry with their two constituent groups (e.g. a White-Black person marrying a White or Black person).

Second, for some parts of the analysis I disaggregate the Hispanic population into three gorups based on three most common single race responses of White, Other, and Black. <!-- TODO: say more about the dangers of reading too much into these categories --> Some scholars have treated these measures as synonymous with skin tone <!-- TODO: cite -->, but I hesitate to do that here as race reporting among Hispanics is complex and multi-faceted <!-- TODO: cite -->. Nonetheless, these racial distinctions partly tap into some potential divisions among the Hispanic population that are worth exploring, particular in relation to marriage with non-Hispanic White and Black partners.

Third, I disaggregate the Asian/Pacific Islander population into the sub-populations of East/Southeast Asian, South Asian, and Pacific Islander. Prior work has shown significant marriage boundaries between South and East/Southeast Asian groups.

## Measuring Interracial Marriage

To help fix ideas, @tbl-cross-product provides an abstract view of the frequency of intergroup marriage for three groups when marriages are cross-tabulated by the race of husband and wife.

```{r}
#| label: tbl-cross-product
#| tbl-cap: Abstract table of the frequency of interracial marriage across three groups, cross-tabulated by husband's race and wife's race.

tibble(
  "hrace" = c("Group 1", "Group 2", "Group 3"),
  "g1" = c("$F_{11}$", "$F_{21}$", "$F_{31}$"),
  "g2" = c("$F_{12}$", "$F_{22}$", "$F_{32}$"),
  "g3" = c("$F_{11}$", "$F_{21}$", "$F_{31}$"),
)  |> 
  gt() |>
  fmt_markdown() |>
  cols_align("center", columns = starts_with("g")) |>
  cols_label(hrace = "Husband",
             g1 = "Group 1", g2 = "Group 2", g3 = "Group 3") |>
  tab_spanner(label = "Wife", columns = starts_with("g"))
```

One common approach to measure interracial marriage "rates" is to calculate the outmarriage ratio, or proportion. This outmarriage ratio is simply the proportion of individuals of a certain group (or the entire population) that are married to someone of a different race. In @tbl-cross-product, the outmarriage ratio for Group 1 is given by:

$$
\text{outmarriage ratio} = \frac{F_{12}+F_{13}+F_{21}+F_{31}}{F_{11}+F_{12}+F_{13}+F_{21}+F_{31}}
$$ {#eq-outmarriage-ratio}

This approach provides a reasonable measure of the experience of interracial marriage, for either the population at large, or for specific groups. However, it has other significant limitations. First, it makes no adjustment for group size in determining the likelihood of an interracial marriage. Groups that make up a larger share of the overall population will, even in the absence of racial preferences for spouses, have lower outmarriage ratios because they have more in-group choices for marriage. Thus, meaningful inferences about social boundaries cannot be drawn by comparing outmarriage ratios across groups. Second, the outmarriage ratio is simply the complement to a measure of endogamy for a single group and therefore does not provide a pairwise measure of the tendency to intermarry between two groups. While outmarriage ratios could be extended into a full conditional distribution of the race of other partner by the race of focal partner, this approach will exacerbate the first problem of group size. Most notably, the outmarriage ratio of Group 1 to Group 2 will not be the same as the outmarriage ratio of Group 2 to Group 1, unless those groups are equal in size. This finding is not a substantive finding of difference, but rather a mechanical limitation.

For these reasons, the preferred measure of the strength of intermarriage across group combinations has long been the odds ratio or cross-product resulting from a cross-tabulation of husband's race by wife's race. From @tbl-cross-product, I can estimate the odds ratio of intergroup marriage between Group 1 and Group 2 as:

$$
\text{odds ratio} = \frac{F_{12}F_{21}}{F_{11}F_{22}}
$$ {#eq-cross-prod}

This cross-product can be interpreted as the odds ratio of an individual from one group having a wife from the other group relative to those same odds for an individual from the other group. When this cross-product equals one, there are no barriers to intergroup marriage. Values lower than one indicate barriers to intergroup marriage. Unlike the outmarriage ratio, the cross-product adjusts for group size and thuse the odds ratios across different groups are directly comparable.

Although this odds ratio and its standard error can be estimated directly from a cross-tabulation shown in @tbl-cross-product, I will prefer a model approach to estimate these values for the analysis. By using a model, I can condition the results on a variety of other factors, discussed below. To estimate the odds ratios, I use a log-linear model approach, where the outcome variable are the frequency counts estimated by a cross-tabulation like @tbl-cross-product. For a single year of the analysis, I can estimate the following model:

$$
\log(F_{ij})=\lambda+\lambda_i+\lambda_j+\beta_{ij}+\alpha_{ij}
$$ {#eq-model-basic}

Where $F_{ij}$ is the frequency count of marriages with a husband of race $i$ and a wife of race $j$. The model accounts for the marginal distributions of husband's race ($\lambda_i$) and wife's race ($\lambda_j$). The $\beta_{ij}$ and $\alpha_{ij}$ parameters account for the saturated interaction between husband's and wife's race. The $\beta_{ij}$ parameters estimate the desired log odds ratios for each non-endogamous pairing of husband's and wife's race. The $\alpha_{ij}$ parameters are residual parameters that must also be estimated to fully saturate the interaction between husband's and wife's race. For example, in a three-by-three table like @tbl-cross-product, one can identify four non-redundant odds ratios, but only three intermarriage odds ratios ($\beta_{ij}$) can be estimated for the three pairwise group comparisons. The fourth unique odds ratio is estimated with $\alpha_{ij}$.

To clarify this approach, @tbl-model-param, shows the parameter coding that would be used for a model in the three group case shown in @tbl-cross-product. The three $\beta$ parameters are estimated by dummy coding the upper off-diagonal cell for each pairwise comparison. The single $\alpha$ parameter is estimated via ANOVA style coding for one of the pairings. Using this modeling approach, the $\beta$ parameters of the log-linear model will exactly correspond to the odds ratios estimated manually by the cross-product.

```{r}
#| label: tbl-model-param
#| tbl-cap: Log-linear model parameters applied to a cross-tabulation of husband's race and wife's race for three groups.

tibble(
  "hrace" = c("Group 1", "Group 2", "Group 3"),
  "g1" = c("$0$", "$0$", "$0$"),
  "g2" = c("$\\beta_{12}$", "$0$", "$-\\alpha_{23}$"),
  "g3" = c("$\\beta_{13}$", "$\\beta_{23}+\\alpha_{23}$", "$0$"),
)  |> 
  gt() |>
  fmt_markdown() |>
  cols_align("center", columns = starts_with("g")) |>
  cols_label(hrace = "Husband",
             g1 = "Group 1", g2 = "Group 2", g3 = "Group 3") |>
  tab_spanner(label = "Wife", columns = starts_with("g"))
```

In most of the models used in the analysis to follow, I make two additional adjustments to the model outlined in @eq-model-basic. First, I use a single model across time and allow all of the parameters from @eq-model-basic to vary by time. Second, I include state of residence to adjust for subnational variation in population distribution by race. CITE have shown that such subnational variation importantly affects the strength of intermarriage. Estimates based on national data will downwardly bias estimates of interracial marriage because they do not fully account for the population share of different racial groups in a spouses marriage market.

The full model is thus given by:

$$
\log(F_{ijkt}) = \lambda+\lambda_i+\lambda_j+\lambda_k+\lambda_t+\lambda_{ik}+\lambda_{jk}+\lambda_{it}+\lambda_{jt}+\lambda_{ikt}+\lambda_{jkt}+\beta_{ijt}+\alpha_{ijt}
$$ {#eq-model-full}

where $j$ is the state of residence and $t$ is the time period. All the two and three way tables between each partner's race, time, and state are added to the model and the $\beta$ and $\alpha$ parameters themselves vary by time. In this model setup, I do not allow the $\beta$ parameters to vary by state, as such models would be very noisy and difficult to estimate for many comparisons. However, I do allow the $\beta$ parameters to vary by region in some of the analyses that follow.

## Measuring Stocks vs. Flows

Ideally, I would use data on the "flow" of recent marriages rather than the "stock" of extant marriages. The stock of current marriages includes marriages that were formed decades prior to the date of data collection and thus will indicate the conditions of that time period rather than the current time period. Furthermore, extant marriages reflect not only marriage patterns but also, implicitly, patterns of marriage dissolution through divorce or widowhood. Prior work has shown that interracial marriages have higher rates of divorce than intraracial marriages<!-- TODO: cite -->. Both of these factors will contribution to underestimation of the tendency toward interracial marriage in recent marriage cohorts.

However, information on marital timing is not available for a substantial subset of the time period here. From 1960 to 1980, the Census asked about the age of first marriage for all married respondents. Marital duration questions were removed from Census 1990. A question on the year of the current marriage was added to the ACS in 2007 onward. Thus, between 1990 and 2006, I have no way to identify recent marriage cohorts. And in the period before 1990, I can only identify timing of marriage for first marriages.

This issue has often been addressed by restricting the sample by the age of one or both partners [@qian_breaking_1997; \@blackwell_mate_2000]. By selecting younger partners, we can be assured that the marriage was relatively recently formed. However, this approach can also introduce a host of other biases as younger couples may differ in other ways from the full set of recent marriage cohorts. Furthermore, the age of marriage for both husbands and wives has increased substantially over this time period, meaning that any static restriction by age will lead to different kinds of couples across the time period.

Rather than subset by age, I take all extant marriages but apply weights to each couple to match the joint age distribution of newlywed partners for that time period. This approach generalizes the method of taking a restricted age group, but instead of arbitrarily deciding on an age group, I allow each age group (or joint age group) to be represented proportionally to its share of the most recent marriage cohort. Because these weights are derived from to the joint age distribution within each time period, this approach also intrinsically adjusts for changes in the timing of marriage for both husbands and wives over the time period.

From 2007 onwards, I can estimate joint age distributions of recent marriage cohorts directly from the ACS data using couples married in the last year. Prior to 2007, I use publicly available tables on the joint distribution of the age of spouses from US vital statistics reports for 1960, 1970, 1980, and 1988 <!-- TODO: cite -->. After 1988, these reports were no longer produced for marriage. From 1990 to 2006, I use linear interpolation of the joint age distribution. Appendix XX <!-- TODO: replace --> of the supplementary materials provides more information on this weighting procedure.

In addition to the age weighting, I also apply sample weights to adjust for sample design in the Census and ACS results.[^2] Weights are applied by taking the sum of weights as frequency counts. Because of the use of weights as well as complex survey design, the analytical standard errors of model parameter estimates will be underestimated. As there is no clear analytical solution to this issue with this modeling approach, I instead calculate standard errors via a boostrap technique by resampling 1000 times from the individual level data within each year.

[^2]: Because the unit of observation is a couple, I use the mean of each partner's person weight to derive the couple weight.

This approach cannot eliminate every bias from the use of the stock of marriages, including bias arise from differential rates of divorce. Appendix XX provides a comparison of the age weighted results to results based on newlyweds (married within the last year) in years where such comparison is possible. The results indicate that the age weighted results are quite close to the results for newlyweds both in magnitude and trend. For some groups, the age-weighted results are slightly lower than the results for newlyweds, as would be expected due to differential union resolution. However, these differences are slight in comparison to the overall differences across groups and time.

For analysis that includes cohabiting couples, I have no information on the duration of cohabitation. For comparative perspective with married couples, I apply the same age weighting to cohabiting couples as I do to married couples. This approach provides a kind of counterfactual case in which I can ask how different the interracial marriage patterns would be if cohabitors of a similar age as newlyweds had gotten married rather than remained in a cohabiting union.

# Results

## Trends in Outmarriage Rates

@fig-outmarriage-overall shows the overall outmarriage ratio across all married couples over time, weighted to match the joint age distribution of newlywed partners.The figure shows a consistent and relatively linear increase over time in the proportion of married couples who belong to different racial groups. In 1960, racially exogamous couples make up less than 1% of all couples <!-- TODO: check this -->. By the end of the time period, about 1 in 6 couples are in an interracial marriage. A larger than average increase in the final time period may reflect some changes to the measurement of race since 2020.

```{r}
#| label: fig-outmarriage-overall
#| fig-cap: The trend in the proportion of marriagest that are racially exogamous. Values in all years are weighted to reflect the joint age distribution of newlywed partners. 95% confidence band shown.
  
ggplot(exogamy_trend, aes(x = year, y = exogamy, group = 1))+
  geom_ribbon(aes(ymin = exogamy-1.96*exogamy_se,
                  ymax = exogamy+1.96*exogamy_se),
              alpha = 0.3, color = NA)+
  geom_line()+
  geom_point()+
  scale_y_continuous(labels = scales::percent)+
  labs(y = "percent of marriages that are racially exogamous")+
  theme_bw()
```

The simple trend shown in @fig-outmarriage-overall suggests rapidly diminishing racial boundaries. However, when we look at the outmarriage ratio separately by racial group, the picture becomes more complicated. @fig-outmarriage-race shows outmarriage ratios for each single race group over time, and separately by the sex of the spouse. As before, all results are weighted to reflect the joint age distribution of newlyweds.

```{r}
#| label: fig-outmarriage-race
#| fig-cap: The trend in the proportion of spouses of a given race and gender that are racially exogamous. Values in all years are weighted to reflect the joint age distribution of newlywed partners. 83% confidence bands shown.

exogamy_trend_race |>
  filter(!str_detect(race, "-")) |>
  ggplot(aes(x = year, y = exogamy, color = spouse, group = spouse))+
  geom_ribbon(aes(ymin = exogamy-1.83*exogamy_se,
                  ymax = exogamy+1.83*exogamy_se, 
                  fill = spouse),
              alpha = 0.3, color = NA)+
  geom_line(alpha = 0.7)+
  geom_point(size = 1, alpha = 0.7)+
  facet_wrap(~race)+
  labs(y = "percent of spouses of given race that are racially exogamous")+
  scale_y_continuous(labels = scales::percent)+
  scale_color_manual(values = c("red", "navy", "gold"))+
  theme_bw()
```

For both the AIAN and API groups, outmarriage ratios largely stopped increasing (after 1990 for AIAN and after 2000 for API) and have remained flat, or have slightly declined since. In contrast, the outmarriage ratios for White, Black, and Hispanic respondents have consistently increased over the time period.

Consistent with prior work <!-- TODO: cite -->, @fig-outmarriage-race also shows that the likelihood of racial exogamy varies by gender for Black and API spouses. Black men are much more likely to be outmarried than Black women and that gap has increased over the time period, although the overall trend is similar for both Black men and women. In contrast, the likelihood of racial exogamy is higher for API women than for API men. This difference has remained largely constistent across time and both API women and men have experienced the same plateau, or possible decline, in the likelihood of racial exogamy.

<!-- TODO: comparison to Pew report -->

@fig-outmarriage-race provides important information on the experience of interracial partnering for different groups, but the results are highly reflective of group size differences. For example, as I will show below, once we adjust for group size, it is actually the Hispanic population which tends to have the highest likelihood of racial exogamy. This pattern is not apparent in @fig-outmarriage-race, because both the level of and change in the relative size of groups are not taken into account. I now turn to an analysis of the odds of racial exogamy, which can better track trends in the likelihood of interracial marriage net of group size differences.

## Overall Trends in Odds Ratios

@fig-intermar-single-overall shows the odds of racial exogamy over time across all possible pairings of single race groups. Results are estimated from log-linear models and weighted by the joint age distribution of newlyweds at each time point. Each panel shows the results of each possible interracial pairing for a particular ego racial group. Thus, each interracial pairing is shown twice across all of the panels.

```{r}
#| label: fig-intermar-single-overall
#| fig-cap: Odds of racial exogamy across all single race pairings, separately by the focal race of spouse. Estimates are derived from log-linear models that control for state racial composition and are weighted to reflect the joint age distribution of spouses.
#| fig-height: 4.2
#| fig-pos: "p"
#| layout-ncol: 2
#| fig-subcap: 
#|    - White partner
#|    - Black partner
#|    - AIAN partner
#|    - API partner
#|    - Hispanic partner

model_single_state_bs <- model_state_bs |>
  filter(!str_detect(term, "-"))

lims <- exp(c(min(model_single_state_bs$conf.low), 
              max(model_single_state_bs$conf.high)))

model_single_state_bs |>
  filter(str_detect(term, "White")) |>
  ggplot(aes(x = year, y = exp(estimate), color = term))+
  geom_vline(xintercept = 2000, linetype = 2)+
  geom_ribbon(aes(ymin = exp(conf.low), ymax = exp(conf.high), fill = term),
              alpha = 0.3, color = NA)+
  geom_line()+
  geom_point(size = 0.5)+
  scale_y_log10(limits = lims)+
  scale_color_manual(values = palette)+
  scale_fill_manual(values = palette)+
  theme_bw()

model_single_state_bs |>
  filter(str_detect(term, "Black")) |>
  ggplot(aes(x = year, y = exp(estimate), color = term))+
  geom_vline(xintercept = 2000, linetype = 2)+
  geom_ribbon(aes(ymin = exp(conf.low), ymax = exp(conf.high), fill = term),
              alpha = 0.3, color = NA)+
  geom_line()+
  geom_point(size = 0.5)+
  scale_y_log10(limits = lims)+
  scale_color_manual(values = palette)+
  scale_fill_manual(values = palette)+
  theme_bw()

model_single_state_bs |>
  filter(str_detect(term, "AIAN")) |>
  ggplot(aes(x = year, y = exp(estimate), color = term))+
  geom_vline(xintercept = 2000, linetype = 2)+
  geom_ribbon(aes(ymin = exp(conf.low), ymax = exp(conf.high), fill = term),
              alpha = 0.3, color = NA)+
  geom_line()+
  geom_point(size = 0.5)+
  scale_y_log10(limits = lims)+
  scale_color_manual(values = palette)+
  scale_fill_manual(values = palette)+
  theme_bw()

model_single_state_bs |>
  filter(str_detect(term, "API")) |>
  ggplot(aes(x = year, y = exp(estimate), color = term))+
  geom_vline(xintercept = 2000, linetype = 2)+
  geom_ribbon(aes(ymin = exp(conf.low), ymax = exp(conf.high), fill = term),
              alpha = 0.3, color = NA)+
  geom_line()+
  geom_point(size = 0.5)+
  scale_y_log10(limits = lims)+
  scale_color_manual(values = palette)+
  scale_fill_manual(values = palette)+
  theme_bw()

model_single_state_bs |>
  filter(str_detect(term, "Hispanic")) |>
  ggplot(aes(x = year, y = exp(estimate), color = term))+
  geom_vline(xintercept = 2000, linetype = 2)+
  geom_ribbon(aes(ymin = exp(conf.low), ymax = exp(conf.high), fill = term),
              alpha = 0.3, color = NA)+
  geom_line()+
  geom_point(size = 0.5)+
  scale_y_log10(limits = lims)+
  scale_color_manual(values = palette)+
  scale_fill_manual(values = palette)+
  theme_bw()
```

The trends in @fig-intermar-single-overall are consistent with the plateau of outmarriage rates observed in @fig-outmarriage-race for AIAN and API partners. Across all possible combinations, racial exogamy involving an API partner had either plateaued by the late 20th century or has shown no evidence of a rise at all over the time period. The results for AIAN partners are somewhat similar, except for some evidence of a moderate rise in the odds of marriage to Hispanic and Black partners in the later time period.

In contrast, in all cases involving combinations of White, Black, and Hispanic partners, the odds of interracial marriage consistently increase exponentially (linear on the scale shown) throughout the time period. Throughout the 20th century, the odds of interracial marriage with a black spouse is the least likely of all possible non-endogamous pairings for each non-Black group. However, in many cases, these odds ratios grow the fastest throughout the time period. By the end of the time period, partnership with a Black spouse is only definitively the least likely from of interracial marriage for White partners. Even in this case, the odds of White/Black partnering is increasing much more rapidly than other pairings involving a White partner. If the trends observed here continue, White/Black interracial marriage will soon become more likely than White/API interracial marriage.

Interracial marriages involving a Hispanic partner are generally some of the most likely for all groups. White/Hispanic partnering is the most likely form of interracial marriage throughout the time period in which Hispanics can be identified in the data (although roughly tied with White/AIAN partnering in 1980 and 1990). Interracial marriage with Hispanic partners is also the most likely form of partnering for Black spouses throughout the time period and the second most likely form of partnering for AIAN and API spouses after marriage with White partners for most of the time period.

## Including Multiracial Groups

To this point, I have only examined the interracial marriage patterns between single race groups. The option to check more than one box on the race question in 2000 allows me to analyze trends in interracial marriage for some large multiracial groups. @fig-intermar-multi-overall shows those trends for the three largest multiracial groups of White-Black, White-AIAN, and White-API. I examine the tendency of each of these groups to marry a member from each of their two constituent groups, in comparison to the odds of a "first generation" marriage between those constituent groups. For example, for the White-Black case, I examine the odds of a White-Black/White and White-Black/Black marriage to a White/Black marriage.

```{r}
#| label: fig-intermar-multi-overall
#| fig-cap: Odds of racial exogamy for biracial groups with component single race groups and in comparison to the odds of racial exogamy for the component single race groups. Estimates are derived from log-linear models that control for state racial composition and are weighted to reflect the joint age distribution of spouses.
#| layout-ncol: 2
#| fig-subcap: 
#|    - White-Black
#|    - White-AIAN
#|    - White-API

model_state_bs |>
  filter(term == "White/Black" | str_detect(term, "White-Black")) |>
  ggplot(aes(x = year, y = exp(estimate), color = term))+
  geom_vline(xintercept = 2000, linetype = 2)+
  geom_ribbon(aes(ymin = exp(conf.low), ymax = exp(conf.high), fill = term),
              alpha = 0.3, color = NA)+
  geom_line()+
  geom_point(size = 0.5)+
  scale_y_log10()+
  scale_color_manual(values = palette)+
  scale_fill_manual(values = palette)+
  theme_bw()

model_state_bs |>
  filter(term == "White/AIAN" | str_detect(term, "White-AIAN")) |>
  ggplot(aes(x = year, y = exp(estimate), color = term))+
  geom_vline(xintercept = 2000, linetype = 2)+
  geom_ribbon(aes(ymin = exp(conf.low), ymax = exp(conf.high), fill = term),
              alpha = 0.3, color = NA)+
  geom_line()+
  geom_point(size = 0.5)+
  scale_y_log10()+
  scale_color_manual(values = palette)+
  scale_fill_manual(values = palette)+
  theme_bw()

model_state_bs |>
  filter(term == "White/API" | str_detect(term, "White-API")) |>
  ggplot(aes(x = year, y = exp(estimate), color = term))+
  geom_vline(xintercept = 2000, linetype = 2)+
  geom_ribbon(aes(ymin = exp(conf.low), ymax = exp(conf.high), fill = term),
              alpha = 0.3, color = NA)+
  geom_line()+
  geom_point(size = 0.5)+
  scale_y_log10()+
  scale_color_manual(values = palette)+
  scale_fill_manual(values = palette)+
  theme_bw()
```

For both the White-Black and the White-API case, we see a similar pattern. Marriages to each of the two constituent groups are considerably more likely than a "first generation" marriage between the two constituent groups. This finding suggests that White-Black and White-API individuals face lower boundaries to interracial marriage with these groups because they are in some sense partially endogamous with both groups [@gullickson_blurring_2025].

For the White-Black group, one might expect that, due to the one-drop rule <!-- TODO: CITE -->, White-Black individuals would be more likely to marry Black rather than White partners. The odds of partnering with Black spouse is slightly higher than the odds of partnering with a white spouse in the early period, but this slight difference has gone away by the end of the time period. Thus, the results here suggest a very weak role of the one-drop rule in the marriage selection choices of White-Black individuals. For White-API individuals, The odds of marriage with White and API partners are very similar. At the very end of the period, we see some evidence that the odds or partnering with White spouse might be growing faster than partnering with API spouse, but this finding should be considered tentative.

For White-AIAN individuals, we observe a very different pattern. Partnering with White spouses is more common than "first generation" White/AIAN interracial marriage and the odds of this form of intermarriage are growing over time, while the odds of partnership with AIAN spouses is less likely than "first generation" White/AIAN interracial marriage and does not appear to be increasing over time. These results are consistent with a norm of hyperdescent in which individuals of White/AIAN ancestry are more likely to be seen and treated as White than AIAN <!-- TODO: CITE -->.

The change in 2000 to multiracial reporting may also affect trends in single race interracial marriage, depending on which individuals were more likely to reclassify themselves as multiracial. While we can never know precisely the affect of this shift in reporting, I develop three counterfactual scenarios to better understand how the trends in interracial marriage might have looked in cases where multiracial respondents remained in single race categories. For sample size reasons, I focus on the three categories of White/Black, White/API, and White/AIAN interracial marriage. In each case, I consider three scenarios for reclassification of part-White multiracial respondents. In the first scenario, all these respondents are re-classified as White. In the second scenario, all these respondents are classified by their minority race. In the third scenario, I use the "best race" variable created by the Minnesota Population Center. This variable is based on multinomial logit models of data from the National Health Interview Survey, which asks multiracial respondent for their preferred single race.

```{r}
#| label: fig-multi-wb-counterf
#| fig-cap: Comparison of part-White exogamy trends with different counterfactual assumptions about single race reporting among multiracial respondents
#| layout-ncol: 2
#| fig-subcap: 
#|    - White/Black
#|    - White/API
#|    - White/AIAN

model_state_bs |>
  mutate(counterfactual = "Actual") |>
  bind_rows(model_cf_white_state_bs) |>
  bind_rows(model_cf_minority_state_bs) |>
  bind_rows(model_cf_best_state_bs) |>
  filter(term == "White/Black") |>
  mutate(counterfactual = if_else(counterfactual == "Minority", "Black",
                                  counterfactual),
         counterfactual = factor(counterfactual, 
                                 levels = c("Actual", "White", "Black", 
                                            "Best"))) |>
  ggplot(aes(x = year, y = exp(estimate), color = counterfactual))+
  geom_vline(xintercept = 2000, linetype = 2)+
  geom_ribbon(aes(ymin = exp(conf.low), ymax = exp(conf.high), 
                  fill = counterfactual),
              alpha = 0.3, color = NA)+
  geom_line()+
  geom_point(size = 0.5)+
  scale_y_log10()+
  scale_color_manual(values = c("red", "gold", "navy", "darkgreen"))+
  scale_fill_manual(values = c("red", "gold", "navy", "darkgreen"))+
  theme_bw()

model_state_bs |>
  mutate(counterfactual = "Actual") |>
  bind_rows(model_cf_white_state_bs) |>
  bind_rows(model_cf_minority_state_bs) |>
  bind_rows(model_cf_best_state_bs) |>
  filter(term == "White/API") |>
  mutate(counterfactual = if_else(counterfactual == "Minority", "API",
                                  counterfactual),
         counterfactual = factor(counterfactual, 
                                 levels = c("Actual", "White", "API", 
                                            "Best"))) |>
  ggplot(aes(x = year, y = exp(estimate), color = counterfactual))+
  geom_vline(xintercept = 2000, linetype = 2)+
  geom_ribbon(aes(ymin = exp(conf.low), ymax = exp(conf.high), 
                  fill = counterfactual),
              alpha = 0.3, color = NA)+
  geom_line()+
  geom_point(size = 0.5)+
  scale_y_log10()+
  scale_color_manual(values = c("red", "gold", "navy", "darkgreen"))+
  scale_fill_manual(values = c("red", "gold", "navy", "darkgreen"))+
  theme_bw()

model_state_bs |>
  mutate(counterfactual = "Actual") |>
  bind_rows(model_cf_white_state_bs) |>
  bind_rows(model_cf_minority_state_bs) |>
  bind_rows(model_cf_best_state_bs) |>
  filter(term == "White/AIAN") |>
  mutate(counterfactual = if_else(counterfactual == "Minority", "AIAN",
                                  counterfactual),
         counterfactual = factor(counterfactual, 
                                 levels = c("Actual", "White", "AIAN", 
                                            "Best"))) |>
  ggplot(aes(x = year, y = exp(estimate), color = counterfactual))+
  geom_vline(xintercept = 2000, linetype = 2)+
  geom_ribbon(aes(ymin = exp(conf.low), ymax = exp(conf.high), fill = counterfactual),
              alpha = 0.3, color = NA)+
  geom_line()+
  geom_point(size = 0.5)+
  scale_y_log10()+
  scale_color_manual(values = c("red", "gold", "navy", "darkgreen"))+
  scale_fill_manual(values = c("red", "gold", "navy", "darkgreen"))+
  theme_bw()
```

@fig-multi-wb-counterf presents the trends resulting from these different scenarios. For White/Black respondents, the overall trend is quite similar across all scenarios, although all three scenarios would have slightly increased the odds of White/Black marriage over that actually observed after 2000.

For both White/API and White/AIAN interracial marriage, the scenario which differs from the other cases is that where multiracial part-White respondents all identify with their single-race minority group (API and AIAN, respectively). In this scenario, the odds of interracial marriage are substantially higher than the other two scenarios and the actual observed trend. The other two scenarios produce results that are statistically indistinguishable from the actual trend. The results here suggest that many part-White multiracial individuals have White spouses and thus reclassifying these individuals into single race minority groups would increase the observed odds of interracial marriage with white spouses. However, this assumption is likely a strong and unrealistic assumption for individuals of partial API and AIAN ancestry. Overall, the results suggest that the shift to multiracial reporting likely had relatively little effect on the trends for interracial marriage among single race groups.

## Trends Among Hispanic Subgroups

I now turn to results that divide the Hispanic group into three separate subgroups, identified by racial identification as "White", "Black", and "Other." @fig-sub-hispanic shows trends in the odds ratio of marriage between these three subgroups and non-Hispanic White and Black respondents. For reference, the trend in White/Black interracial marriage is also shown.

```{r}
#| label: fig-sub-hispanic
#| fig-cap: Exogamy of Hispanic subgroups with White and Black partners. White/Black exogamy shown for reference
#| layout-ncol: 2
#| fig-subcap:
#|   - White partner
#|   - Black partner

lims <- exp(c(min(model_hispanic_sub_state_bs$conf.low), 
              max(model_hispanic_sub_state_bs$conf.high)))

model_hispanic_sub_state_bs |>
  filter(!str_detect(term, "Hispanic/") & !str_detect(term, "^Black")) |>
  ggplot(aes(x = year, y = exp(estimate), color = term))+
  geom_vline(xintercept = 2000, linetype = 2)+
  geom_ribbon(aes(ymin = exp(conf.low), ymax = exp(conf.high), fill = term),
              alpha = 0.3, color = NA)+
  geom_line()+
  geom_point(size = 0.5)+
  scale_y_log10(limits = lims)+
  scale_color_manual(values = palette)+
  scale_fill_manual(values = palette)+
  theme_bw()

model_hispanic_sub_state_bs |>
  filter((str_detect(term, "^Black/") & !str_detect(term, "Hispanic/")) |
           str_detect(term, "^White/Black$")) |>
  ggplot(aes(x = year, y = exp(estimate), color = term))+
  geom_vline(xintercept = 2000, linetype = 2)+
  geom_ribbon(aes(ymin = exp(conf.low), ymax = exp(conf.high), fill = term),
              alpha = 0.3, color = NA)+
  geom_line()+
  geom_point(size = 0.5)+
  scale_y_log10(limits = lims)+
  scale_color_manual(values = palette)+
  scale_fill_manual(values = palette)+
  theme_bw()
```

In terms of interracial marriage with White spouses shown in the left panel, the ordering of the three subgroups is as expected. Hispanic individuals who also identify as White are the most likely to marry non-Hispanic White spouses. The odds of marriage to White spouses is somewhat lower for Hispanic individuals who identify as "Other" by race. For comparison, both of these odds ratios are substantially higher than the odds of White/Black marriage. The odds of marriage to a White spouse are much lower for Hispanic individuals who identify as "Black" and are similar to the odds of White/Black marriage.

Although the odds are lowest for Black Hispanics, the trend suggests much a faster increase in this odds ratio over time, on par, with the increase in the odds of White/Black marriage. Thus, while there is a clear stratification of the odds of marriage with Whites by Hispanic racial self-identification, these differences have been getting smaller over time.

The right panel shows the odds of interracial marriage between non-Hispanic Black respondents and the three Hispanic subgroups. As one might expect, these odds are inverted relative to the odds of interracial marriage with non-Hispanic White respondents. Interracial marriage between non-Hispanic Black spouses and Hispanics who identify as either Other or White are both similar to the odds of White/Black interracial marriage over time. The odds of interracial marriage between non-Hispanic Black spouses and Hispanics who identify as Black, on the other hand, are much higher and are comparable to the odds of White/White Hispanic interracial marriage. All three odds ratios are growing exponentially over time.

## Trends Among API Subgroups

@fig-sub-asian-simple shows the odds of partnership with white spouses across three subgroups of the API category. I divide the API group into Pacific Islanders, South Asians, and East/Southeast Asians. The results show that the plateau trends seen for the API group overall is only the trend for the East/Southeast Asian group. This group saw rapid increase in the odds of partnership with White spouses until 1990, when those odds leveled off. Since 1990, the odds have declined slightly, but have largely remained stationary.

```{r}
#| label: fig-sub-asian-simple
#| fig-cap: Exogamy of Asian subgroups with White partners

model_asian_sub_simple_state_bs |>
  filter(str_detect(term, "^White/")) |>
  mutate(term = str_replace(term, "/EastAsian", "/E&SE Asian"),
         term = str_replace(term, "/PI", "/Pac. Islander"),
         term = str_replace(term, "/SouthAsian", "/South Asian")) |>
  ggplot(aes(x = year, y = exp(estimate), color = term))+
  geom_vline(xintercept = 2000, linetype = 2)+
  geom_ribbon(aes(ymin = exp(conf.low), ymax = exp(conf.high), fill = term),
              alpha = 0.3, color = NA)+
  geom_line()+
  geom_point(size = 0.5)+
  scale_y_log10()+
  scale_color_manual(values = palette)+
  scale_fill_manual(values = palette)+
  theme_bw()
```

In contrast, the results for South Asians suggest increasing odds of marriage with White spouse throughout the time period, although sparsity of data from before 1980 limits the full range of the time period. Point estimates suggest a decline from 1980 to 1990, but the wide error margins on these estimates make such a decline questionable at best. Since 1990, there has been steady growth in the odds of South Asian/White partnering. However, those odds are begin and still remain substantially lower than the odds of East and South Asian/White partnering, even as they are converging across time.

The results for Pacific Islanders show a decline in the odds of marriage with White partners throughout the time period. The decline is particularly concentrated in the period before 2000. Although this group is small, this trend before 2000 is not simply a result of wide error margins on estimates, and suggests a true decline in the likelihood of marriage between these groups. After 2000, the decline is slight and may simply be a result of measurement error.

The decline for Pacific Islander is notable as the only trend examined here to show a substantial decline across this time period, which is otherwise characterized by substantial increases in the odds of interracial marriage in the late 20th century. Notably, the starting odds of White/Pacific Islander marriage are relatively very high. Given that Pacific Islander identification has been unstable over the time period (CITE), it is possible that this trend might in some part reflect changing self-identification practices.

## Including Cohabitation

@fig-intermar-cohab compares the odds of an interracial union across the ten single race combinations, separately for marriage and cohabitation. The results for cohabitation are only available since 1990. As expected, the odds of interracial cohabitation are generally higher than the odds of interracial marriage. The White/AIAN case, where the odds are extremely similar across the time period, is the only case to deviate from this pattern. Furthermore, the overall trend in cohabitation matches the trend for interracial marriage for each combination. Therefore, it is unlikely that trends in interracial marriage are driven by a changes in the frequency of cohabitation over the time period.

```{r}
#| label: fig-intermar-cohab
#| fig-cap: Odds of racial exogamy for single race groups by different union types

model_state_bs <- model_state_bs |>
  mutate(union_type = "marriage")

#model_state_combine_bs <- model_state_combine_bs |>
#  mutate(union_type = "both")

model_state_cohab_bs |>
  mutate(union_type = "cohabitation") |>
  #bind_rows(model_state_combine_bs) |>
  bind_rows(model_state_bs) |>
  filter(!str_detect(term, "-")) |>
  mutate(union_type = factor(union_type, 
                             levels = c("marriage", "cohabitation", "both"))) |>
  ggplot(aes(x = year, y = exp(estimate), color = union_type))+
  geom_vline(xintercept = 2000, linetype = 2)+
  geom_ribbon(aes(ymin = exp(conf.low), ymax = exp(conf.high), fill = union_type),
              alpha = 0.3, color = NA)+
  geom_line()+
  geom_point(size = 0.5)+
  facet_wrap(~term)+
  scale_y_log10()+
  scale_color_manual(values = c("red", "navy", "darkgreen"))+
  scale_fill_manual(values = c("red", "navy", "darkgreen"))+
  theme_bw()
```

For most of the cases, there is also some convergence over time in the odds of interracial cohabitation and marriage, although the odds of interracial marriage are still lower at the end of the time period in all cases except for White/AIAN. Thus, although the barriers of entry into interracial cohabitation are lower than the barriers of entry into interracial marriage, the difference in those barriers is weakening over time, possibly reflecting a greater deterioration in meaning betweeen cohabitation and marriage over time.

# Conclusions
