---
title: "Trends and Heterogeneity in Interracial Marriage in the United States, 1960-2023"
shorttitle: "Trends in Interracial Marriage, 1960-2023"
shortauthors: "Gullickson"
keywords: [interracial marriage, trends, racial exogamy]
thanks: Prepared for the 2026 Annual Meeting of the Population Association of America, St. Louis, Missouri
reference-section-title: References
bibliography: ../bibliography/project.bib
csl: ../bibliography/chicago-parenthetical.csl
filters:
  - abstract-section
format:
  aog-article-pdf:
    papersize: letter
    keep-tex: true
    #mainfont: Baskerville 
    #sansfont: Futura-Medium
    #fontsize: 12pt   
    #geometry: margin=1in
    #fig-height: 4     # smaller fig heights make floating easier
    fig-width: 7.5    # set to the (full width - margins) of letter 
    fig-pos: "!t"
    colorlinks: true
    urlcolor: red
  #submittable-docx: 
  #  fig-format: pdf
  #  out-width: "100%"
  #  fig-width: 6.25 # theoretically it should be 6.5 with 1 inch margins
  #submittable-pdf:
  #  keep-tex: false
  #  fig-pos: "!t"
# header-includes to resposition subcaptions
header-includes:
  - \usepackage{caption}
  - \captionsetup[subfigure]{justification=centering}
---

```{r}
#| label: setup
#| include: false
library(here)
source(here("utils","check_packages.R"))
source(here("utils","functions.R"))
```

```{r}
#| label: load-data

load(here("data", "data_constructed", "data_couples.RData"))
load(here("data", "data_constructed", "exogamy_trends.RData"))
load(here("data", "data_constructed", "model_state_bs.RData"))
load(here("data", "data_constructed", "model_basic_bs.RData"))
load(here("data", "data_constructed", "models_subgroups.RData"))
load(here("data", "data_constructed", "models_counterfactual.RData"))
load(here("data", "data_constructed", "model_region.RData"))
load(here("data", "data_constructed", "models_state_gen_bs.RData"))
load(here("data", "data_constructed", "model_recent_mar.RData"))
load(here("data", "data_constructed", "models_cohab.RData"))
load(here("data", "data_constructed", "model_recent_mar.RData"))
load(here("data", "data_constructed", "model_same_sex.RData"))
```

```{r}
#| label: set-color-palette

palette <- c(
  # Single race combinations
  "White/Black"                   = "#FF3B00",  # rich orange
  "White/AIAN"                    = "#00785E",  # teal green
  "White/API"                     = "#00A6D6",  # muted magenta
  "White/Hispanic"                = "#661100",  # deep rust
  "Black/AIAN"                    = "#E6D945",  # warm brown
  "Black/API"                     = "#74AA99",  # muted cyan
  "Black/Hispanic"                = "#0B3D91",  # deep burgundy
  "AIAN/API"                      = "#F05A1A",  # desaturated turquoise
  "AIAN/Hispanic"                 = "#C0408C",  # dusty rose
  "API/Hispanic"                  = "#FFC107",  # yellow-gold
  
  # Single race to multiracial combinations
  "White/White-Black"             = "#3E8F0D",  # vivid spring green
  "Black/White-Black"             = "#471624",  # deep maroon-brown
  "White/White-AIAN"              = "#361892",  # royal indigo
  "AIAN/White-AIAN"               = "#C9A227",  # bright teal-blue
  "White/White-API"               = "#CC6666",  # slate blue
  "API/White-API"                 = "#6D7003",  # olive drab
  "Black/Black-API"               = "#1117AA",  # bold cobalt blue
  "API/Black-API"                 = "#6D4113",  # dark umber
  "Black/Black-AIAN"              = "#7171BB",  # muted periwinkle
  "AIAN/Black-AIAN"               = "#422D13",  # earthy espresso

  # Hispanic subgroups with White and Black
  "White/White Hispanic"           = "#552288",  # soft purple
  "White/Black Hispanic"           = "#E1A400",  # rose plum
  "White/Other Hispanic"           = "#287B33",  # leafy green
  "Black/White Hispanic"           = "#33A02C",  # olive yellow
  "Black/Black Hispanic"           = "#9C2C5C",  # reddish brick
  "Black/Other Hispanic"           = "#2C37DA",  # vivid sapphire

  # Asian subgroups with White
  "White/EastAsian"               = "#3366AA",  # steely blue
  "White/E&SE Asian"              = "#3311EE",  # steel sky blue
  "White/SEAsian"                 = "#118877",  # sea green
  "White/South Asian"             = "#009E73",  # muted violet-magenta

  # Indigenous subgroups with White
  "White/AIAN"                    = "#CC4477",  # muted raspberry rose
  "White/Pac. Islander"           = "#FF8C00",  # orange red
  
  # Intra-Hispanic
  "WhiteHispanic/BlackHispanic"   = "#BB8866",  # muted tan
  "WhiteHispanic/OtherHispanic"   = "#885577",  # dusty mauve
  "BlackHispanic/OtherHispanic"   = "#AA8844",  # muted amber
  
  # Intra-Asian
  "EastAsian/SEAsian"             = "#336633",  # forest moss
  "EastAsian/SouthAsian"          = "#993399",  # grape violet
  "SEAsian/SouthAsian"            = "#CC6666",  # muted rose

  # Intra-Indigenous
  "AIAN/PI"                       = "#447777"   # muted teal gray
)

```

# Abstract

While scholars recognize that interracial marriage in the United States has grown exponentially since the 1960s,  our understanding of this trend for specific racial pairings is piecemeal and thin, owing to issues with data comparability and lack of information on marriage timing between 1980 and 2007. In this article, I use log-linear models of Census and American Community Survey data to analyze trends in interracial marriage from 1960 to 2023 across major single-race groups, as well as a some important subgroups. Weights are applied to match the joint age distribution of currently married partners to the age distribution of newlyweds in the same time period. While the overall trend is one of exponential increase, important heterogeneity is visible when examining particular racial pairings or subgroups. Most notably, the likelihood of interracial marriage for East/Southeast Asian and American Indian/Alaska Native spouses plateaued by the end of the 20th century and has not continued to grow. On the other hand, the likelihood of interracial marriage for Black spouses, although far lower than any other group at the start of the period, has increased more rapidly than any other group.  Overall, these results indicate an ongoing transformation of group boundaries. 

# Introduction

Since the 1960s, the frequency of marriage between spouses who identify with different racial groups has risen at an unprecedented pace in the United States [@gullickson_black_2006; @rosenfeld_racial_2008]. Although interracial marriage remains rarer than other forms of heterogamy such as educational or religious intermarriage, the growth in interracial marriage has been more substantial than for other forms of heterogamy over this same time period [@rosenfeld_racial_2008].

This trend over roughly the last half century has been one of the most important features of the changing racial demography of the United States. Interracial marriage is widely seen as the engine of the "biracial baby boom" in which an increasing number of people desire to identify with more than one race [@root_racially_1992]. It is a general presumption among the public that interracial marriage rates have grown exponentially and will continue to rise for the foreseeable future, signaling weakening social boundaries between racial groups. This belief in the inevitability of interracial marriage helps to inform conceptions of a "post-racial" America in which everyone is mixed to some degree or another.

Given its importance, our knowledge of trends in interracial marriage over this period remains surprisingly piecemeal and thin. We lack a systematic examination of the trend in interracial marriage between most specific racial groups over the entire time period. @qian_changing_2011, which is one of the most cited scholarly treatments of the subject, only examines two points in time, 1980 and 2008. @rosenfeld_racial_2008 only measures endogamy rather than the likelihood of specific interracial combinations (e.g. Black/White, Black/Asian). @gullickson_black_2006 only examines interracial marriage between White and Black partners. None of these studies examine data beyond 2008. The most comprehensive recent analysis comes from the Pew Research Center [@livingston_intermarriage_2017]. This report again uses two points in time, 1980 and 2015, to draw conclusions and focuses on simple outmarriage rates that do not account for group size differences and cannot be used to measure the frequency of particular interracial pairings.

In this article, I develop a systematic examination of the trends in interracial marriage from 1960 to 2023, using data from the US Census and the American Community Survey. I address the primary methodological challenge in developing such measures - the lack of comparable information on marital duration for much of the time period - by the use of weights that make each sample representative of the joint distribution of newlyweds in the same time period. Additionally, I handle a variety of other methodological challenges arising from changes in measurement across the time period, including the switch to a multiple check-off response to the race question since 2000. The goals of this paper are largely descriptive. However, given the overall importance of these trends and the piecemeal nature of our existing body of knowledge, such a comprehensive descriptive analysis is long overdue.

Many of my findings echo and reinforce previous work. I find, for example, that the odds of interracial marriage between Black and non-Black spouses have been lower than for other types of interracial marriage, but these odds have also grown the fastest over time. However, importantly, I also find that the story of uninterrupted weakening of boundaries is inaccurate for some groups. The odds of exogamy for Asian/Pacific Islanders and for American Indian/Alaska Natives increased in the early period under observation, but plateaued by 1990 and have not increased since. Further analysis across a variety of subgroups reveals important heterogeneity in trends and levels of interracial marriage.

# Background

The significance of interracial marriage to social systems is twofold.[^1] First, the likelihood of interracial marriage is seen as a key indicator of the social distance between groups and the strength and rigidity of the boundary separating those groups [@gordon_assimilation_1964]. Interracial marriage provides a window into the assimilation and integration of groups within the most intimate sphere of social life. For example, the almost complete social (and often legal) prohibition on marriage between Black and White partners for much of US history <!-- TODO: CITE --> is a testament to the stark impermeability of the White/Black boundary.

[^1]: I use the term "interracial" here because I will focus on marriage between groups that are identified by a racialized boundary, but the logic could apply equally well to "intergroup" marriage where groups are defined by ethnicity, religion, etc.

Second, an increase in the likelihood of interracial marriage is seen as a harbinger of intergenerational change because the children from such unions can plausibly claim possession of both group identifications [@root_racially_1992]. Such children then potentially represent another source by which group boundaries can be weakened, by invalidating the distinctiveness of group identity itself. This argument comes with important caveats. Marriage is not the only means by which to produce "mixed" children. For example, for most of US history, White/Black mixing predominantly occurred outside of marriage [@hodes_white_1997; @kennedy_interracial_2003]. Furthermore, norms of classification such as the "one drop rule" can often effectively stunt any challenge to group boundaries arising from "mixed" progeny [@johnston_race_1970; @davis_who_1991]. Nonetheless, the "biracial baby boom" resulting from rising interracial marriage in the post-Civil Rights period has played an important role in restructuring our understanding, and even measurement, of racial boundaries [@lee_americas_2004].

Measuring the frequency of interracial marriage is complicated by differences in group size. These differences will play an important role in who marries whom, because they determine the racial distribution of potential partners. Whether or not group size is accounted for in a measure of interracial marriage can lead to dramatic differences in our assessment of its frequency. A popular measure of interracial marriage is the outmarriage rate, or ratio. Typically, the outmarriage ratio measures the percent of individuals from a given group that are married to someone belonging to a different group, although it can also be measured across the population at large. The measure is intuitive and easy to understand. However, it is also heavily influenced by relative group size.

A toy example easily demonstrates why this is so. Imagine a population divided by two groups A and B, where A constitutes 85% of the population and B constitutes the remaining 15%. Among B members looking for a romantic partner, 85% of the possible partners they can choose will belong to the other group, while among A members only 15% of possible partners will belong to the other group. The two groups have very different opportunities for intergroup partnership. If both groups were to prefer an outgroup marriage to an ingroup marriage, we would expect a 100% outmarriage rate for group B, whose members can easily find a partner among the larger group A, while the  outmarriage rate for group A is limited to 15% even when all group B members marry a member of group A. This difference in outmarriage rates between groups tells us nothing useful except that the groups differ in size.

For this reason, most analyses of intergroup marriage uses an odds ratio to measure the likelihood of group exogamy and endogamy. The odds ratio implicitly controls for the marginal distribution of groups and thus provides a measure that is independent of group size differences. In the example above, a single odds ratio would estimate the likelihood of marriage between groups A and B. 

Because it adjusts for group size, the odds ratio is considered a superior measure of underlying social distance. However, the outmarriage ratio also provides important information about the experience of intergroup contact across different populations. The tension between these two measures reflects the two primary reasons for studying intermarriage noted above. The odds ratio provides a better measure of the strength and rigidity of boundaries, while the outmarriage ratio provides a better measure of the actual frequency of interracial families in the population. 

Scholars also differ in whether they measure the likelihood of group endogamy/exogamy or the likelihood of specific pairings of racial groups. @rosenfeld_racial_2008 and @livingston_intermarriage_2017, for example, only measure general patterns of endogamy/exogamy for each focal racial group. These general patterns are easier to estimate because they pool the specific racial outgroups together, but they also make it impossible to detect different trends for specific pairings of groups, such as White/Black or Black/Asian interracial marriage. @qian_changing_2011 provide estimates of exogamy for specific pairings of racial groups in their log-linear models, but only report results for pairings involving a White partner, meaning we know little about trends or even levels of interracial marriage for dual minority pairings.

Despite different measurement approaches and the piecemeal nature of prior work, some potential results have been identified across studies. First, the racial exogamy of Black partners is historically lower than for other groups, but exogamy has increased faster and more consistently for Black partners than for other groups, suggesting some convergence [@rosenfeld_racial_2008; @qian_changing_2011; @gullickson_patterns_2022]. <!-- TODO: Rework discussion to include Qian and Lichter and discuss the more ambivalent results for Latinos --> Second, the increase in exogamy for Asian/Pacific Islander and American Indian/Alaska Native populations may have slowed down or stalled in more recent years, although the magnitude of this slowdown is unclear [@qian_social_2007;@qian_changing_2011; @livingston_intermarriage_2017]. @livingston_intermarriage_2017 actually find a decline in the outmarriage rate for Asian spouses from 1980 to 2015 and no change for Latino spouses as well.  However, these results are somewhat difficult to interpret due to the use of outmarriage rates that do not adjust for group size and the inclusion of foreign-born individuals into the sample (an issue I discuss more below). Other work, using log-linear modeling, has shown Latino exogamy to be higher than for other groups, although the trend in Latino exogamy is less clear [@rosenfeld_racial_2008; @qian_changing_2011].

<!-- TODO: add a paragraph to discuss the theory that immigration accounts for the leveling off/decline of interracial marriage for Asians -->

Several other limitations are present in our existing body of knowledge. First, most of the prior work is limited in the years that are used. @qian_changing_2011, @livingston_intermarriage_2017, and @gullickson_patterns_2022 all use two time points starting from 1980, but two time points do not allow us to fully extrapolate a trend. The reason for this approach is the lack of information on marital duration in many data sources. Ideally, to measure trends in interracial marriage, researchers would use information on newlyweds. However, information on marital timing was removed after the 1980 Census and only re-introduced to the American Community Survey from 2008 onward. Furthermore, the 1980 data only provided marital duration for first marriages, making the comparison between 1980 and more recent years somewhat difficult. Researchers either have to decide on a slightly apples-to-oranges comparison between first marriages in 1980 and all marriages in more recent data, or they must restrict the analysis to first marriages throughout.

Second, while odds ratios adjust for group size, they only do so at the national level. Because racial groups are not equally distributed throughout the US, these national differences in group size may not reflect the actual distribution of groups in a person's local marriage market [@harris_how_2005]. These differences will tend to make interracial marriage appear less common than it is in practice. Furthermore, shifts in the geographic distribution of racial groups over time could be misinterpreted as trends in interracial marriage.

Third, existing work has largely focused on large single race groups, for reasons of sample size. However, we know relatively little about how these trends might change by analyzing subgroups within these larger aggregates, nor do we have much knowledge of trends or tendencies in interracial marriage for multiracial groups. @gullickson_patterns_2022 analyzes patterns of interracial marriage for Asian and Latino ethnonational groups and finds substantial differences in the patterns and trends for East/Southeast Asians and South Asians, but this analysis only uses two points in time (1980 and 2014-2018). @littlejohn_race_2019 and @gullickson_blurring_2025 both examine the marriage choices of multiracial individuals, but only at one point in time.

Fourth, the rise in cohabitation and same-sex marriage over this time period may complicate our overall understanding of interracial union formation. Prior work has shown that exogamy across a variety of characteristics is more common in cohabitation than marriage [@blackwell_homogamy_2004;@joyner_interracial_2005], likely reflecting a winnowing process as individuals move toward more permanent unions. However, the rising prevalence of cohabitation as a potential alternative to marriage may increasingly offer an alternative form of union for interracial couple and thus affect overall trends in interracial marriage.

Early work on same-sex cohabitors demonstrated that same-sex unions are less exogamous than comparable opposite-sex unions [@jepsen_empirical_2002; @rosenfeld_independence_2005; @schwartz_assortative_2009] and more recent data on same-sex marriages has found similar results [@manning_cohabitation_2022; @hirschl_eight_2024; @song_crossing_2025]. The greater exogamy observed in same-sex unions might reflect differential availability of partners, more liberal attitudes overall, or the fact that the transgression of one norm may increase the likelihood of other transgressions. As same-sex marriage becomes more normalized in American discourse, we might expect the gap in exogamy between same-sex and opposite-sex couples to diminish. Yet, existing work has largely used simple tabular comparisons at a single point in time.

In what follows, I attempt to resolve many of these shortcomings to produce an analysis of trends in interracial marriage over time from 1960 to 2023. By using weights to adjust extant marriages to the joint age distribution of newlywed partners, I am able to produce accurate results for the entire time period, and leverage larger sample sizes to examine interracial marriages across a wider variety of groups.

# Data and Methods

Data for this analysis are derived from the microdata Census samples for Census 1960, 1970, 1980, 1990, and 2000, and American Community Survey (ACS) data from 2001 to 2023, with the exception of 2020. All data were extracted via the Integrated Public Use Microdata Sample system [@ruggles_ipums_2025]. ACS data from 2020 is removed due to issues with data collection during the COVID pandemic. To improve statistical power for the smaller ACS samples, single year samples are pooled into intervals from 2001-2005, 2006-2010, 2011-2015, 2016-2019, and 2021-2023. The midpoint of the interval is used for analyzing trends.

The primary data used for analysis include all extant heterosexual married couples in each sample. I focus primarily on heterosexual marriages because this is the only union type that is consistently identifiable throughout the entire time period. Cohabiting partners can also be identified since 1990 at which point "unmarried partner" was provided as an option for relationship to head of household. Same-sex marriages can be identified since 2013 when the Census Bureau stopped recoding same-sex married partners to unmarried partners. To compare the trends for these union types with heterosexual married couples, I also analyze trends in overall interracial coupling for opposite-sex cohabitations from 1990 to 2023 and same-sex marriages from 2013-2023.[^same-sex-cohab] @tbl-sample-size shows the sample size of each union type by time period for the full analysis.

```{r}
#| label: tbl-sample-size
#| tbl-cap: Data source and sample size of unions among native-born partners by time period and union type.

census |>
  filter(gen_husband == "2nd+" & gen_wife == "2nd+") |>
  group_by(year, union_type) |>
  summarize(n = n(), .groups = "drop") |>
  pivot_wider(names_from = union_type, values_from = n) |>
  rename(ssmarriage = `same-sex marriage`) |>
  mutate(
    year = as.character(year),
    year = case_when(
      year == "2003" ~ "2001-2005",
      year == "2008" ~ "2006-2010",
      year == "2013" ~ "2011-2015",
      year == "2017.5" ~ "2016-2019",
      year == "2022" ~ "2021-2023",
      TRUE ~ year
    ),
    data_source = c(rep("Census", 5), rep("ACS", 5))
  ) |>
  select(year, data_source, marriage, cohabitation, ssmarriage) |>
  gt() |>
  sub_missing(columns = everything(), missing_text = "-") |>
  fmt_integer(c(marriage, cohabitation, ssmarriage)) |>
  cols_label(
    year ~ "Time period",
    marriage ~ "marriage",
    cohabitation ~ "cohabitation",
    ssmarriage ~ "marriage",
    data_source ~ "Data source"
  ) |>
  tab_spanner(c(marriage, cohabitation), label = "Opposite-sex") |>
  tab_spanner(c(ssmarriage), label = "Same-sex") |>
  tab_footnote(
    footnote = "Same-sex marriages only identifiable beginning in 2013",
    locations = cells_body(columns = ssmarriage, rows = 8)
  )
```

[^same-sex-cohab]: While it is possible to identify same-sex cohabitations from 1990 onward, comparison across time is complicated by changes to coding practices  and the possibility of response errors overwhelming the small number of same-sex cohabitations [@black_demographics_2000;@black_measurement_2007]. Therefore, I restrict my analysis to same-sex marriages since 2013.

For most of the analysis, I also restrict the sample to couples in which both partners are native born. Foreign-born individuals may have been married prior to arrival in the US. Even in cases where foreign-born individuals arrived in the US unmarried, they will have substantial preferences for ethnonational endogamy that could easily be conflated with racial preferences, although this may be less true of individuals who arrived as children (i.e. the 1.5 generation). Figure S4 in the supplementary appendix shows how the results would change for single-race groups involving the Hispanic or Asian/Pacific Islander groups when including some or all of the foreign-born population. The results show that the overall odds of interracial marriage would decline substantially, but the trends are generally similar. The results are similar regardless of whether the entire foreign-born population or only the 1.5 generation is included.

## Measuring Race

In order to measure the frequency of marriage across groups, I first need to establish what those groups will be. Ideally, group distinctions would capture meaningful and salient ideas within a given society of where group boundaries lie. However, over the substantial period of time analyzed here, the meaning, placement, salience, and measurement of boundaries can change. For these reasons, I confine most of the analysis to the large single race groups of White, Black, American Indian/Alaska Native (AIAN), Asian/Pacific Islander (API), and Hispanic, where conceptual and methodological changes are less likely to cause discrepancy over time.

Nonetheless, several changes in Census coding of race are noteworthy. First, Hispanic identification is not determined by a category on the race question but by a separate question on Hispanicity that precedes the race question. This question was not asked on all Census forms until 1980.[^2] Thus, I cannot measure Hispanic interracial marriage patterns prior to 1980 and Hispanics in the Census in 1960 and 1970 will instead appear in other racial categories. From 1980 onward, individuals who identified as Hispanic on the Hispanicity question are identified as Hispanic regardless of their response to the race question in most of the analysis. I do conduct a supplementary analysis that uses Hispanic subgroups identified by their response to the race question. For this supplementary analysis, I use the three largest categories of White, Black, and Other. While some scholars have treated this approach as synonymous with race or skin tone [@qian_latinos_2004; @choi_racial_2020], I do not assume that here as racial identification choices among Hispanics are complex, multi-faceted, and fluid [@rodriguez_changing_2000]. Nonetheless, these identification choices may reflect some role of both phenotype and cultural assimilation into the American racial schema (or lack thereof) and are therefore worth exploring, particularly in relation to marriage with non-Hispanic White and Black partners.

[^2]: Form 1 of the 1970 Census also included a Hispanicity question, but the results from this Hispanicity question have been found to be not comparable to the Hispanicity question asked in 1980 onward [@uscensusbureau_1990_1992, p. B-13].

Second, combining all Asian and Pacific Islander respondents into a single category is less than ideal because of a great deal of heterogeneity within this panethnic group. Prior work has found that although South Asians are nominally categorized as Asian, their actual racial placement in the US is more ambiguous and they are less attached to Asian panethnic identity [@kibria_not_1996; @morning_racial_2001; @schachter_finding_2014]. Pacific Islander populations were aggregated into the API category based on the now defunct 1977 Office of Management and Budget (OMB) standards, yet share a history of settler colonialism more similar to other indigenous groups [@sasa_just_2022]. Although it is possible to identify more specific ethnonational identities (e.g. Chinese, Indian, Hawaiian) in the data, the categories included in a given year are not consistent across time. Furthermore, the sample size of specific ethnonational groups would preclude analysis for most groups. For most of the analysis, I therefore rely upon the Asian/Pacific Islander (API) aggregate. However, in supplementary analyses, I also explore trends where this group is broken into three subgroups of East/Southeast Asian, South Asian, and Pacific Islander.[^3]

[^3]: The South Asian category is only identifiable from 1980 onward. Native Hawaiians are only identifiable from 1970 onward and other specific Pacific Islander groups (e.g. Tongan, Chamorro) are only identifiable from 1990 onward. The 1980 Census includes a category of "Other Asian and Pacific Islander" that makes it impossible to fully distinguish these groups in that year. I treat this issue as a missing data problem. Individuals in this category in 1980 are assigned to an API sub-group category via predictive mean matching based on 1990 Census data. 

Third, starting in 2000, individuals are allowed to check all categories that apply in response to the race question. This shift will likely affect trends in interracial marriage, but not in a manner that is easy to predict a priori, because we don't know who is mostly likely to change their race response. For the main analysis, I separate multiracial respondents from monoracial respondents and I measure trends in interracial marriage for the largest multiracial groups of White-Black, White-AIAN, and White-API.[^slashuse] In a supplementary analysis, I consider trends in interracial marriage among single race groups under different counterfactual scenarios where multiracial respondents are forced back into a single race option.

[^slashuse]: Throughout the article, I use a slash ("/") to distinguish a union across two groups (e.g. White/Black for marriage between a White and Black spouse) and a dash ("-") to indicate a multiracial individual with ancestry from two groups (e.g. White-Black to indicate an individual that identified as both White and Black). 

Finally, from 2020 onward, the Census and ACS substantially changed the format of the race and Hispanicity question, although the large panethnic groups used here remained the same. However, the Census Bureau changed its method of allocation of race responses in such a way that the number of multiracial respondents increased explosively [@starr_multiracial_2024; @ventura_rise_2025]. This increase was most notable for individuals who identified as Hispanic and were placed in a multiracial category that included the "Other" category. Because I am identifying Hispanics by the Hispanicity question regardless of their answer to the race question, this measurement issue should not lead to large discrepancies in the results presented here, but any substantial changes after 2020 could be driven in part by this change. As a sensitivity test, Table SX from the supplementary materials shows the distribution of race for the two five year periods immediately prior to the 2020 shift in comparison to the 2020-2023 period. For most groups, the shifts are minimal, but there are substantially increases in the multiracial groups of White-Black and White-AIAN. Therefore, shifts in the last period for interracial marriage involving these groups should be viewed with some skepticism.

<!-- TODO: leaving this out for now, as its a bit too in the weeds. Even apart from these specific changes to measurement, it is possible that some patterns of change in interracial marriage may be driven by changes in patterns of racial identification over time. Race is not fixed and longitudinal data has shown that people do change their racial identification over time [@saperstein_racial_2012; @liebler_america_2017]. Some of this change may be a kind of "churning" with no particular direction or trend over time, in which case, we can think of the racial categories here as measuring a probabilistic assessment of one's race. Other changes may reflect long term secular change which could affect patterns of interracial marriage. The clearest case of this kind of change over the time period is the renewal and growth of American Indian identification from the mid-Twentieth century to at least the end of that century [@nagel_american_1995; @eschbach_changing_1993]. Such growth may have contributed to increases in interracial marriage when formerly non-American Indian spouses switched their identification to American Indian. However, it is also possible that non-American Indians with American Indian spouses were more likely to switch which would then reduce the measure of interracial marriage.-->

## Measuring Interracial Marriage

```{r}
#| label: tbl-cross-product
#| tbl-cap: Abstract table of the frequency of interracial marriage across three groups, cross-tabulated by husband's race and wife's race.

tibble(
  "hrace" = c("Group 1", "Group 2", "Group 3"),
  "g1" = c("$F_{11}$", "$F_{21}$", "$F_{31}$"),
  "g2" = c("$F_{12}$", "$F_{22}$", "$F_{32}$"),
  "g3" = c("$F_{11}$", "$F_{21}$", "$F_{31}$"),
)  |> 
  gt() |>
  fmt_markdown() |>
  cols_align("center", columns = starts_with("g")) |>
  cols_label(hrace = "Husband",
             g1 = "Group 1", g2 = "Group 2", g3 = "Group 3") |>
  tab_spanner(label = "Wife", columns = starts_with("g"))
```

To help fix ideas, @tbl-cross-product provides a view of the frequency of interracial marriage for three abstract groups when marriages are cross-tabulated by the race of husband and wife.[^same-sex-cross-tab] From this table, I can calculate either outmarriage ratios or odds ratios. For example, the outmarriage ratio for husbands from Group 1 would be:

[^same-sex-cross-tab]: The same-sex marriage sample cannot be cross-tabulated by gender. Because this sample only includes heads of households and their partners, I instead cross-tabulate same-sex marriages by head of household status instead of gender. Because all the models used in the analysis are symmetric with respect to gender, this approach is applied for convenience, but I would get the same result by randomly assigning each partner to the row or column.

$$
\text{outmarriage ratio} = \frac{F_{21}+F_{31}}{F_{11}+F_{21}+F_{31}}
$$ {#eq-outmarriage-ratio}

This outmarriage ratio measures the proportion of Group 1 husbands who are married to a member of a different group. Parallel outmarriage ratios can be constructed for wives. An outmarriage ratio for all marriages can also be calculated by taking the sum of the off-diagonal cells in @@tbl-cross-product divided by the total number of marriages.

As discussed earlier, the outmarriage ratio provides a useful summary of the experience of interracial marriage but can be heavily influenced by relative group size. For this reason, the odds ratio is often preferred as a measure of social distance because it factors out relative group size. The odds ratio can be calculated as the inverse cross-product between any two groups in @tbl-cross-product. For example, the odds ratio of marriage between Groups 1 and 2 is given by:

$$
\text{odds ratio} = \frac{F_{12}F_{21}}{F_{11}F_{22}}
$$ {#eq-cross-prod}

This cross-product can be interpreted as the odds ratio of an individual from one of the groups having a partner from the other group relative to those same odds for an individual from the other group. If this cross-product is equal to one then group membership is not a factor in partner selection. Values lower than one indicate barriers to interracial marriage.

Although this odds ratio and its standard error can be estimated directly, I instead use a model approach to estimate these values for the analysis. By using a model, I can condition the results on other factors, discussed below. To estimate the odds ratios, I use a log-linear model approach. The log-linear model has long been used by sociologists to estimate association in contingency tables of varying dimension. Log-linear models differ from more familiar models because they predict the frequency counts of a contingency table and association between variables is determined by interaction terms rather than main effects. For @tbl-cross-product above, I could estimate a null model that just predicts cell frequencies by the marginal distribution of husband's and wife's races:

$$
\log(F_{ij}) = \lambda+\lambda_i+\lambda_j
$$

where $F_{ij}$ is the frequency count in row $i$ (husband's race) and column $j$ (wife's race). This model assumes there is no association between the row and column variables. The model is typically estimated by a poisson generalized linear model with dummy coding of the row and column categories. Fitted cell counts from this model predict the number of marriages expected under the assumption that individuals partner without regard for race. These fitted cell counts will be identical to those used for a chi-squared test of independence for contingency tables.

To account for the association between husband's and wife's race, a two-way interaction term can be added:

$$
\log(F_{ij}) = \lambda+\lambda_i+\lambda_j+\lambda_{ij}
$$
The $\lambda_{ij}$ interaction term contains $(I-1)(J-1)$ parameters, where $I$ and $J$ are the number of categories for the row and column variables, respectively. These interaction parameters saturate the model, such that the fitted values will exactly match the observed frequency counts for the contingency table. The interaction parameters themselves identify all of the non-redundant log-odds ratios that can be derived from the contingency table. It is also possible use a reduced parameterization of the interaction that does not saturate the model, yielding a more parsimonious representation, although that is not my goal here.

Default dummy coding of categories will not produce $\lambda_{ij}$ estimates that directly match the desired log odds ratios between each racial pairing. Instead, I use a custom parameterization as follows:

$$
\log(F_{ij})=\lambda+\lambda_i+\lambda_j+\beta_{ij}+\alpha_{ij}
$$ {#eq-model-basic}

The $\beta_{ij}$ and $\alpha_{ij}$ parameters account for the saturated interaction between husband's and wife's race. The $\beta_{ij}$ parameters estimate the desired log odds ratios for each non-endogamous pairing of husband's and wife's race. The $\alpha_{ij}$ parameters are nuisance parameters required to completely saturate the model and constraining the $\beta_{ij}$ parameters to equal the desired log odds ratios. More details on this parameterization are provided in the supplementary materials. 

One advantage of log-linear models is that they naturally extend to higher-dimensional contingency tables. I take advantage of this property to estimate models based on a four-way contingency table defined by husband's race, wife's race, state of residence, and year. The full model is given by:

$$
\log(F_{ijkt}) = \lambda+\lambda_i+\lambda_j+\lambda_k+\lambda_t+\lambda_{ik}+\lambda_{jk}+\lambda_{it}+\lambda_{jt}+\lambda_{ikt}+\lambda_{jkt}+\beta_{ijt}+\alpha_{ijt}
$$ {#eq-model-full}

where $k$ is the state of residence and $t$ is the time period. All the two and three way interactions between each partner's race, time, and state are added to the model and the $\beta$ and $\alpha$ parameters themselves vary by time.[^zero-cells]

This specification allows the $\beta$ parameters to vary over time, allowing for an analysis of trends. Including state of residence adjusts for geographic differences in the racial composition of local populations. This geographic adjustment is important although imperfect as states are usually too large to serve as a marriage market and I only have the state of residence at the time of data collection, not at the time of the marriage. Nonetheless, adjusting for state variation will account for much of the geographic variation in population distribution by race.

[^zero-cells]: Zero cell counts in the cells forming the odds ratio in @eq-cross-prod will prevent estimation of the parameters of the model for that group comparison. Such zero cell counts are more likely when sample size for one or both groups is small and when the underlying likelihood of interracial marriage is small. Determining the degree of subgroup analysis that could be sustained without being affected by such zero cell counts was determined by investigation of cell frequencies. For the main results, the log-odds ratio is not reported for the AIAN/API pairing in 1960 due to zero cell counts.

<!-- For example, in a three-by-three table like @tbl-cross-product, one can identify four non-redundant odds ratios, but only three intermarriage odds ratios ($\beta_{ij}$) can be estimated for the three pairwise group comparisons. The fourth unique odds ratio is estimated with $\alpha_{ij}$.

To clarify this approach, @tbl-model-param, shows the parameter coding that would be used for a model in the three group case shown in @tbl-cross-product. The three $\beta$ parameters are estimated by dummy coding the upper off-diagonal cell for each pairwise comparison. The single $\alpha$ parameter is estimated via ANOVA style coding for one of the pairings. Using this modeling approach, the exponentiated $\beta$ parameters of the log-linear model will exactly correspond to the odds ratios estimated manually by the cross-product.

```{r}
#| label: tbl-model-param
#| tbl-cap: Log-linear model parameters applied to a cross-tabulation of husband's race and wife's race for three groups.

tibble(
  "hrace" = c("Group 1", "Group 2", "Group 3"),
  "g1" = c("$0$", "$0$", "$0$"),
  "g2" = c("$\\beta_{12}$", "$0$", "$-\\alpha_{23}$"),
  "g3" = c("$\\beta_{13}$", "$\\beta_{23}+\\alpha_{23}$", "$0$"),
)  |> 
  gt() |>
  fmt_markdown() |>
  cols_align("center", columns = starts_with("g")) |>
  cols_label(hrace = "Husband",
             g1 = "Group 1", g2 = "Group 2", g3 = "Group 3") |>
  tab_spanner(label = "Wife", columns = starts_with("g"))
``` 

-->


## Measuring Stocks vs. Flows

Ideally, I would use data on the "flow" of recent marriages rather than the "stock" of extant marriages. The stock of extant marriages includes marriages that were formed decades prior to the date of data collection and thus will indicate the conditions of that time period rather than the current time period. Furthermore, extant marriages reflect not only marriage patterns but also, implicitly, patterns of marriage dissolution through divorce or widowhood. Scholars have theorized that crossing racial boundaries may put greater stress on interracial marriages, leading to a higher rate of divorce [@zhang_marital_2009]. However, empirical work has found somewhat mixed and inconclusive evidence of differential dissolution for interracial spouses [@bratter_will_2008;@zhang_marital_2009;@fu_broken_2011].

However, information on marital timing is not available for a substantial subset of the time period. From 1960 to 1980, the Census asked about the age of first marriage for all married respondents. Marital duration questions were removed from Census 1990. A question on the year of the current marriage was added to the ACS in 2007. Thus, between 1990 and 2006, I have no way to identify recent marriage cohorts and in the period before 1990, I can only identify timing of marriage for first marriages.

This issue has often been addressed by restricting the sample by the age of one or both partners [@qian_breaking_1997; @blackwell_mate_2000; @gullickson_education_2006]. By selecting younger partners, we can be assured that the marriage was relatively recently formed. However, this approach can also introduce a host of other biases as younger couples may differ in other ways from the full set of recent marriage cohorts. Furthermore, the age of marriage for both husbands and wives has increased substantially over this time period, meaning that any static restriction by age will lead to different kinds of couples across the time period.

Rather than subset by age, I take all extant marriages but apply weights to each couple to match the joint age distribution of newlywed partners for that time period. This approach can be seen as a generalization of the technique of restricting the analysis to certain age groups. Rather than arbitrarily decide on a restriction, I use the data to represent each (joint) age group proportionally to its share among actual newlyweds. For example, in 2023, the marriages where both husband and wife were between 25-29 years old make up about 2.3% of all extant marriages, but 15.3% of all marriages formed within the last year. Therefore, the age weight for this group is 6.65 (15.3/2.3) to represent its higher frequency among newlyweds than extant marriages. Because these weights are derived from the joint age distribution within each time period, this approach also intrinsically adjusts for temporal changes in the timing of marriage for both spouses.

From 2007, I can estimate joint age distributions of recent marriage cohorts directly from the ACS data using couples married in the last year. Prior to 2007, I use publicly available tables on the joint distribution of the age of spouses from US vital statistics reports for 1960, 1970, 1980, and 1988 [@nationalcenterforhealthstatistics_vital_1960; @nationalcenterforhealthstatistics_vital_1970; @nationalcenterforhealthstatistics_vital_1980; @nationalcenterforhealthstatistics_vital_1988]. After 1988, these reports were no longer produced for marriage. From 1990 to 2006, I use linear interpolation of the joint age distribution. The supplementary appendix provides a full discussion of the weighting procedure and interpolation.

This approach cannot eliminate every bias from the use of the stock of marriages, including bias arising from differential rates of union dissolution. Figure S3 in the supplementary appendix provides a comparison of the age weighted results to results based on newlyweds (married within the last year) in years where such comparison is possible. The results indicate that the age weighted results are quite close to the results for newlyweds both in magnitude and trend, although the odds ratios for newlyweds tend to be a little higher, as expected.

I apply age weights separately to opposite-sex and same-sex marriages. I have no information on the duration of cohabitation. For comparative perspective with married couples, I apply the same age weighting of married couples to cohabiting couples.

In addition to the age weighting, I also apply sample weights to adjust for sample design in the Census and ACS results.[^4] Weights are applied by taking the sum of weights as frequency counts. Because of the use of weights as well as complex survey design, the analytical standard errors of model parameter estimates will be underestimated. No clear analytical solution exists for this issue, so I use a bootstrap technique to generate standard errors by resampling 1000 times from the individual level data within each year to re-generate parameter estimates.

[^4]: Because the unit of observation is a couple, I use the mean of both partner's person weight to derive the couple weight. All sample weights are divided by the mean sample weight for each year.

# Results

All results are presented graphically as the trend in the outmarriage or odds ratio over time.[^rep-data] All figures involving odds ratios use logarithmic scaling of the y-axis to highlight differences over time and across groups, so an observed linear trend in these figures corresponds to exponential growth on the original scale. The colors used for each type of interracial marriage pairing are maintained across separate figures. All figures include 83% confidence bands unless otherwise stated. An 83% confidence band facilitates easier comparison between parameter estimates than a traditional 95% band because an overlap in these bands will be approximately equal to a statistically significant difference at $p<0.05$ [@knol_misuse_2011].

[^rep-data]: The replication package included with this article contains all model estimates in a machine readable format.

## Trends in Outmarriage Ratios

```{r}
#| label: fig-outmarriage-overall
#| fig-cap: The trend in the share of all marriages that are racially exogamous. Values in all years are weighted to reflect the joint age distribution of newlywed partners. 95% confidence band shown.
  
ggplot(exogamy_trend, aes(x = year, y = exogamy, group = 1))+
  geom_ribbon(aes(ymin = exogamy-1.96*exogamy_se,
                  ymax = exogamy+1.96*exogamy_se),
              alpha = 0.3, color = NA)+
  geom_line()+
  geom_point()+
  scale_y_continuous(labels = scales::percent)+
  labs(y = "percent racially exogamous")+
  theme_bw(base_size = 14)
```

@fig-outmarriage-overall shows the overall outmarriage ratio across all married couples over time, weighted to match the joint age distribution of newlywed partners. The figure shows a consistent and relatively linear increase over time in the proportion of married couples who belong to different racial groups. In 1960, racially exogamous couples make up less than 1% of all couples. By the end of the time period, about 1 in 6 couples are in an interracial marriage. A larger than average increase in the final time period may reflect some changes to the measurement of race since 2020.

The simple trend shown in @fig-outmarriage-overall suggests rapidly diminishing racial boundaries. However, when we look at the outmarriage ratio separately by racial group, the picture becomes more complicated. @fig-outmarriage-race shows outmarriage ratios for each single race group over time, and separately by the sex of the spouse. As before, all results are weighted to reflect the joint age distribution of newlyweds.

```{r}
#| label: fig-outmarriage-race
#| fig-cap: The trend in the share of spouses of a given race and gender that are in a racially exogamous marriage. Values in all years are weighted to reflect the joint age distribution of newlywed partners. 83% confidence bands shown.

exogamy_trend_race |>
  filter(!str_detect(race, "-")) |>
  ggplot(aes(x = year, y = exogamy, color = spouse, group = spouse))+
  geom_ribbon(aes(ymin = exogamy-1.83*exogamy_se,
                  ymax = exogamy+1.83*exogamy_se, 
                  fill = spouse),
              alpha = 0.3, color = NA)+
  geom_line(alpha = 0.7)+
  geom_point(size = 1, alpha = 0.7)+
  facet_wrap(~race)+
  labs(y = "percent racially exogamous")+
  scale_y_continuous(labels = scales::percent)+
  scale_color_manual(values = c("red", "navy", "gold"))+
  theme_bw(base_size = 14)+
  theme(panel.spacing.x = unit(0.75, "lines")) 
```


For both the AIAN and API groups, outmarriage ratios largely stopped increasing (after 1990 for AIAN and after 2000 for API) and have remained flat, or have slightly declined since. In contrast, the outmarriage ratios for White, Black, and Hispanic partners have increased over the entire time period.

Consistent with prior work, @fig-outmarriage-race also shows that the likelihood of racial exogamy varies by gender for Black and API spouses [@jacobs_gender_2002; @gullickson_black_2006]. Black men are much more likely to marry a non-Black partner than Black women and that gap has increased over the time period, although the likelihood has been growing exponentially for both Black men and women. In contrast, the likelihood of racial exogamy is higher for API women than for API men. This difference has remained largely consistent across time and both API women and men experienced the same peak in the likelihood of racial exogamy in 2000, followed by a moderate decline.

@fig-outmarriage-race provides important information on the experience of interracial partnering for different groups, but the results are highly reflective of group size differences. For example, as I will show below, once we adjust for group size, it is actually the Hispanic population which tends to have the highest likelihood of racial exogamy. This pattern is not apparent in @fig-outmarriage-race, because no adjustment is made for relative group size. I now turn to an analysis of the odds of racial exogamy, which provide a better measure of the underlying social distance between groups.

## Overall Trends in Odds Ratios

@fig-intermar-single-overall shows the odds of racial exogamy over time across all possible pairings of single race groups. Results are estimated from log-linear models and weighted by the joint age distribution of newlyweds at each time point. Each panel shows the results of each possible interracial pairing for a particular focal racial group. Thus, each interracial pairing is shown twice across all of the panels.

```{r}
#| label: fig-intermar-single-overall
#| fig-cap: Odds of racial exogamy across all single race pairings, separately by the focal race of spouse. Estimates are derived from log-linear models that control for state racial composition and are weighted to reflect the joint age distribution of spouses. 83% confidence bands shown.
#| fig-height: 8
#| fig-pos: "p"

model_single_state_bs <- model_state_bs |>
  filter(!str_detect(term, "-"))

lims <- exp(c(min(model_single_state_bs$conf.low), 
              max(model_single_state_bs$conf.high)))

# we need to facet instead of subcap because word doesn't like it. Also it will
# be a pain for journal editorial teams (sigh)
model_single_state_bs |>
  mutate(focal1 = str_extract(term, "^[^/]+"),
         focal2 = str_extract(term, "(?<=/).*")) |>
  pivot_longer(cols = starts_with("focal"), values_to = "focal") |>
  mutate(focal = factor(
    focal, 
    levels = c("White", "Black", "AIAN", "API", "Hispanic"),
    labels = paste0("Involving ", 
                    c("White", "Black", "AIAN", "API", "Hispanic"),
                    " spouse")
    )) |>
  ggplot(aes(x = year, y = exp(estimate), color = term))+
  #geom_vline(xintercept = 2000, linetype = 2)+
  geom_ribbon(aes(ymin = exp(conf.low), ymax = exp(conf.high), fill = term),
              alpha = 0.3, color = NA)+
  geom_line()+
  geom_point(size = 0.5)+
  facet_wrap(~focal, ncol = 2)+
  labs(y = NULL, color = "pairing", fill = "pairing")+
  scale_y_log10(limits = lims, labels = scales::comma)+
  scale_color_manual(values = palette)+
  scale_fill_manual(values = palette)+
  theme_bw(base_size = 13)
```

The trends in @fig-intermar-single-overall are consistent with the plateau of outmarriage rates observed in @fig-outmarriage-race for AIAN and API partners. Across all possible combinations, racial exogamy involving an API partner had either plateaued by the late 20th century or has shown no evidence of a rise at all over the time period. The results for AIAN partners are somewhat similar, except for some evidence of a moderate rise in the odds of marriage to Hispanic and Black partners in the later time period.

In contrast, in all cases involving combinations of White, Black, and Hispanic partners, the odds of interracial marriage consistently increase throughout the time period. In the early part of the time period, the odds of interracial marriage with a Black spouse are clearly the lowest of all possible exogamous pairings for each non-Black group. However, in many cases, these odds ratios grow the fastest throughout the time period. By the end of the time period, partnership with a Black spouse is only definitively the least likely form of interracial marriage for White partners. Even in this case, the odds of White/Black partnering are increasing much more rapidly than other pairings involving a White partner. If the trends observed here continue, White/Black interracial marriage will soon become more likely than White/API interracial marriage.

Interracial marriages involving an Hispanic partner are generally among the most likely for all groups. White/Hispanic partnering is the most likely form of interracial marriage throughout the time period in which Hispanics can be identified in the data. Interracial marriage with Hispanic partners is also the most likely form of partnering for Black spouses throughout the time period and the second most likely form of partnering for AIAN and API spouses after marriage with White partners for most of the time period.

The substantial changes to the measurement of race at 2000 and 2020 do not generate any substantial discontinuities to the trends observed here, suggesting that the results for single-race groups are robust to such changes. 

## Including Multiracial Groups

To this point, I have only examined the interracial marriage patterns between single race groups. The option to check more than one box on the race question since 2000 allows me to analyze trends in interracial marriage for some large multiracial groups for part of the time period. @fig-intermar-multi-overall shows those trends for the three largest multiracial groups of White-Black, White-AIAN, and White-API. I examine the tendency of each of these groups to marry a member from each of their two constituent groups, in comparison to the odds of a "first generation" marriage between those constituent groups. For example, for the White-Black case, I compare the odds of a White-Black/White or a White-Black/Black marriage to a White/Black marriage.

```{r}
#| label: fig-intermar-multi-overall
#| fig-cap: Odds of racial exogamy for biracial groups with component single race groups and in comparison to the odds of racial exogamy for the component single race groups. Estimates are derived from log-linear models that control for state racial composition and are weighted to reflect the joint age distribution of spouses. 83% confidence bands shown.


model_state_bs |>
  mutate(group = case_when(
    term == "White/Black" | str_detect(term, "White-Black") ~ "White-Black",
    term == "White/AIAN" | str_detect(term, "White-AIAN") ~ "White-AIAN",
    term == "White/API" | str_detect(term, "White-API") ~ "White-API",
    TRUE ~ NA_character_
  )) |>
  filter(!is.na(group)) |>
  mutate(group = factor(group, 
                        levels = c("White-Black", "White-AIAN", "White-API"))) |>
  ggplot(aes(x = year, y = exp(estimate), color = term))+
  #geom_vline(xintercept = 2000, linetype = 2)+
  geom_ribbon(aes(ymin = exp(conf.low), ymax = exp(conf.high), fill = term),
              alpha = 0.3, color = NA)+
  geom_line()+
  geom_point(size = 0.5)+
  facet_wrap(~group, ncol = 2)+
  labs(y = NULL, color = "pairing", fill = "pairing")+
  scale_y_log10(labels = scales::comma)+
  scale_color_manual(values = palette)+
  scale_fill_manual(values = palette)+
  theme_bw(base_size = 14)
```

For both the White-Black and the White-API case, I observe a similar pattern. Multiracial individuals are considerably more likely to marry a partner from either constituent group than members of the constituent groups are to marry one another. This finding suggests that White-Black and White-API individuals face lower boundaries to interracial marriage with these groups because they are in some sense partially endogamous with both groups [@gullickson_blurring_2025].

For the White-Black group, one might expect that, due to the one-drop rule, White-Black individuals would be more likely to marry Black rather than White partners. The odds of partnering with a Black spouse are slightly higher than the odds of partnering with a White spouse in the early period, but this slight difference has gone away by the end of the time period, Although the last period should be viewed with some caution due to measurement issues discussed earlier, the closing of the gap is apparent in earlier periods as well. Thus, the results here suggest a very weak role of the one-drop rule in the marriage selection choices of White-Black individuals. 

For White-API individuals, The odds of marriage with White and API partners are very similar. By the end of the period, the odds of a White-API individual partnering with a White spouse appear to be growing faster than the odds of partnering with an API spouse, but this finding should be considered tentative because of potential issues with measurement for this group in the last period, discussed earlier.

For White-AIAN individuals, we observe a very different pattern. Partnering with White spouses is more common than "first generation" White/AIAN interracial marriage and the odds of this form of intermarriage are growing over time. In contrast, the odds of partnership with AIAN spouses are less likely than those odds for "first generation" White/AIAN interracial marriage and do not appear to be increasing over time. These results are consistent with a norm of hyperdescent in which individuals of White-AIAN ancestry are more likely to be seen and treated as White than AIAN [@gullickson_choosing_2011; @iverson_regimes_2022].

The change in 2000 to multiracial reporting may also affect the trends for single race groups, depending on which individuals were more likely to reclassify themselves as multiracial. While we can never know precisely the affect of this shift in reporting, I develop three counterfactual scenarios to better understand how the trends in interracial marriage might have looked if multiracial respondents had remained in single race categories. For sample size reasons, I focus on the three categories of White/Black, White/API, and White/AIAN interracial marriage. In each case, I consider three scenarios for reclassification of part-White multiracial respondents. In the first scenario, all these respondents are re-classified as White. In the second scenario, all these respondents are classified by their minority race. In the third scenario, I use the "best race" variable created by the Minnesota Population Center. This variable is based on multinomial logit models of data from the National Health Interview Survey, which asks multiracial respondents for their preferred single race.

```{r}
#| label: fig-multi-wb-counterf
#| fig-cap: Comparison of odds of racial exogamy with different counterfactual assumptions about single race reporting among multiracial respondents. Estimates are derived from log-linear models that control for state racial composition and are weighted to reflect the joint age distribution of spouses. 83% confidence bands shown.

model_state_bs |>
  mutate(counterfactual = "Actual") |>
  bind_rows(model_cf_white_state_bs) |>
  bind_rows(model_cf_minority_state_bs) |>
  bind_rows(model_cf_best_state_bs) |>
  filter(counterfactual == "Actual" | year >= 1990) |>
  filter(term == "White/Black" | term == "White/API" | term == "White/AIAN") |>
  mutate(counterfactual = factor(counterfactual, 
                                 levels = c("Actual", "White", "Minority", 
                                            "Best"))) |>
  ggplot(aes(x = year, y = exp(estimate), color = counterfactual))+
  #geom_vline(xintercept = 2000, linetype = 2)+
  geom_ribbon(aes(ymin = exp(conf.low), ymax = exp(conf.high), 
                  fill = counterfactual),
              alpha = 0.3, color = NA)+
  geom_line()+
  geom_point(size = 0.5)+
  facet_wrap(~term, ncol = 2)+
  scale_y_log10(labels = scales::comma)+
  labs(y = NULL)+
  scale_color_manual(values = c("navy", "gold", "red", "#004C2E"))+
  scale_fill_manual(values = c("navy", "gold", "red", "#004C2E"))+
  theme_bw(base_size = 14)
```

@fig-multi-wb-counterf presents the trends resulting from these different scenarios. For White/Black respondents, the overall trend is quite similar across all scenarios, although all three scenarios would have slightly increased the odds of White/Black marriage over that actually observed after 2000.

For both White/API and White/AIAN interracial marriage, the scenario which differs from the other cases is when multiracial part-White respondents are identified by their single-race minority group. In this scenario, the odds of interracial marriage are substantially higher than the other two scenarios and the actual observed trend. The other two scenarios produce results that are statistically indistinguishable from the actual trend. The results here suggest that many part-White multiracial individuals have White spouses and thus reclassifying these individuals into single race minority groups would increase the observed odds of interracial marriage with White spouses. However, this assumption is likely unrealistic for individuals of part API and AIAN ancestry, both of which have been shown not to follow the logic of the one-drop rule [@gullickson_choosing_2011].

## Trends Among Hispanic Subgroups

I now turn to results that divide the Hispanic group into three separate subgroups, identified by racial identification as "White", "Black", and "Other." @fig-sub-hispanic shows trends in the odds ratio of marriage between these three subgroups and non-Hispanic White and Black partners. For reference, the trend in White/Black interracial marriage is also shown.

```{r}
#| label: fig-sub-hispanic
#| fig-cap: Odds of racial exogamy for Hispanic subgroups with non-Hispanic White and Black partners. White/Black exogamy odds are also shown for reference. Estimates are derived from log-linear models that control for state racial composition and are weighted to reflect the joint age distribution of spouses. 83% confidence bands shown.
#| fig-height: 3

lims <- exp(c(min(model_hispanic_sub_state_bs$conf.low), 
              max(model_hispanic_sub_state_bs$conf.high)))

temp <- model_hispanic_sub_state_bs |>
  filter(term == "White/Black") |>
  mutate(group = "Black")

model_hispanic_sub_state_bs |>
  filter(!str_detect(term, "Hispanic/")) |>
  mutate(group = if_else(str_detect(term, "^Black"), "Black", "White"),
         term = str_replace(term, "Hispanic", " Hispanic")) |>
  bind_rows(temp) |>
  mutate(group = factor(group, levels = c("White", "Black")),
         term = factor(term, 
                       levels = c("White/White Hispanic",
                                  "White/Other Hispanic",
                                  "White/Black Hispanic",
                                  "Black/Black Hispanic",
                                  "Black/Other Hispanic",
                                  "Black/White Hispanic",
                                  "White/Black"))) |>
  ggplot(aes(x = year, y = exp(estimate), color = term))+
  #geom_vline(xintercept = 2000, linetype = 2)+
  geom_ribbon(aes(ymin = exp(conf.low), ymax = exp(conf.high), fill = term),
              alpha = 0.3, color = NA)+
  geom_line()+
  geom_point(size = 0.5)+
  facet_wrap(~group)+
  labs(y = NULL, color = "pairing", fill = "pairing")+
  scale_y_log10(limits = lims, labels = scales::comma)+
  scale_color_manual(values = palette)+
  scale_fill_manual(values = palette)+
  theme_bw(base_size = 12)
```

In terms of interracial marriage with White spouses, the ordering of the three subgroups is as expected. Hispanic individuals who identify as White by race are the most likely to marry non-Hispanic White spouses. The odds of marriage to White spouses are somewhat lower for Hispanic individuals who identify as Other by race. For comparison, both of these odds ratios are substantially higher than the odds of White/Black marriage. The odds of marriage to a White spouse are much lower for Hispanic individuals who identify as Black by race and are slightly lower than the odds of White/Black marriage.

Although the odds are lowest for Black Hispanics, the trend suggests a much faster increase in this odds ratio over time, on par, with the increase in the odds of White/Black marriage. Thus, while there is a clear stratification of the odds of marriage with White partners by Hispanic racial self-identification, these differences have been getting smaller over time.

As one might expect, the ordering of Hispanic groups by the odds of partnership with a non-Hispanic Black spouse are inverted relative to those for a non-Hispanic White spouse. Interracial marriage between non-Hispanic Black spouses and Hispanics who identify racially as either Other or White are both similar to the odds of White/Black interracial marriage over time. The odds of interracial marriage between non-Hispanic Black spouses and Hispanics who identify as Black, on the other hand, are much higher and are comparable to the odds of White/White Hispanic interracial marriage. All three odds ratios are growing at a similar exponential rate over time.

## Trends Among API Subgroups

@fig-sub-asian-simple shows the odds of partnership with White spouses across three subgroups of the API category. I divide the API group into Pacific Islanders, South Asians, and East/Southeast Asians. The results show that the plateau trend seen for the API group overall is only the trend for the East/Southeast Asian group. This group experienced a rapid increase in the odds of partnership with White spouses until 1990, when those odds leveled off. Since 1990, the odds have declined slightly, but have largely remained stationary.

```{r}
#| label: fig-sub-asian-simple
#| fig-cap: Odds of racial exogamy for Asian subgroups with White partners. Estimates are derived from log-linear models that control for state racial composition and are weighted to reflect the joint age distribution of spouses. 83% confidence bands shown.

model_asian_sub_simple_state_bs |>
  filter(str_detect(term, "^White/")) |>
  mutate(term = str_replace(term, "/EastAsian", "/E&SE Asian"),
         term = str_replace(term, "/PI", "/Pac. Islander"),
         term = str_replace(term, "/SouthAsian", "/South Asian")) |>
  ggplot(aes(x = year, y = exp(estimate), color = term))+
  #geom_vline(xintercept = 2000, linetype = 2)+
  geom_ribbon(aes(ymin = exp(conf.low), ymax = exp(conf.high), fill = term),
              alpha = 0.3, color = NA)+
  geom_line()+
  geom_point(size = 0.5)+
  scale_y_log10(labels = scales::comma)+
  labs(y = NULL, color = "pairing", fill = "pairing")+
  scale_color_manual(values = palette)+
  scale_fill_manual(values = palette)+
  theme_bw(base_size = 14)
```

In contrast, the results for South Asians suggest increasing odds of marriage with White spouses throughout the time period, although sparsity of data from before 1980 limits the full range of the time period. Point estimates suggest a decline from 1980 to 1990, but the wide error margins on these estimates make such a decline questionable at best. Since 1990, there has been steady growth in the odds of South Asian/White partnering. However, those odds begin and remain substantially lower than those same odds for South/Southeast Asian partners, even as they are converging across time.

The results for Pacific Islanders show a decline in the odds of marriage with White partners throughout the time period. The decline is particularly concentrated in the period before 2000. Although this group is small, this trend before 2000 is not simply a result of wide error margins. After 2000, the decline is slight and may simply be a result of measurement error.

The decline for Pacific Islanders is notable as the only trend examined here to show a substantial decline across the time period, which is otherwise characterized by substantial increases in the odds of interracial marriage for at least the early part of the time period. Notably, the starting odds of White/Pacific Islander marriage are relatively very high. Given that Pacific Islander identification has been unstable over the time period [@liebler_americas_2017b], it is possible that this trend might reflect changing self-identification practices.

## Trends for Alternate Union Types

To this point, I have presented results only for heterosexual marriages because these marriages are identifiable throughout the time period. I can also identify opposite-sex cohabitations since 1990 and same-sex marriages since 2013. @fig-intermar-alt-unions shows trends in the likelihood of interracial marriage across the three different union types since 1990. Because the sample size is considerably smaller for opposite-sex cohabitations and same-sex marriages, I focus only on unions involving a White partner. Full results for all partnering combinations are available in the supplementary appendix.

```{r}
#| label: fig-intermar-alt-unions
#| fig-cap: Odds of racial exogamy across single race pairings involving a White partner, separately by opposite-sex marriage, opposite-sex cohabitation, and same-sex marriage. Estimates are derived from log-linear models that control for state racial composition and are weighted to reflect the joint age distribution of spouses. 83% confidence bands shown.

model_state_bs <- model_state_bs |>
  mutate(union_type = "opposite-sex marriage")

model_state_cohab_bs <- model_state_cohab_bs |>
  mutate(union_type = "opposite-sex cohabitation")

model_state_same_sex_bs |>
  mutate(union_type = "same-sex marriage") |>
  bind_rows(model_state_bs) |>
  bind_rows(model_state_cohab_bs) |>
  filter(!str_detect(term, "-") & str_detect(term, "White") &
           year >= 1990) |>
  mutate(union_type = factor(union_type,
                             levels = c("opposite-sex marriage",
                                        "opposite-sex cohabitation",
                                        "same-sex marriage"))) |>
  ggplot(aes(x = year, y = exp(estimate), color = union_type))+
  geom_ribbon(aes(ymin = exp(conf.low), ymax = exp(conf.high), fill = union_type),
              alpha = 0.3, color = NA)+
  geom_line()+
  geom_point(size = 0.5)+
  facet_wrap(~term)+
  scale_y_log10(labels = scales::comma)+
  scale_color_manual(values = c("navy", "red", "darkgreen"))+
  scale_fill_manual(values = c("navy", "red", "darkgreen"))+
  labs(y = NULL, color = "union type", fill = "union type")+
  theme_bw(base_size = 14)+
  theme(legend.position = "bottom")
```

As expected from prior work, @fig-intermar-alt-unions shows that the odds of interracial marriage are generally higher in opposite-sex cohabitations and same-sex marriages than in opposite-sex marriages. The one exception to this pattern is for the White/AIAN case, where all three union types have similar odds of interracial partnering. 

While the odds of interracial partnering are higher in alternate union types, the rank-ordering of pairings is similar across all union types. The trends over time are also broadly similar across union types, although the trend line for same-sex marriages is relatively short. Overall, the results here are consistent with the expectation that less "permanent" unions (cohabitation) and unions that already transgress norms in other ways (same-sex marriages) will also be more racially heterogamous than opposite-sex marriages. However, while racial boundaries may be somewhat weaker for these alternate union types, they still play a strong role in predicting who partners with whom. I also observe no narrowing of the gap between union types over time, suggesting that the increasing prevalence of alternate union types is not leading to a convergence in the likelihood of exogamy.

# Conclusions

In general, the likelihood that a marriage will cross racial lines has gone up substantially between 1960 and 2023, regardless of whether one measures that likelihood by outmarriage ratios or odds ratios. However, that general trend masks a great deal of heterogeneity that emerges when examining this trend separately by groups and pairings of groups. This heterogeneity reveals information about the changes in group boundaries over time. 

The most notable deviation from the general trend is for marriages involving an AIAN or API partner. For these cases, the increase in the likelihood of interracial marriage reached a peak by the end of the 20th century for some or all possible pairings. This finding is most consistent for API partners, where I observe either a moderately declining or stationary likelihood of partnership with all other single-race groups since 1990. Subgroup analysis indicates that the White/API trend is largely driven by the experiences of East and Southeast Asian partners. South Asian partners experienced an increase in the likelihood of interracial marriage with White spouses throughout the period, while Pacific Islander partners experienced a decline.

The plateau pattern is only visible for AIAN partners with regard to their marriage with White and API partners. The likelihood of interracial marriage with Hispanic and Black partners has increased consistently throughout the time period.

On the other hand, I  observe a consistent exponential increase in the likelihood of interracial marriage for pairings that involve some combination of White, Black, and Latino partners. Notably, the greatest increase in the likelihood of interracial marriage is for marriages involving a Black partner. In 1960, the likelihood of marrying a Black partner was by far the lowest among all possible pairings for every other racial group. While this hierarchy still persists in many cases, there has been substantial convergence in these likelihoods over time, suggesting that the Black/non-Black boundary has diminished over time. 

At the other end of the spectrum, interracial marriage with an Hispanic partner often ranks as one of the most likely forms of interracial marriage. White/Hispanic interracial marriage has the highest likelihood among all possible single-race pairings throughout the time period. Yet, that pattern is also structured by the self-identified race of the Hispanic partner. Hispanics who identify as White are most likely to marry White partners and least likely to marry Black partners. Hispanics who identify as Black are most likely to marry a Black partner, and least likely to marry a White partner. Hispanics who identify as Other are somewhere in-between in their likelihood to marry a White partner, and similar to White Hispanics in terms of their likelihood of marrying a Black partner. In this way, the interracial marriage patterns of Hispanics tend to reinforce an existing White/Black divide. 

Multiracial groups also tend to have a higher likelihood of interracial marriage with single-race constituent groups than single-race groups have with one another, although I only analyze patterns for the three largest multiracial groups. The White/AIAN group is the exception to this rule. White/AIAN partners are much more likely to marry a White partner than an AIAN partner, consistent with a pattern of hyperdescent. In contrast, I observe only a small weak effect of hypodescent for White/Black partners.

My goals in this article are largely descriptive and thus I can only speculate as to the causes of changing patterns for some groups observed over time. However, these results present some challenges to the argument that increasing immigration accounts for the plateauing of interracial marriage for Asian partners [@qian_social_2007;@qian_generation_2020]. That pattern holds even when analysis is restricted to native-born spouses, so it is not simply a result of greater opportunity for native-immigrant intermarriage among Asians. @qian_generation_2020[pg. 2892] argue that growing immigration may also increase endogamy by reinforcing ethnoracial identity and boundaries more broadly (although see @jimenez_mexican_2008 regarding the complicated relationship between immigration and ethnoracial identity). If this were true, we would expect similar results for Hispanic partners, yet the growth of interracial marriage among Hispanic partners increased at a consistent and rapid pace throughout the time period. Furthermore, when disaggregating by subgroups, I find that the plateau trend only holds for East and Southeast Asians and not for South Asians. Instead, the group with a pattern most similar to the East and Southeast Asian population is the American Indian/Alaska Native population for which immigration plays no role. While these two similar trends may arise from different underlying causes, the overall patterns suggest that factors other than growing immigration are responsible for the plateau and even slight decline in the likelihood of interracial marriage for some groups.

These results are robust to a variety of specification tests including how multiracial responses are recorded, the exclusion of the foreign-born population, or direct use of newlyweds rather than age-weighted data. Although these factors can change the levels observed somewhat, they do not substantially alter the findings with regard to trends. Furthermore, an analysis of opposite-sex cohabitation and same-sex marriage shows that while interracial partnership is more likely in these alternate union types, the overall trends are similar.

Ironically, historically high rates of interracial marriage for API and AIAN individuals combined with more freedom of identification for part-API and part-AIAN individuals may have contributed to the plateau in rates of interracial marriage for these groups. If the biracial children of such unions were identified solely as members of these groups, the observed plateau in interracial marriage rates would be less severe.

Importantly, the odds ratios used for much of the analysis help to factor out group size, which is critical when using interracial marriage as a benchmark to establish the  strength of group boundaries. However, the experiences of groups and the actual frequency of interracial marriages and families are better reflected by the outmarriage ratios. For example, while the odds of interracial marriage involving API partners remain low, roughly 40-50% of API partners are married to someone of a different race. Group size does matter for the overall distribution of multiracial children produced by such interracial unions. Nonetheless, the outmarriage ratios show the same general plateau effect for AIAN and API partners. 

Overall, these results do not consistently point to a pattern of dissolving barriers even though many groups do continue to experience rapid exponential increase in the likelihood of interracial marriage. Rather they point to the ways that pre-existing norms around group boundaries are being challenged, transformed, and re-interpreted. In all cases, the likelihood of interracial marriage remains well below what one would expect if race was not a factor in partner selection at all.

# References

::: {#refs}
:::
